---
title: "Lasso Regression on Cellphone Use Addiction "
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: visual
---

Slides: [slides.html](slides.html){target="_blank"}

## Introduction

In today's ever-growing digital era, what began as a tool of
convenience, smartphones and devices have become a vital piece of our
daily life. While these items enable instant communication across the
globe, growing usage has evolved into an increasing psychological and
social crisis: smartphone addiction. Characterized by the compulsive
need to check notifications and apps, withdrawal symptoms, and severe
interference with daily life, smartphone dependency shows an increasing
link to rising rates of anxiety, sleep deprivation, and diminished
academic performance among young adults [@lin2014]. Identifying the
precise behavioral, social, and psychological drivers of this addiction
is critical for developing effective interventions. Available data
tracking screen time, app usage, and psychological metrics frequently
contain dozens of interconnected variables. Traditional linear
regression models struggle under this weight, frequently from
overfitting and multicollinearity, which obscures the true narrative
hidden within the data [@james2013].

In this project, we will explore the Least Absolute Shrinkage and
Selection Operator (LASSO) method, commonly known as LASSO regression,
to isolate the most critical predictors of smartphone addiction.
Introduced by Robert Tibshirani [-@tibshirani1996], LASSO is a powerful
statistical method designed for both regularization and variable
selection. Traditional ordinary least squares (OLS) regression treats
every available variable as important, which often introduces noise and
reduces predictive accuracy when dealing with large datasets. LASSO
solves this problem by adding a mathematical constraint—known as an L1
penalty—to the sum of the absolute values of the regression coefficients
[@tibshirani1996].

This L1 penalty performs a unique function: it actively shrinks the
coefficients of less important variables toward zero [@james2013].
Notably, unlike other methods such as Ridge regression, LASSO can force
coefficients to become exactly zero [@tibshirani1996]. LASSO
automatically performs variable selection, completely removing
irrelevant or redundant features from the final model, resulting in a
highly interpretable, sparse model consisting only of the most
statistically significant predictors.

Our dataset explores problematic smartphone usage from survey data
collected in China using a Computer Adaptive Testing (CAT) item bank
[@gao2024]. It is comprised of 1,619 respondents with complete data on
six instrument families plus the demographics.

Here we will apply LASSO regression to a comprehensive smartphone
addiction dataset to break down which specific behaviors and
demographics truly drive the dependencies. Smartphone datasets are
particularly noisy, with variables like social media screen time, daily
notification counts, and self-reported anxiety scores often overlapping
and distorting traditional models. By implementing LASSO, we can
mathematically weed out any statistical noise, isolating the leading
factors [@james2013]. Ideally, this approach will aid us in answering a
crucial question: with the countless digital metrics tracked today,
which factors truly lead to smartphone addictions? Our findings aim to
provide a streamlined, data-driven framework that peers and clinicians
alike can use to understand and combat digital dependency.

| Abbreviation | Scale                                             |
|--------------|---------------------------------------------------|
| NMP          | Nomophobia                                        |
| SAPS         | Smartphone Addiction Proneness                    |
| SPAI         | Smartphone Addiction Inventory                    |
| MPAS         | Mobile Phone Addiction Scale                      |
| MPATS        | Mobile Phone Addiction Tendency Scale             |
| SAS          | Smartphone Addiction Scale (SAS_C / SAS_CA items) |

## Methods

We model smartphone addiction severity with penalized linear regression.
Each respondent contributes item-level responses from six validated
instrument families (NMP, SAPS, SPAI, MPAS, MPATS, SAS) plus
demographics. After listwise deletion of incomplete records,
$n = 1{,}619$ complete cases remain. The primary continuous outcome is
**SPAI total**, the sum of 20 SPAI items (range 20–80) [@lin2014]. SPAI
items are excluded from the predictor set when predicting SPAI total to
avoid circularity; we instead ask whether nomophobia, proneness, and
related scales track overall SPAI severity. A secondary categorical
outcome, **SPAI item 12**, captures perceived negative impact on study
or work on a four-point Likert scale.

### Ordinary least squares and why LASSO

A standard linear model for continuous outcome $y_i$ and predictors
$\mathbf{x}_i \in \mathbb{R}^p$ is

$$y_i = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} + \varepsilon_i,
\qquad i = 1,\ldots,n.
\tag{1}$$

Ordinary least squares (OLS) minimizes the residual sum of squares
$\sum_{i=1}^{n}(y_i - \hat{y}_i)^2$. With dozens of correlated survey
items, OLS tends to overfit, inflate coefficient variance, and produce
unstable signs under multicollinearity [@james2013; @mcneish2015]. Ridge
regression adds an $L_2$ penalty that shrinks coefficients but rarely
sets them exactly to zero [@hoerl1970]. LASSO replaces that penalty with
an $L_1$ term, enabling simultaneous shrinkage and variable selection
[@tibshirani1996; @hesterberg2008].

### LASSO objective and tuning

LASSO estimates coefficients by minimizing the penalized least-squares
objective [@tibshirani1996; @santosa1986]:

$$\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}.
\tag{2}$$

The tuning parameter $\lambda \geq 0$ controls the bias–variance
trade-off. When $\lambda = 0$, Equation (2) reduces to OLS; as $\lambda$
increases, more coefficients are driven to exactly zero, yielding a
sparse, interpretable model [@efron2004; @dalalyan2017]. We select
$\lambda$ with 10-fold cross-validation on an 80/20 train–test split,
reporting both **`lambda.min`** (lowest CV error) and **`lambda.1se`**
(simplest model within one standard error of the minimum) [@james2013].
For the main paper result we emphasize plain LASSO at `lambda.1se` as a
deliberate parsimony choice: it retains nearly the same test $R^2$ while
selecting far fewer predictors.

Test-set performance for continuous models is summarized by

$$R^2 = 1 - \frac{\sum_{i}(y_i - \hat{y}_i)^2}{\sum_{i}(y_i - \bar{y})^2},
\qquad
\mathrm{RMSE} = \sqrt{\frac{1}{n}\sum_{i}(y_i - \hat{y}_i)^2}.
\tag{3}$$

### Plain LASSO versus group LASSO

We fit two complementary LASSO variants implemented in **`glmnet`**
(plain) and **`grpreg`** (group) [@james2013; @huang2024].

**Plain LASSO** (`alpha = 1`, `family = "gaussian"`) performs item-level
selection: any individual survey item may enter or leave the model.
**Group LASSO** assigns all items within a scale (e.g., SAPS) to a
single group and selects entire scales at once [@yuan2006; @huang2024].
Plain LASSO answers which specific items matter; group LASSO answers
which whole instruments matter when scales are treated as coherent
units.

For classification of SPAI item 12, we additionally fit multinomial
LASSO (`family = "multinomial"`) and refit a multinomial log-linear
model on the selected predictors (post-LASSO) to improve class
probabilities [@james2013]. Class imbalance in SPAI item 12 is evaluated
with balanced accuracy and Cohen's $\kappa$ on the held-out test set.

### Justification for this problem and data

Our CAT item bank contains on the order of 100 item-level predictors
that measure overlapping facets of problematic phone use [@gao2024].
Pearson correlations among scale totals exceed $r = 0.64$ for several
pairs (see Data Exploration), so multicollinearity is expected rather
than exceptional. LASSO is well suited to high-dimensional, collinear
psychological survey data because it stabilizes estimation and returns a
sparse subset of items or scales that best predict addiction severity
[@mcneish2015; @dalalyan2017]. Comparing plain and group LASSO,
alongside `lambda.min` and `lambda.1se`, lets us balance predictive
accuracy against interpretability—an important consideration when
communicating results to clinicians and educators [@james2013;
@huang2024].

## Analysis and Results

### Data Exploration and Visualization

Data come from a Computer Adaptive Testing (CAT) item bank on
problematic mobile phone use [@gao2024]. Respondents completed items
from six instrument families—nomophobia (NMP), smartphone addiction
proneness (SAPS), the Smartphone Addiction Inventory (SPAI) [@lin2014],
the Mobile Phone Addiction Scale (MPAS), the Mobile Phone Addiction
Tendency Scale (MPATS), and the Smartphone Addiction Scale (SAS-C and
SAS-CA)—along with demographics (age, gender, grade). The raw file
contains $n = 1{,}619$ respondents with complete records on all
variables; no listwise missingness remains after filtering. Scale totals
are computed as the sum of item scores within each instrument (higher
totals indicate more problematic use, per the item bank codebook).

The sample is predominantly young adults (mean age $= 18.8$ years,
$SD = 2.9$), with more female ($n = 991$) than male ($n = 628$)
respondents. SPAI total—the primary outcome for regression—has mean
$= 41.8$ ($SD = 10.2$) on the 20–80 scale, with a roughly symmetric
distribution centered below the midpoint.

```{r viz-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(patchwork)

dat <- read.csv("data/DATA_Problematic Mobile Phone Use CAT.csv")
dat <- dat[complete.cases(dat), ]

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$SPAI_total  <- rowSums(dat[, paste0("SPAI_",  1:20)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)
```

```{r fig-spai-dist, fig.cap="Distribution of SPAI total score (primary outcome; n = 1,619 complete cases)."}
ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  annotate("text",
           x = mean(dat$SPAI_total) + 1.5, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Mean = ", round(mean(dat$SPAI_total), 1)),
           color = "#C44E52", size = 3.8) +
  labs(
    title = "Distribution of SPAI Total Score",
    x = "SPAI Total (20–80)", y = "Count"
  )
```

Pearson correlations with SPAI total rank the predictor scales as
follows: SAS ($r = 0.81$), SAPS ($r = 0.71$), MPATS ($r = 0.71$), MPAS
($r = 0.65$), and NMP ($r = 0.55$). The scatter plot below shows the
strongest of these relationships—SAS against SPAI total—with a strong
linear trend and modest residual spread.

```{r fig-scatter-sas, fig.cap="SAS total versus SPAI total with OLS trend line and 95% CI band."}
ggplot(dat, aes(x = SAS_total, y = SPAI_total)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "#C44E52",
              fill = "#f4a7a9", linewidth = 1.1) +
  annotate("text",
           x = max(dat$SAS_total) * 0.05,
           y = max(dat$SPAI_total) * 0.96, hjust = 0,
           label = paste0("r = ", round(cor(dat$SAS_total, dat$SPAI_total), 2)),
           size = 4.5, fontface = "bold", color = "#C44E52") +
  labs(
    title = "SAS Total vs SPAI Total",
    x = "SAS Total (Smartphone Addiction Scale)", y = "SPAI Total (Addiction Severity)"
  )
```

All six scale totals move together (inter-scale $r$ often exceeds 0.60),
confirming substantial **convergent validity** among instruments but
also highlighting the multicollinearity that motivates LASSO
regularization.

```{r fig-corr-heatmap, fig.cap="Pearson correlation heatmap among scale totals (complete cases)."}
corr_cols <- c("NMP_total", "SAPS_total", "SPAI_total", "MPAS_total",
               "MPATS_total", "SAS_total")
corr_mat <- cor(dat[, corr_cols], use = "complete.obs")

corr_long <- as.data.frame(corr_mat) %>%
  mutate(Scale1 = rownames(.)) %>%
  pivot_longer(-Scale1, names_to = "Scale2", values_to = "r") %>%
  mutate(
    Scale1 = gsub("_total", "", Scale1),
    Scale2 = gsub("_total", "", Scale2)
  )

ggplot(corr_long, aes(x = Scale2, y = Scale1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3, color = "black") +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r",
       title = "Scale Total Correlations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Two patterns warrant attention before modeling. First, **SPAI item 12**
(perceived impact on study/work) is highly imbalanced: levels 1–4
contain 262, 577, 688, and 92 responses respectively, so the rarest
category is only $5.7\%$ of the sample—classification models must
account for this skew. Second, although NMP shows the highest average
endorsement when rescaled to a percentage of maximum ($= 68.8\%$), it
correlates **weakest** with SPAI total among the five non-SPAI scales,
hinting that nomophobia may be partially redundant once proneness (SAPS)
and SAS symptoms are considered—a pattern consistent with preliminary
LASSO results where NMP items are often zeroed out. The figure below
summarizes the age and gender composition of the sample.

```{r fig-demographics, fig.cap="Age distribution and demographic breakdown (complete cases).", fig.width=10, fig.height=4}
gender_labels <- c("1" = "Male", "2" = "Female")
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)],
                         levels = unname(gender_labels))

p_age <- ggplot(dat, aes(x = age)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(title = "Age", x = "Years", y = "Count")

p_gender <- dat %>% count(gender_lab) %>%
  ggplot(aes(x = gender_lab, y = n, fill = gender_lab)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Gender", x = NULL, y = "Count")

p_age + p_gender + plot_layout(widths = c(2, 1))
```

Together, these exploratory plots motivate our modeling strategy: SPAI
total is suitable as a continuous Gaussian outcome; the strong SAPS and
SAS relationships suggest they will dominate LASSO selection; and the
shared variance across scales justifies penalized regression over
unregularized OLS [@james2013; @gao2024].

### Modeling and Results

#### Data preprocessing

We began with the raw CAT item bank file ($n = 1{,}619$ rows, 101
variables) and retained only complete cases; no listwise missingness
required exclusion. The primary outcome **SPAI total** was computed as
the sum of 20 SPAI items. All other SPAI item columns were removed from
the predictor matrix when modeling SPAI total to avoid circularity.
Demographics (age, gender, grade) and item-level responses from NMP,
SAPS, MPAS, MPATS, and SAS (SAS-C and SAS-CA) entered as predictors.
Gender and grade were coded as factors; all item responses were treated
as numeric. Predictors were assembled with a model matrix (no intercept
column in the penalized fit). Respondents were split 80/20 into training
and test sets using stratification on SPAI item 12
(`caret::createDataPartition`) so that the imbalanced categorical
outcome remained represented in both partitions; all continuous models
below use the same split.

#### LASSO regression for SPAI total

We fit plain LASSO (`glmnet`, `alpha = 1`, Gaussian family) and group
LASSO (`grpreg`, scales as groups) on the training set with 10-fold
cross-validation to select $\lambda$. Test-set performance is reported
with $R^2$ and RMSE (Equation 3). Following our Methods section, we
emphasize **plain LASSO at `lambda.1se`** as the recommended model: it
achieves strong predictive accuracy while retaining far fewer predictors
than group LASSO at comparable $R^2$ [@james2013; @huang2024].

```{r modeling-setup, warning=FALSE, message=FALSE}
library(glmnet)
library(grpreg)
library(caret)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  mat <- as.matrix(raw_coef)
  vars <- rownames(mat)[mat[, 1] != 0]
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(
    R2 = 1 - ss_res / ss_tot,
    RMSE = sqrt(mean((y_true - y_pred)^2))
  )
}

mod_dat <- dat
mod_dat$gender <- as.factor(mod_dat$gender)
mod_dat$grade <- as.factor(mod_dat$grade)
mod_dat$SPAI_12 <- as.factor(mod_dat$SPAI_12)

train_index <- createDataPartition(mod_dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(mod_dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
```

On the held-out test set ($n = `r nrow(mm_reg$x) -
length(train_index)`$), plain LASSO at`lambda.1se`achieved $R^2 =`r
gauss_results$R2[1]`$ and RMSE
$= `r gauss_results$RMSE\[1\]`$, selecting`r
gauss_results\$N_selected\[1\]`predictors. Group LASSO at`lambda.min`reached a similar $R^2 =`r
gauss_results$R2[2]`$ but retained `r gauss_results$N_selected[2]`
predictors—substantially more complexity for little additional
explanatory gain. In practical terms, the non-SPAI scales explain
roughly three-quarters of the variation in SPAI total severity on
respondents the model had not seen during training.

```{r tbl-gaussian, tbl.cap="Gaussian LASSO models predicting SPAI total from non-SPAI items (test set)."}
kable(gauss_results, align = c("l", "r", "r", "r"),
      col.names = c("Model", "R²", "RMSE", "Predictors selected"))
```

Item-level plain LASSO selection was concentrated in **SAPS** (addiction
proneness) and **SAS** (cognitive and behavioral addiction symptoms).
Nomophobia (NMP) contributed few or no items once these stronger scales
were penalized into the model, consistent with the weaker zero-order
correlation observed in exploration. Demographics played a minor role
relative to the addiction instrument items.

```{r fig-selected-scales, fig.cap="Number of item-level predictors selected by plain LASSO (lambda.1se), grouped by instrument."}
ggplot(scale_counts, aes(x = reorder(Scale, N_items), y = N_items, fill = Scale)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Selected items",
       title = "Plain LASSO Selected Predictors by Scale") +
  theme_minimal(base_size = 12)
```

These results align with the exploratory correlation structure: SAS and
SAPS track SPAI total most closely, and LASSO uses that shared variance
efficiently while zeroing out redundant nomophobia items. The high test
$R^2$ should be interpreted as **convergent validity** among overlapping
addiction measures rather than evidence of causal effects [@gao2024;
@lin2014].

### Conclusion

Using LASSO regression on a Chinese CAT item bank of problematic mobile
phone use [@gao2024], we showed that a sparse plain-LASSO model can
predict SPAI total addiction severity with strong held-out accuracy
($R^2 = 0.75$) from other validated scales and demographics. **SAPS**
and **SAS** emerged as the dominant predictors; **NMP** (nomophobia)
added little once those instruments were in the model. Plain LASSO at
`lambda.1se` offered the best balance of fit and parsimony relative to
group LASSO, which selected many more items for comparable performance.

Several implications follow from these findings. First, the instruments
in this item bank measure closely related facets of problematic
smartphone use; high predictive overlap supports their use as a coherent
battery but also underscores the multicollinearity that motivates
penalized regression over ordinary least squares [@james2013;
@mcneish2015]. Second, because the sample was collected in **China**
from predominantly young adults (mean age $= 18.8$ years), results
should not be generalized uncritically to other countries or age groups
without external validation [@gao2024]. Cultural norms around mobile
use, educational pressure, and access to digital services differ across
settings; replication in U.S. or other populations would clarify whether
SAPS and SAS remain the leading predictors of SPAI severity. Third, for
clinicians and educators working with similar screening batteries, a
short subset of SAPS and SAS items identified by LASSO may suffice to
estimate overall severity, though prospective validation is still
needed.

Future work could examine post-LASSO refitting, robustness across
train–test splits and random seeds, and ordered-outcome models for SPAI
item 12. Extending the analysis to new geographic samples would
strengthen claims about which behavioral and psychological drivers of
smartphone addiction are universal versus context-specific.

## References