---
title: "Lasso Regression on Cellphone Use Addiction "
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}

## Introduction

In today's ever-growing digital era, what began as a tool of convenience,
smartphones and devices have become a vital piece of our daily life. While
these items enable instant communication across the globe, growing usage has
evolved into an increasing psychological and social crisis: smartphone
addiction. Characterized by the compulsive need to check notifications and
apps, withdrawal symptoms, and severe interference with daily life,
smartphone dependency shows an increasing link to rising rates of anxiety,
sleep deprivation, and diminished academic performance among young adults
[@lin2014]. Identifying the precise behavioral, social, and psychological
drivers of this addiction is critical for developing effective interventions.
Available data tracking screen time, app usage, and psychological metrics
frequently contain dozens of interconnected variables. Traditional linear
regression models struggle under this weight, frequently from overfitting and
multicollinearity, which obscures the true narrative hidden within the data
[@james2013].

In this project, we will explore the Least Absolute Shrinkage and Selection
Operator (LASSO) method, commonly known as LASSO regression, to isolate the
most critical predictors of smartphone addiction. Introduced by Robert
Tibshirani [-@tibshirani1996], LASSO is a powerful statistical method designed
for both regularization and variable selection. Traditional ordinary least
squares (OLS) regression treats every available variable as important, which
often introduces noise and reduces predictive accuracy when dealing with large
datasets. LASSO solves this problem by adding a mathematical constraint—known
as an L1 penalty—to the sum of the absolute values of the regression
coefficients [@tibshirani1996].

This L1 penalty performs a unique function: it actively shrinks the
coefficients of less important variables toward zero [@james2013]. Notably,
unlike other methods such as Ridge regression, LASSO can force coefficients to
become exactly zero [@tibshirani1996]. LASSO automatically performs variable
selection, completely removing irrelevant or redundant features from the final
model, resulting in a highly interpretable, sparse model consisting only of the
most statistically significant predictors.

Our dataset explores problematic smartphone usage from survey data from a
Computer Adaptive Testing (CAT) item bank. It is comprised of 1,619
respondents with complete data on six instrument families plus the demographics.

Here we will apply LASSO regression to a comprehensive smartphone addiction
dataset to break down which specific behaviors and demographics truly drive the
dependencies. Smartphone datasets are particularly noisy, with variables like
social media screen time, daily notification counts, and self-reported anxiety
scores often overlapping and distorting traditional models. By implementing
LASSO, we can mathematically weed out any statistical noise, isolating the
leading factors [@james2013]. Ideally, this approach will aid us in answering a
crucial question: with the countless digital metrics tracked today, which
factors truly lead to smartphone addictions? Our findings aim to provide a
streamlined, data-driven framework that peers and clinicians alike can use to
understand and combat digital dependency.

| Abbreviation | Scale |
|--------------|-------|
| NMP | Nomophobia |
| SAPS | Smartphone Addiction Proneness |
| SPAI | Smartphone Addiction Inventory |
| MPAS | Mobile Phone Addiction Scale |
| MPATS | Mobile Phone Addiction Tendency Scale |
| SAS | Smartphone Addiction Scale (SAS_C / SAS_CA items) |

## Methods

We model smartphone addiction severity with penalized linear regression.
Each respondent contributes item-level responses from six validated
instrument families (NMP, SAPS, SPAI, MPAS, MPATS, SAS) plus
demographics. After listwise deletion of incomplete records, $n = 1{,}619$
complete cases remain. The primary continuous outcome is **SPAI total**,
the sum of 20 SPAI items (range 20–80) [@lin2014]. SPAI items are
excluded from the predictor set when predicting SPAI total to avoid
circularity; we instead ask whether nomophobia, proneness, and related
scales track overall SPAI severity. A secondary categorical outcome,
**SPAI item 12**, captures perceived negative impact on study or work on
a four-point Likert scale.

### Ordinary least squares and why LASSO

A standard linear model for continuous outcome $y_i$ and predictors
$\mathbf{x}_i \in \mathbb{R}^p$ is

$$
y_i = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} + \varepsilon_i,
\qquad i = 1,\ldots,n.
\tag{1}
$$

Ordinary least squares (OLS) minimizes the residual sum of squares
$\sum_{i=1}^{n}(y_i - \hat{y}_i)^2$. With dozens of correlated survey
items, OLS tends to overfit, inflate coefficient variance, and produce
unstable signs under multicollinearity [@james2013; @mcneish2015].
Ridge regression adds an $L_2$ penalty that shrinks coefficients but
rarely sets them exactly to zero [@hoerl1970]. LASSO replaces that
penalty with an $L_1$ term, enabling simultaneous shrinkage and variable
selection [@tibshirani1996; @hesterberg2008].

### LASSO objective and tuning

LASSO estimates coefficients by minimizing the penalized least-squares
objective [@tibshirani1996; @santosa1986]:

$$
\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}.
\tag{2}
$$

The tuning parameter $\lambda \geq 0$ controls the bias–variance
trade-off. When $\lambda = 0$, Equation (2) reduces to OLS; as
$\lambda$ increases, more coefficients are driven to exactly zero,
yielding a sparse, interpretable model [@efron2004; @dalalyan2017].
We select $\lambda$ with 10-fold cross-validation on an 80/20
train–test split, reporting both **`lambda.min`** (lowest CV error) and
**`lambda.1se`** (simplest model within one standard error of the minimum)
[@james2013]. For the main paper result we emphasize plain LASSO at
`lambda.1se` as a deliberate parsimony choice: it retains nearly the
same test $R^2$ while selecting far fewer predictors.

Test-set performance for continuous models is summarized by

$$
R^2 = 1 - \frac{\sum_{i}(y_i - \hat{y}_i)^2}{\sum_{i}(y_i - \bar{y})^2},
\qquad
\mathrm{RMSE} = \sqrt{\frac{1}{n}\sum_{i}(y_i - \hat{y}_i)^2}.
\tag{3}
$$

### Plain LASSO versus group LASSO

We fit two complementary LASSO variants implemented in **`glmnet`**
(plain) and **`grpreg`** (group) [@james2013; @huang2024].

**Plain LASSO** (`alpha = 1`, `family = "gaussian"`) performs item-level
selection: any individual survey item may enter or leave the model.
**Group LASSO** assigns all items within a scale (e.g., SAPS) to a single
group and selects entire scales at once [@yuan2006; @huang2024]. Plain
LASSO answers which specific items matter; group LASSO answers which
whole instruments matter when scales are treated as coherent units.

For classification of SPAI item 12, we additionally fit multinomial
LASSO (`family = "multinomial"`) and refit a multinomial log-linear
model on the selected predictors (post-LASSO) to improve class
probabilities [@james2013]. Class imbalance in SPAI item 12 is evaluated
with balanced accuracy and Cohen's $\kappa$ on the held-out test set.

### Justification for this problem and data

Our CAT item bank contains on the order of 100 item-level predictors that
measure overlapping facets of problematic phone use [@gao2024]. Pearson
correlations among scale totals exceed $r = 0.64$ for several pairs (see
Data Exploration), so multicollinearity is expected rather than
exceptional. LASSO is well suited to high-dimensional, collinear
psychological survey data because it stabilizes estimation and returns a
sparse subset of items or scales that best predict addiction severity
[@mcneish2015; @dalalyan2017]. Comparing plain and group LASSO, alongside
`lambda.min` and `lambda.1se`, lets us balance predictive accuracy against
interpretability—an important consideration when communicating results to
clinicians and educators [@james2013; @huang2024].

## Analysis and Results

### Data Exploration and Visualization

Data come from a Computer Adaptive Testing (CAT) item bank on problematic
mobile phone use [@gao2024]. Respondents completed items from six
instrument families—nomophobia (NMP), smartphone addiction proneness
(SAPS), the Smartphone Addiction Inventory (SPAI) [@lin2014], the Mobile
Phone Addiction Scale (MPAS), the Mobile Phone Addiction Tendency Scale
(MPATS), and the Smartphone Addiction Scale (SAS-C and SAS-CA)—along with
demographics (age, gender, grade). The raw file contains
$n = 1{,}619$ respondents with complete records on all variables; no
listwise missingness remains after filtering. Scale totals are computed as
the sum of item scores within each instrument (higher totals indicate more
problematic use, per the item bank codebook).

The sample is predominantly young adults (mean age $\approx 18.8$ years,
$SD \approx 2.9$), with more female ($n = 991$) than male ($n = 628$)
respondents. SPAI total—the primary outcome for regression—has mean
$\approx 41.8$ ($SD \approx 10.2$) on the 20–80 scale, with a roughly
symmetric distribution centered below the midpoint. When scale totals are
rescaled to percentages of each instrument's maximum, **NMP** shows the
highest average endorsement ($\approx 69\%$ of maximum), whereas SAPS,
SPAI, MPAS, MPATS, and SAS cluster between 40% and 52%—suggesting that
fear of disconnection is reported more intensely than other addiction
symptoms in this cohort.

Pearson correlations with SPAI total rank the predictor scales as follows:
SAS ($r \approx 0.81$), SAPS ($r \approx 0.71$), MPATS ($r \approx 0.71$),
MPAS ($r \approx 0.65$), and NMP ($r \approx 0.55$). Scatter plots of SAPS
and SAS against SPAI total show strong linear trends with modest residual
spread. A correlation heatmap further reveals that all six scale totals
move together (inter-scale $r$ often exceeds 0.60), confirming substantial
**convergent validity** among instruments but also highlighting the
multicollinearity that motivates LASSO regularization.

Two patterns warrant attention before modeling. First, **SPAI item 12**
(perceived impact on study/work) is highly imbalanced: levels 1–4 contain
262, 577, 688, and 92 responses respectively, so the rarest category is
only $\approx 6\%$ of the sample—classification models must account for
this skew. Second, although NMP averages highest in percentage terms, it
correlates **weakest** with SPAI total among the five non-SPAI scales, hinting
that nomophobia may be partially redundant once proneness (SAPS) and SAS
symptoms are considered—a pattern consistent with preliminary LASSO results
where NMP items are often zeroed out. Demographic boxplots show modest SPAI
differences by gender and grade but no single grade level dominates severity
scores.

```{r viz-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(patchwork)

dat <- read.csv("data/DATA_Problematic Mobile Phone Use CAT.csv")
dat <- dat[complete.cases(dat), ]

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$SPAI_total  <- rowSums(dat[, paste0("SPAI_",  1:20)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)
```

```{r fig-spai-dist, fig.cap="Distribution of SPAI total score (primary outcome; n = 1,619 complete cases)."}
ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  annotate("text",
           x = mean(dat$SPAI_total) + 1.5, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Mean = ", round(mean(dat$SPAI_total), 1)),
           color = "#C44E52", size = 3.8) +
  labs(
    title = "Distribution of SPAI Total Score",
    x = "SPAI Total (20–80)", y = "Count"
  )
```

```{r fig-scale-boxplots, fig.cap="Scale totals rescaled to percent of each instrument maximum for comparability."}
scale_max <- c(NMP = 80, SAPS = 50, SPAI = 80, MPAS = 55,
               MPATS = 80, SAS = 150)

long_scales <- dat %>%
  select(NMP_total, SAPS_total, SPAI_total, MPAS_total,
         MPATS_total, SAS_total) %>%
  pivot_longer(everything(), names_to = "Scale", values_to = "Score") %>%
  mutate(
    Scale = str_remove(Scale, "_total"),
    Pct = Score / scale_max[Scale] * 100,
    Scale = factor(Scale, levels = names(scale_max))
  )

ggplot(long_scales, aes(x = Scale, y = Pct, fill = Scale)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.6, outlier.alpha = 0.4) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "Score Distributions Across All Six Scales",
    x = NULL, y = "Score (% of maximum)"
  )
```

```{r fig-correlations, fig.cap="Pearson correlations between non-SPAI scale totals and SPAI total."}
corr_df <- dat %>%
  summarise(
    NMP   = cor(NMP_total,   SPAI_total),
    SAPS  = cor(SAPS_total,  SPAI_total),
    MPAS  = cor(MPAS_total,  SPAI_total),
    MPATS = cor(MPATS_total, SPAI_total),
    SAS   = cor(SAS_total,   SPAI_total)
  ) %>%
  pivot_longer(everything(), names_to = "Scale", values_to = "r") %>%
  mutate(Scale = fct_reorder(Scale, r))

ggplot(corr_df, aes(x = Scale, y = r, fill = r)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = round(r, 2)), hjust = -0.15, fontface = "bold") +
  scale_fill_gradient(low = "#9ecae1", high = "#08519c", guide = "none") +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(title = "Pearson Correlations with SPAI Total", x = NULL, y = "Pearson r")
```

```{r fig-corr-heatmap, fig.cap="Pearson correlation heatmap among scale totals (complete cases)."}
corr_cols <- c("NMP_total", "SAPS_total", "SPAI_total", "MPAS_total",
               "MPATS_total", "SAS_total")
corr_mat <- cor(dat[, corr_cols], use = "complete.obs")

corr_long <- as.data.frame(corr_mat) %>%
  mutate(Scale1 = rownames(.)) %>%
  pivot_longer(-Scale1, names_to = "Scale2", values_to = "r") %>%
  mutate(
    Scale1 = gsub("_total", "", Scale1),
    Scale2 = gsub("_total", "", Scale2)
  )

ggplot(corr_long, aes(x = Scale2, y = Scale1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3, color = "black") +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r",
       title = "Scale Total Correlations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig-scatter-saps, fig.cap="SAPS total versus SPAI total with OLS trend line and 95% CI band."}
ggplot(dat, aes(x = SAPS_total, y = SPAI_total)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "#C44E52",
              fill = "#f4a7a9", linewidth = 1.1) +
  annotate("text",
           x = max(dat$SAPS_total) * 0.05,
           y = max(dat$SPAI_total) * 0.96, hjust = 0,
           label = paste0("r = ", round(cor(dat$SAPS_total, dat$SPAI_total), 2)),
           size = 4.5, fontface = "bold", color = "#C44E52") +
  labs(
    title = "SAPS Total vs SPAI Total",
    x = "SAPS Total (Addiction Proneness)", y = "SPAI Total (Addiction Severity)"
  )
```

```{r fig-demographics, fig.cap="Age distribution and demographic breakdown (complete cases).", fig.width=10, fig.height=4}
gender_labels <- c("1" = "Male", "2" = "Female")
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)],
                         levels = unname(gender_labels))

p_age <- ggplot(dat, aes(x = age)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(title = "Age", x = "Years", y = "Count")

p_gender <- dat %>% count(gender_lab) %>%
  ggplot(aes(x = gender_lab, y = n, fill = gender_lab)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Gender", x = NULL, y = "Count")

p_age + p_gender + plot_layout(widths = c(2, 1))
```

Together, these exploratory plots motivate our modeling strategy: SPAI
total is suitable as a continuous Gaussian outcome; the strong SAPS and SAS
relationships suggest they will dominate LASSO selection; and the shared
variance across scales justifies penalized regression over unregularized OLS
[@james2013; @gao2024].

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
