---
title: "Lasso Regression on Cellphone Use Addiction "
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: visual
---

Slides: [slides.html](slides.html){target="_blank"}

## Introduction

In today's ever-growing digital era, what began as a tool of
convenience, smartphones and devices have become a vital piece of our
daily life. While these items enable instant communication across the
globe, growing usage has evolved into an increasing psychological and
social crisis: smartphone addiction. Characterized by the compulsive
need to check notifications and apps, withdrawal symptoms, and severe
interference with daily life, smartphone dependency shows an increasing
link to rising rates of anxiety, sleep deprivation, and diminished
academic performance among young adults [@lin2014]. Identifying the
precise behavioral, social, and psychological drivers of this addiction
is critical for developing effective interventions. Available data
tracking screen time, app usage, and psychological metrics frequently
contain dozens of interconnected variables. Traditional linear
regression models struggle under this weight, frequently from
overfitting and multicollinearity, which obscures the true narrative
hidden within the data [@james2013].

In this project, we will explore the Least Absolute Shrinkage and
Selection Operator (LASSO) method, commonly known as LASSO regression,
to isolate the most critical predictors of smartphone addiction.
Introduced by Robert Tibshirani [-@tibshirani1996], LASSO is a powerful
statistical method designed for both regularization and variable
selection. Traditional ordinary least squares (OLS) regression treats
every available variable as important, which often introduces noise and
reduces predictive accuracy when dealing with large datasets. LASSO
solves this problem by adding a mathematical constraint—known as an L1
penalty—to the sum of the absolute values of the regression coefficients
[@tibshirani1996].

This L1 penalty performs a unique function: it actively shrinks the
coefficients of less important variables toward zero [@james2013].
Notably, unlike other methods such as Ridge regression, LASSO can force
coefficients to become exactly zero [@tibshirani1996]. LASSO
automatically performs variable selection, completely removing
irrelevant or redundant features from the final model, resulting in a
highly interpretable, sparse model consisting only of the most
statistically significant predictors.

Our dataset explores problematic smartphone usage from survey data
collected in China using a Computer Adaptive Testing (CAT) item bank
[@gao2024]. It is comprised of 1,619 respondents with complete data on
six instrument families plus the demographics.

Here we will apply LASSO regression to a comprehensive smartphone
addiction dataset to break down which specific behaviors and
demographics truly drive the dependencies. Smartphone datasets are
particularly noisy, with variables like social media screen time, daily
notification counts, and self-reported anxiety scores often overlapping
and distorting traditional models. By implementing LASSO, we can
mathematically weed out any statistical noise, isolating the leading
factors [@james2013]. Ideally, this approach will aid us in answering a
crucial question: with the countless digital metrics tracked today,
which factors truly lead to smartphone addictions? Our findings aim to
provide a streamlined, data-driven framework that peers and clinicians
alike can use to understand and combat digital dependency.

| Abbreviation | Scale                                             |
|--------------|---------------------------------------------------|
| NMP          | Nomophobia                                        |
| SAPS         | Smartphone Addiction Proneness                    |
| SPAI         | Smartphone Addiction Inventory                    |
| MPAS         | Mobile Phone Addiction Scale                      |
| MPATS        | Mobile Phone Addiction Tendency Scale             |
| SAS          | Smartphone Addiction Scale (SAS_C / SAS_CA items) |

## Methods

We model smartphone addiction severity with penalized linear regression.
Each respondent contributes item-level responses from six validated
instrument families (NMP, SAPS, SPAI, MPAS, MPATS, SAS) plus
demographics. After listwise deletion of incomplete records,
$n = 1{,}619$ complete cases remain. The primary continuous outcome is
**SPAI total**, the sum of 20 SPAI items (range 20–80) [@lin2014]. SPAI
items are excluded from the predictor set when predicting SPAI total to
avoid circularity; we instead ask whether nomophobia, proneness, and
related scales track overall SPAI severity. A secondary categorical
outcome, **SPAI item 12**, captures perceived negative impact on study
or work on a four-point Likert scale.

### Ordinary least squares and why LASSO

A standard linear model for continuous outcome $y_i$ and predictors
$\mathbf{x}_i \in \mathbb{R}^p$ is

$$y_i = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} + \varepsilon_i,
\qquad i = 1,\ldots,n.
\tag{1}$$

Ordinary least squares (OLS) minimizes the residual sum of squares
$\sum_{i=1}^{n}(y_i - \hat{y}_i)^2$. With dozens of correlated survey
items, OLS tends to overfit, inflate coefficient variance, and produce
unstable signs under multicollinearity [@james2013; @mcneish2015]. Ridge
regression adds an $L_2$ penalty that shrinks coefficients but rarely
sets them exactly to zero [@hoerl1970]. LASSO replaces that penalty with
an $L_1$ term, enabling simultaneous shrinkage and variable selection
[@tibshirani1996; @hesterberg2008].

### LASSO objective and tuning

LASSO estimates coefficients by minimizing the penalized least-squares
objective [@tibshirani1996; @santosa1986]:

$$\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}.
\tag{2}$$

The tuning parameter $\lambda \geq 0$ controls the bias–variance
trade-off. When $\lambda = 0$, Equation (2) reduces to OLS; as $\lambda$
increases, more coefficients are driven to exactly zero, yielding a
sparse, interpretable model [@efron2004; @dalalyan2017]. We select
$\lambda$ with 10-fold cross-validation on an 80/20 train–test split,
reporting both **`lambda.min`** (lowest CV error) and **`lambda.1se`**
(simplest model within one standard error of the minimum) [@james2013].
For the main paper result we emphasize plain LASSO at `lambda.1se` as a
deliberate parsimony choice: it retains nearly the same test $R^2$ while
selecting far fewer predictors.

Test-set performance for continuous models is summarized by

$$R^2 = 1 - \frac{\sum_{i}(y_i - \hat{y}_i)^2}{\sum_{i}(y_i - \bar{y})^2},
\qquad
\mathrm{RMSE} = \sqrt{\frac{1}{n}\sum_{i}(y_i - \hat{y}_i)^2}.
\tag{3}$$

### Iterative LASSO variable reduction

Rather than fitting a single LASSO model and choosing between lambda.min
and lambda.1se, we used an iterative, sequential-refitting procedure in
two phases.

Phase 1: Initial selection and stability check. LASSO (glmnet, alpha =
1, Gaussian family) was fit using 10-fold CV, with coefficients
extracted at lambda.min. Predictors with a nonzero coefficient were
retained and all others were dropped. We then refit LASSO on the
surviving predictors, again at lambda.min. Because the penalty was held
constant, this refit was not expected to drop further predictors. Its
purpose was to confirm that the first fit had already removed the noise
variables and to let the remaining coefficients re-adjust now that the
eliminated predictors no longer competed for shared variance.

Phase 2: Escalation. We then switched to lambda.1se, which by
construction is more aggressive about zeroing out coefficients, and
refit on the Phase 1 predictor set. From there we continued refitting
the surviving predictors at successively larger multiples of lambda.1se.
The size of each successive multiplier was chosen adaptively. After each
refit, we evaluated test RMSE, MAE, R², and MAPE. Only then decided
whether a further, larger penalty was worth trying, since the effect of
additional shrinkage on accuracy could not be anticipated in advance.
This loop continued until a refit produced a simultaneous, meaningful
change in RMSE, MAE, R², and MAPE relative to the preceding step, which
we treated as the signal to stop further reduction.

Model Selection: Every candidate model produced across both phases, not
just the last one fit, were compared on test RMSE, MAE, R², and MAPE.
The candidate with the strongest metrics overall was treated as the
leading model. Where two or more candidates had near-equivalent metrics,
we selected the most parsimonious of that group (i.e., the one with
fewer predictors), on the reasoning that a simpler model is preferable
when it sacrifices little to no measurable accuracy. Which candidate
this rule selected, and its exact predictor count and performance
metrics, are reported in the Results section.

### Post-LASSO OLS on top predictors

To assess how much predictive power could be retained with an even more
minimal item set, we additionally identified, from the selected LASSO
model, the five predictors with the largest absolute standardized
coefficients, and refit that five-predictor subset using ordinary least
squares (OLS) on the same train/test split. This follow-up analysis
addresses a practical question distinct from model selection itself: how
many survey items would be needed to approximate the full model's
predictive accuracy in a short- form instrument. Which items were
selected and how much variance they explained are reported in the
Results section.

### Justification for this problem and data

Our CAT item bank contains on the order of 100 item-level predictors
that measure overlapping facets of problematic phone use [@gao2024].
Pearson correlations among scale totals exceed $r = 0.64$ for several
pairs (see Data Exploration), so multicollinearity is expected rather
than exceptional. LASSO is well suited to high-dimensional, collinear
psychological survey data because it stabilizes estimation and returns a
sparse subset of items or scales that best predict addiction severity
[@mcneish2015; @dalalyan2017]. Comparing models reduced by `lambda.min`
and `lambda.1se`, lets us balance predictive accuracy against
interpretability—an important consideration when communicating results
to clinicians and educators [@james2013; @huang2024].

## Analysis and Results

### Data Exploration and Visualization

Data come from a Computer Adaptive Testing (CAT) item bank on
problematic mobile phone use [@gao2024]. Respondents completed items
from six instrument families—nomophobia (NMP), smartphone addiction
proneness (SAPS), the Smartphone Addiction Inventory (SPAI) [@lin2014],
the Mobile Phone Addiction Scale (MPAS), the Mobile Phone Addiction
Tendency Scale (MPATS), and the Smartphone Addiction Scale (SAS-C and
SAS-CA)—along with demographics (age, gender, grade). The raw file
contains $n = 1{,}619$ respondents with complete records on all
variables; no listwise missingness remains after filtering. Scale totals
are computed as the sum of item scores within each instrument (higher
totals indicate more problematic use, per the item bank codebook).

The sample is predominantly young adults (mean age $= 18.8$ years,
$SD = 2.9$), with more female ($n = 991$) than male ($n = 628$)
respondents. SPAI total—the primary outcome for regression—has mean
$= 41.8$ ($SD = 10.2$) on the 20–80 scale, with a roughly symmetric
distribution centered below the midpoint.

```{r viz-setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(patchwork)

dat <- read.csv("cellphone.csv")
dat <- dat[complete.cases(dat), ]

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$SPAI_total  <- rowSums(dat[, paste0("SPAI_",  1:20)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)
```

```{r fig-spai-dist, fig.cap="Distribution of SPAI total score (primary outcome; n = 1,619 complete cases)."}
ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  annotate("text",
           x = mean(dat$SPAI_total) + 1.5, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Mean = ", round(mean(dat$SPAI_total), 1)),
           color = "#C44E52", size = 3.8) +
  labs(
    title = "Distribution of SPAI Total Score",
    x = "SPAI Total (20–80)", y = "Count"
  )
```

Pearson correlations with SPAI total rank the predictor scales as
follows: SAS ($r = 0.81$), SAPS ($r = 0.71$), MPATS ($r = 0.71$), MPAS
($r = 0.65$), and NMP ($r = 0.55$). The scatter plot below shows the
strongest of these relationships—SAS against SPAI total—with a strong
linear trend and modest residual spread.

```{r fig-scatter-sas, fig.cap="SAS total versus SPAI total with OLS trend line and 95% CI band."}
ggplot(dat, aes(x = SAS_total, y = SPAI_total)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "#C44E52",
              fill = "#f4a7a9", linewidth = 1.1) +
  annotate("text",
           x = max(dat$SAS_total) * 0.05,
           y = max(dat$SPAI_total) * 0.96, hjust = 0,
           label = paste0("r = ", round(cor(dat$SAS_total, dat$SPAI_total), 2)),
           size = 4.5, fontface = "bold", color = "#C44E52") +
  labs(
    title = "SAS Total vs SPAI Total",
    x = "SAS Total (Smartphone Addiction Scale)", y = "SPAI Total (Addiction Severity)"
  )
```

All six scale totals move together (inter-scale $r$ often exceeds 0.60),
confirming substantial **convergent validity** among instruments but
also highlighting the multicollinearity that motivates LASSO
regularization.

```{r fig-corr-heatmap, fig.cap="Pearson correlation heatmap among scale totals (complete cases)."}
corr_cols <- c("NMP_total", "SAPS_total", "SPAI_total", "MPAS_total",
               "MPATS_total", "SAS_total")
corr_mat <- cor(dat[, corr_cols], use = "complete.obs")

corr_long <- as.data.frame(corr_mat) %>%
  mutate(Scale1 = rownames(.)) %>%
  pivot_longer(-Scale1, names_to = "Scale2", values_to = "r") %>%
  mutate(
    Scale1 = gsub("_total", "", Scale1),
    Scale2 = gsub("_total", "", Scale2)
  )

ggplot(corr_long, aes(x = Scale2, y = Scale1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3, color = "black") +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r",
       title = "Scale Total Correlations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Two patterns warrant attention before modeling. First, **SPAI item 12**
(perceived impact on study/work) is highly imbalanced: levels 1–4
contain 262, 577, 688, and 92 responses respectively, so the rarest
category is only $5.7\%$ of the sample—classification models must
account for this skew. Second, although NMP shows the highest average
endorsement when rescaled to a percentage of maximum ($= 68.8\%$), it
correlates **weakest** with SPAI total among the five non-SPAI scales,
hinting that nomophobia may be partially redundant once proneness (SAPS)
and SAS symptoms are considered—a pattern consistent with preliminary
LASSO results where NMP items are often zeroed out. The figure below
summarizes the age and gender composition of the sample.

```{r fig-demographics, fig.cap="Age distribution and demographic breakdown (complete cases).", fig.width=10, fig.height=4}
gender_labels <- c("1" = "Male", "2" = "Female")
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)],
                         levels = unname(gender_labels))

p_age <- ggplot(dat, aes(x = age)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(title = "Age", x = "Years", y = "Count")

p_gender <- dat %>% count(gender_lab) %>%
  ggplot(aes(x = gender_lab, y = n, fill = gender_lab)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Gender", x = NULL, y = "Count")

p_age + p_gender + plot_layout(widths = c(2, 1))
```

Together, these exploratory plots motivate our modeling strategy: SPAI
total is suitable as a continuous Gaussian outcome; the strong SAPS and
SAS relationships suggest they will dominate LASSO selection; and the
shared variance across scales justifies penalized regression over
unregularized OLS [@james2013; @gao2024].

### Modeling and Results

#### Data preprocessing

We began with the raw CAT item bank file ($n = 1{,}619$ rows, 101
variables) and retained only complete cases; no listwise missingness
required exclusion. The primary outcome **SPAI total** was computed as
the sum of 20 SPAI items. All other SPAI item columns were removed from
the predictor matrix when modeling SPAI total to avoid circularity.
Demographics (age, gender, grade) and item-level responses from NMP,
SAPS, MPAS, MPATS, and SAS (SAS-C and SAS-CA) entered as predictors.
Gender and grade were coded as factors; all item responses were treated
as numeric. Predictors were assembled with a model matrix (no intercept
column in the penalized fit). Respondents were split 80/20 into training
and test sets. All continuous models below use the same split.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
#

library(glmnet)
library(dplyr)

###################################################
#
#      Reduces from full model if 81 Variables down to 52 using Lasso Lambda.min
#
##################################################


# ===============================
# 1. LOAD DATA
# ===============================
cells_data <- read.csv("cellphone.csv", header = TRUE)

# ===============================
# 2. CLEAN GRADE and Gender
# ===============================
grade_map <- c(
  "1" = "F", "2" = "Sph", "3" = "J", "4" = "Sen", 
  "5" = "M1", "6" = "M2", "7" = "M3", "8" = "Doc", 
  "12" = "8th", "13" = "9th", "14" = "10th", "15" = "11th", "16" = "12th"
)
cells_data$grade <- factor(grade_map[as.character(cells_data$grade)])

# ADD THIS LINE TO FIX GENDER:
cells_data$gender <- factor(cells_data$gender)

# ===============================
# 3. DEFINE OUTCOME (TOTAL SPAI ADDICTION)
# ===============================
# Generate names for all 20 SPAI items
spai_all <- paste0("SPAI_", 1:20)

# Convert all 20 columns to numeric
cells_data[spai_all] <- lapply(cells_data[spai_all], function(x) as.numeric(as.character(x)))

# MODIFIED: Calculate the sum of ALL 20 SPAI variables
cells_data$sum_SPAI <- rowSums(cells_data[, spai_all], na.rm = TRUE)

# MODIFIED: Set sum_SPAI as the new dependent variable
y <- cells_data$sum_SPAI

# ===============================
# 4. PREDICTORS (NO SPAI LEAKAGE)
# ===============================
# MODIFIED: Drop the new target variable (sum_SPAI) and all individual SPAI columns 
x <- cells_data %>% dplyr::select(-sum_SPAI, -all_of(spai_all))

# Safety check: Ensure no lingering SPAI or target columns exist in your predictors
stopifnot(!any(grepl("SPAI", names(x))))

# ===============================
# 5. FIXED: MODEL MATRIX BEFORE SPLIT
# ===============================
x_full_mat <- model.matrix(~ . - 1, data = x)

# ===============================
# 6. TRAIN / TEST SPLIT
# ===============================
set.seed(123)
n <- nrow(x_full_mat)
train_index <- sample(1:n, size = 0.8 * n)

x_train_mat <- x_full_mat[train_index, ]
x_test_mat  <- x_full_mat[-train_index, ]
y_train     <- y[train_index]
y_test      <- y[-train_index]

# ===============================
# 7. LASSO (TRAIN ONLY)
# ===============================
set.seed(123)
cv_lasso <- cv.glmnet(
  x = x_train_mat, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# ===============================
# 8. TRAIN VS TEST EVALUATION (UPDATED WITH NMAE & MAPE)
# ===============================
pred_train <- as.numeric(predict(cv_lasso, s = "lambda.min", newx = x_train_mat))
pred_test  <- as.numeric(predict(cv_lasso, s = "lambda.min", newx = x_test_mat))

rmse_train <- sqrt(mean((y_train - pred_train)^2))
mae_train  <- mean(abs(y_train - pred_train))
r2_train   <- 1 - (sum((y_train - pred_train)^2) / sum((y_train - mean(y_train))^2))

rmse_test  <- sqrt(mean((y_test - pred_test)^2))
mae_test   <- mean(abs(y_test - pred_test))
r2_test    <- 1 - (sum((y_test - pred_test)^2) / sum((y_test - mean(y_test))^2))

# FIXED: Removed the duplicate uncorrected line beneath this one
cv_rmse    <- sqrt(cv_lasso$cvm[cv_lasso$lambda == cv_lasso$lambda.min])

# ADDED: Calculation for Scale-Agnostic Percentage Metrics 
range_y_test   <- max(y_test) - min(y_test)
norm_mae_test  <- (mae_test / range_y_test) * 100
mape_test      <- mean(abs((y_test - pred_test) / y_test)) * 100


# ==============================================================================
# ADDITIONAL CODE: PLOT FIT AND EXTRACT TOP PREDICTORS (TOTAL SPAI MODEL)
# ==============================================================================

# ------------------------------------------------------------------------------
# 1. VISUALIZE MODEL FIT (ACTUAL VS. PREDICTED)
# ------------------------------------------------------------------------------
plot_data <- data.frame(
  Actual = y_test,
  Predicted = pred_test
)
# Determine coordinate boundaries for a perfectly square 1:1 plot
min_val <- min(c(plot_data$Actual, plot_data$Predicted), na.rm = TRUE)
max_val <- max(c(plot_data$Actual, plot_data$Predicted), na.rm = TRUE)

plot(
  x = plot_data$Actual, 
  y = plot_data$Predicted,
  xlab = "Actual Total SPAI Score",
  ylab = "Predicted Total SPAI Score",
  pch = 16,
  col = rgb(0.2, 0.4, 0.6, 0.5),
  xlim = c(min_val, max_val),
  ylim = c(min_val, max_val)
)

abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

legend(
  "topleft",
  legend = "Perfect Prediction Line",
  col = "red",
  lty = 2,
  lwd = 2
)
```

The figure shows predicted versus actual Total SPAI scores from a
10-fold cross-validated LASSO regression model. The model began with 81
candidate predictors and retained 52 predictors using the lambda-min
penalty. The dashed reference line represents perfect prediction. There
is a strong positive linear correlation and a good fit with test RMSE =
5.1702, test MAE = 3.9714 and test MAPE = 10.58 percent. The optimal
lambda = 0.074363.

Lasso regression was preformed one more time on the remaining 52
variables using lambda.min again. Even though the optimal lambda for
this model is 0.020216, no variables were dropped which shows that the
remaining variables have sufficient information to contribute to the
model such that lambda.min will not eliminate any additional variables.
The test RMSE = 5.2142, test MAE = 3.9902 and the test MAPE = 10.6
percent. Because this is the best that lambda.min can do and because
there are still 52 variables being used as predictors, lambda.1se will
be used to reduce the model even further.

Lambda.min is the exact value of $\lambda$ that results in the lowest
possible mean squared error (MSE) during cross-validation. The goal of
lambda.min is to maximize the prediction accuracy which means that it is
less aggressive at dropping variables. Thus variables with noise or
little contribution to the model will often remain in the model leading
to a large and often complex model. Since the focus of Lasso Regression
is to minimize the number of predictor variables while maintaining
accuracy, lambda.1se was used for the next reduction. Lambda.1se is the
largest value of $\lambda$ whose cross-validation error is still within
one standard error of the absolute minimum error. This will force more
variables that have minimum contribution to the model to be dropped and
yielding a simpler model that gives similar results to lambda.min.

Using lambda.1se = 0.227095 reduced the model from 52 predictor
variables down to 33 predictor variables. That is a 19 variable drop in
the model which is the exact goal of Lasso Regression. While this Lasso
created a simpler, less complex model, the test RMSE improved slightly
to 5.1626; the test R2 improved slightly to 0.7498 and the Normalized
Test MAE remained unchanged through this reduction. Based on this
information, this is the best performing model yet and it is the least
complex model so far as it has the lowest number of predictor variables.

Another Lasso Regression was preformed to see if there are still
predictors that can be removed without sacrificing the accuracy of the
model. This time the lambda used was 1.75\*lambda.1se which forces even
more predictor variables to zero. With this penalty, the model was
reduced from 33 predictors down to 28 predictors. The test RMSE
increased slightly to 5.2553, the test MAE slightly increased to 4.1075
and the test MAPE = 11.21 percent. The optimal lambda is 0.576584. Even
though there was a slight increase in the test RMSE and MAE, another
Lasso was performed to evaluate the number of variables that would be
reduced at the cost of losing accuracy.

This sixth Lasso Regression brought the total variables down to 22.
While the test RMSE, test MAE and test MAPE all had slight increases,
this was not sufficient to stop another Lasso Regression as the simplest
model with the best accuracy is the objective. The lambda for the sixth
Lasso Regression is 1.195034, test RMSE = 5.5280, test MAE = 4.1075
which is unchanged from the last lasso and test MAPE = 12.14.

The seventh Lasso Regression reduced the model from 22 predictor
variables to 19 predictor variables but the most important change was in
the R2 value which dropped below 70% on the test data for the first
time. Due to the drop of R2 and the increase in the test RMSE, test MAE
and test MAPE, the threshold benefit of dropping variables at the cost
of accuracy has been reached and no more reduction of predictor
variables will be done on the model. Here is the summary table of all of
the models:

```{r}
# ------------------------------------------------------------------------------
# 2. EXTRACT AND CLEAN TOP PREDICTORS
# ------------------------------------------------------------------------------
# Extract Lasso coefficients at the optimal lambda (lambda.min)
coef_lasso <- coef(cv_lasso, s = "lambda.min")

# Convert to a clean matrix format
coef_matrix <- as.matrix(coef_lasso)
coef_df <- data.frame(
  Dummy_Variable = rownames(coef_matrix),
  Coefficient = coef_matrix[, 1],
  Absolute_Impact = abs(coef_matrix[, 1]),
  stringsAsFactors = FALSE
)

# Remove the Intercept and filter to only selected features (non-zero)
selected_features <- coef_df %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

# Combine everything into a clean presentation table with exact string matching
first_top_predictors <- selected_features %>%
  rowwise() %>%
  mutate(Original_Variable = {
    # Find which original column names match the start of the dummy name
    matches <- names(x)[sapply(names(x), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    # Pick the longest match to avoid sub-string confusion
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)


# MODIFIED: Saved with a unique filename specific to this new total SPAI outcome
write.csv(first_top_predictors, "total_spai_first_predictors.csv", row.names = FALSE)




# ===============================
# 9. OUTPUT RESULTS TABLE (UPDATED HORIZONTAL LAYOUT)
# ===============================
# 1. Build the base vertical results data frame
results_vertical <- data.frame(
  Metric = c("RMSE", "MAE", "R-Squared"),
  Train  = c(round(rmse_train, 4), round(mae_train, 4), round(r2_train, 4)),
  Test   = c(round(rmse_test, 4), round(mae_test, 4), round(r2_test, 4))
)

# 2. Transpose the metrics to make them columns and include scale-agnostic metrics
results_horizontal <- data.frame(
  Model = "First_Lasso_Reduction",
  OVR = length(unique(first_top_predictors$Original_Variable)),
  `Train RMSE` = results_vertical[1, "Train"],
  `Test RMSE`  = results_vertical[1, "Test"],
  `Train MAE`   = results_vertical[2, "Train"],
  `Test MAE`    = results_vertical[2, "Test"],
  `Train R-Squared` = results_vertical[3, "Train"],
  `Test R-Squared`  = results_vertical[3, "Test"],
  `Test Norm MAE (%)` = round(norm_mae_test, 2),
  `Test MAPE (%)`     = round(mape_test, 2),
  `Optimal Lambda`    = round(cv_lasso$lambda.min, 6), # ADDED: Lambda column
  check.names = FALSE
)


# ==============================================================================
# DISPLAY MODEL SIZE AND TOP 5 PREDICTOR VARIABLES
# ==============================================================================
# Calculate the count of unique original variables with non-zero coefficients
total_surviving_vars <- length(unique(first_top_predictors$Original_Variable))


####################################################################
###############################################################################
#
#       Second Lasso  lambda.min again on 52 to modify coefficients and still has 52 after this run
#
###############################################################################
####################################################################


# ==============================================================================
# 1. EXTRACT SELECTED VARIABLE NAMES (REMOVING DUMMY FACTOR SUFFIXES)
# ==============================================================================
# Get unique names of the original variables that survived the first round
remaining_vars <- unique(first_top_predictors$Original_Variable)

# ==============================================================================
# 2. SUBSET THE ORIGINAL X PREDICTORS
# ==============================================================================
# Create a new predictor dataframe containing only the surviving variables
x_reduced <- cells_data %>% dplyr::select(all_of(remaining_vars))

# ==============================================================================
# 3. GENERATE NEW MODEL MATRIX & SPLIT DATA
# ==============================================================================
x_reduced_mat <- model.matrix(~ . - 1, data = x_reduced)

# Re-use the exact same training indexes from your first split for consistency
x_train_mat2 <- x_reduced_mat[train_index, ]
x_test_mat2  <- x_reduced_mat[-train_index, ]
# Note: y_train and y_test remain exactly the same as your sum_SPAI vectors

# ==============================================================================
# 4. RUN SECOND LASSO MODEL
# ==============================================================================
set.seed(123) # Reset seed for reproducibility in cross-validation folds
cv_lasso2 <- cv.glmnet(
  x = x_train_mat2, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# ==============================================================================
# 6. EXTRACT, CLEAN, AND SAVE SECOND REDUCTION VARIABLES
#    (MOVED UP: Must create 'second_top_predictors' before evaluating performance)
# ==============================================================================
coef_lasso2  <- coef(cv_lasso2, s = "lambda.min")
coef_matrix2 <- as.matrix(coef_lasso2)

coef_df2 <- data.frame(
  Dummy_Variable = rownames(coef_matrix2),
  Coefficient = coef_matrix2[, 1],
  Absolute_Impact = abs(coef_matrix2[, 1]),
  stringsAsFactors = FALSE
)

selected_features2 <- coef_df2 %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

second_top_predictors <- selected_features2 %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_reduced)[sapply(names(x_reduced), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)


# Saved with a unique filename specific to this second total SPAI reduction
write.csv(second_top_predictors, "total_spai_second_predictors.csv", row.names = FALSE)

# ==============================================================================
# 5. EVALUATE SECOND MODEL PERFORMANCE (UPDATED WITH NMAE, MAPE, & LAMBDA)
# ==============================================================================
pred_train2 <- as.numeric(predict(cv_lasso2, s = "lambda.min", newx = x_train_mat2))
pred_test2  <- as.numeric(predict(cv_lasso2, s = "lambda.min", newx = x_test_mat2))

# Compute baseline metrics for vertical extraction
mae_train2      <- mean(abs(y_train - pred_train2)) # ADDED FOR MASTER TABLE CONSISTENCY
mae_test2       <- mean(abs(y_test - pred_test2))
range_y_test    <- max(y_test) - min(y_test)
norm_mae_test2  <- (mae_test2 / range_y_test) * 100
mape_test2      <- mean(abs((y_test - pred_test2) / y_test)) * 100
lambda_used2    <- cv_lasso2$lambda.min

# 1. Build the base vertical results data frame
results2_vertical <- data.frame(
  Metric = c("RMSE", "MAE", "R-Squared"),
  Train  = c(round(sqrt(mean((y_train - pred_train2)^2)), 4), round(mae_train2, 4), round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4)),
  Test   = c(round(sqrt(mean((y_test - pred_test2)^2)), 4), round(mae_test2, 4), round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4))
)

# 2. Transpose the metrics to make them columns with the correct OVR mapping
results2_horizontal <- data.frame(
  Model = "Second_Lasso_Reduction",
  OVR = length(unique(second_top_predictors$Original_Variable)),
  `Train RMSE` = results2_vertical[1, "Train"],
  `Test RMSE`  = results2_vertical[1, "Test"],
  `Train MAE`   = results2_vertical[2, "Train"],
  `Test MAE`    = results2_vertical[2, "Test"],
  `Train R-Squared` = results2_vertical[3, "Train"],
  `Test R-Squared`  = results2_vertical[3, "Test"],
  `Test Norm MAE (%)` = round(norm_mae_test2, 2), 
  `Test MAPE (%)`     = round(mape_test2, 2),     
  `Optimal Lambda`    = round(lambda_used2, 6),   
  check.names = FALSE
)


#------------------------------------------------------------------
#   Comparison Tables Code
#------------------------------------------------------------------

# ==============================================================================
# 1. CALCULATE VARIABLE COUNTS FOR EACH STAGE
# ==============================================================================
initial_var_count <- ncol(x)
first_var_count   <- length(unique(first_top_predictors$Original_Variable))
second_var_count  <- length(unique(second_top_predictors$Original_Variable))

# ==============================================================================
# 2. RUN BASELINE MULTIPLE LINEAR REGRESSION (FOR COMPARISON PERFORMANCE)
# ==============================================================================
baseline_model <- lm(y_train ~ ., data = as.data.frame(x_train_mat))

pred_train_base <- predict(baseline_model, newdata = as.data.frame(x_train_mat))
pred_test_base  <- predict(baseline_model, newdata = as.data.frame(x_test_mat))

# Baseline metrics (and new context variables)
rmse_tr_b  <- sqrt(mean((y_train - pred_train_base)^2, na.rm=TRUE))
rmse_te_b  <- sqrt(mean((y_test - pred_test_base)^2, na.rm=TRUE))
mae_tr_b   <- mean(abs(y_train - pred_train_base), na.rm=TRUE) # ADDED
mae_te_b   <- mean(abs(y_test - pred_test_base), na.rm=TRUE)
r2_tr_b    <- 1 - (sum((y_train - pred_train_base)^2, na.rm=TRUE) / sum((y_train - mean(y_train))^2))
r2_te_b    <- 1 - (sum((y_test - pred_test_base)^2, na.rm=TRUE) / sum((y_test - mean(y_test))^2))
norm_mae_b <- (mae_te_b / range_y_test) * 100
mape_b     <- mean(abs((y_test - pred_test_base) / y_test), na.rm=TRUE) * 100

# ==============================================================================
# 3. CONSTRUCT THE MASTER COMPARISON TABLE (UPDATED FOR ALL METRICS)
# ==============================================================================
master_comparison <- data.frame(
  Metric = c(
    "OVR", 
    "Train RMSE", 
    "Test RMSE", 
    "Train MAE",         # ADDED
    "Test MAE",          # ADDED
    "Train R-Squared", 
    "Test R-Squared",
    "Test Norm MAE (%)", 
    "Test MAPE (%)",     
    "Optimal Lambda"     
  ),
  Baseline_Full_Model = c(
    initial_var_count, 
    round(rmse_tr_b, 4), 
    round(rmse_te_b, 4), 
    round(mae_tr_b, 4),  # ADDED
    round(mae_te_b, 4),  # ADDED
    round(r2_tr_b, 4), 
    round(r2_te_b, 4),
    round(norm_mae_b, 2), 
    round(mape_b, 2),     
    NA                    
  ),
  First_Lasso_Reduction = c(
    first_var_count, 
    round(rmse_train, 4), 
    round(rmse_test, 4), 
    round(mae_train, 4),  # ADDED (Assumed variables exist from step 1)
    round(mae_test, 4),   # ADDED
    round(r2_train, 4), 
    round(r2_test, 4),
    round(norm_mae_test, 2),
    round(mape_test, 2),
    round(cv_lasso$lambda.min, 6)
  ),
  Second_Lasso_Reduction = c(
    second_var_count, 
    round(sqrt(mean((y_train - pred_train2)^2)), 4), 
    round(sqrt(mean((y_test - pred_test2)^2)), 4), 
    round(mae_train2, 4), # ADDED
    round(mae_test2, 4),  # ADDED
    round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4), 
    round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4),
    round(norm_mae_test2, 2),
    round(mape_test2, 2),
    round(lambda_used2, 6)
  ),
  stringsAsFactors = FALSE
)

# ==============================================================================
# 4. TRANSPOSE, PRINT, AND SAVE THE SUMMARY TABLE
# ==============================================================================
# Clean transposition keeping matrix string issues isolated
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Strip default sequence index numbering from print layout
row.names(transposed_comparison) <- NULL


write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)




#______________________________________________________________________________________
#######################################################################################
#
#
#  New Lasso to actually do a second reduction  Modified lambda to lambda.1se Reduces from 52 to 33
#
#
#######################################################################################
#______________________________________________________________________________________




# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 3 MODELS BEFORE APPENDING
# ==============================================================================
# CORRECTED: Ensures the matrix object clears any downstream models cleanly
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:4]
}

# ==============================================================================
# 1. EXTRACT COEFFICIENTS AND COUNT VARIABLES FOR LAMBDA.1SE
# ==============================================================================
# Extract Lasso coefficients using the stricter 1-standard-error rule
coef_lasso_1se  <- coef(cv_lasso, s = "lambda.1se")
coef_matrix_1se <- as.matrix(coef_lasso_1se)

coef_df_1se <- data.frame(
  Dummy_Variable = rownames(coef_matrix_1se),
  Coefficient = coef_matrix_1se[, 1],
  Absolute_Impact = abs(coef_matrix_1se[, 1]),
  stringsAsFactors = FALSE
)

# Filter out the intercept and zeroed-out variables
selected_features_1se <- coef_df_1se %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0)

# Clean variable names to count unique original variables
top_predictors_1se <- selected_features_1se %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x)[sapply(names(x), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup()

se1_var_count <- length(unique(top_predictors_1se$Original_Variable))

# ==============================================================================
# 2. EVALUATE LAMBDA.1SE MODEL PERFORMANCE (FIXED: COMPUTING ALL REQUIRED METRICS)
# ==============================================================================
# Generate predictions using the 1se penalty strength
pred_train_1se <- as.numeric(predict(cv_lasso, s = "lambda.1se", newx = x_train_mat))
pred_test_1se  <- as.numeric(predict(cv_lasso, s = "lambda.1se", newx = x_test_mat))

# Compute metrics
rmse_tr_1se   <- sqrt(mean((y_train - pred_train_1se)^2))
rmse_te_1se   <- sqrt(mean((y_test - pred_test_1se)^2))
mae_tr_1se    <- mean(abs(y_train - pred_train_1se))
mae_te_1se    <- mean(abs(y_test - pred_test_1se))
r2_tr_1se     <- 1 - (sum((y_train - pred_train_1se)^2) / sum((y_train - mean(y_train))^2))
r2_te_1se     <- 1 - (sum((y_test - pred_test_1se)^2) / sum((y_test - mean(y_test))^2))

# Compute new horizontal metrics for the 1se model
range_y_test      <- max(y_test) - min(y_test)
norm_mae_test_1se <- (mae_te_1se / range_y_test) * 100
mape_test_1se     <- mean(abs((y_test - pred_test_1se) / y_test)) * 100

# ==============================================================================
# 3. CONSTRUCT THE MASTER COMPARISON TABLE (FIXED VARIABLES SYNCHRONIZATION)
# ==============================================================================
master_comparison <- data.frame(
  Metric = c(
    "OVR", 
    "Train RMSE", 
    "Test RMSE", 
    "Train MAE",         
    "Test MAE",          
    "Train R-Squared", 
    "Test R-Squared",
    "Test Norm MAE (%)", 
    "Test MAPE (%)",     
    "Optimal Lambda"     
  ),
  Baseline_Full_Model = c(
    initial_var_count, 
    round(rmse_tr_b, 4), 
    round(rmse_te_b, 4), 
    round(mae_tr_b, 4),  
    round(mae_te_b, 4),  
    round(r2_tr_b, 4), 
    round(r2_te_b, 4),
    round(norm_mae_b, 2), 
    round(mape_b, 2),     
    NA                    
  ),
  First_Lasso_Reduction = c(
    first_var_count, 
    round(rmse_train, 4), 
    round(rmse_test, 4), 
    round(mae_train, 4),  
    round(mae_test, 4),   
    round(r2_train, 4), 
    round(r2_test, 4),
    round(norm_mae_test, 2),
    round(mape_test, 2),
    round(cv_lasso$lambda.min, 6)
  ),
  Second_Lasso_Reduction = c(
    second_var_count, 
    round(sqrt(mean((y_train - pred_train2)^2)), 4), 
    round(sqrt(mean((y_test - pred_test2)^2)), 4), 
    round(mae_train2, 4), 
    round(mae_test2, 4),  
    round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4), 
    round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4),
    round(norm_mae_test2, 2),
    round(mape_test2, 2),
    round(lambda_used2, 6)
  ),
  Lambda_1se_Parsimonious = c(
    se1_var_count,                                                  # OVR
    round(rmse_tr_1se, 4),                                          # Train RMSE
    round(rmse_te_1se, 4),                                          # Test RMSE
    round(mae_tr_1se, 4),                                           # Train MAE
    round(mae_te_1se, 4),                                           # Test MAE
    round(r2_tr_1se, 4),                                            # Train R-Squared
    round(r2_te_1se, 4),                                            # Test R-Squared
    round(norm_mae_test_1se, 2),                                    # Test Norm MAE (%)
    round(mape_test_1se, 2),                                        # Test MAPE (%)
    round(cv_lasso$lambda.1se, 6)                                   # Optimal Lambda (1se)
  ),
  stringsAsFactors = FALSE
)

# ==============================================================================
# 4. TRANSPOSE, PRINT, AND SAVE THE SUMMARY TABLE
# ==============================================================================
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Ensure numeric columns print correctly and don't convert back to raw characters
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL



write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)


#-----------------------------------------------------------------------------------------------------

#####################################################################################################
######################################################################################################
#
#      Third Lasso Reduction  Modified Lambda = 1.75Lambda.1se  Reduces from 33 to 28 
#
#####################################################################################################
#####################################################################################################



# ==============================================================================
# 0. ENVIRONMENT RESET: TRUNCATE TABLE TO THE INITIAL 4 MODELS BEFORE APPENDING
# ==============================================================================
# Clears out any lingering downstream models from your active R workspace session
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:5]
}

# ==============================================================================
# 1. EXTRACT THE SURVIVING VARIABLES FROM THE 1SE MODEL
# ==============================================================================
# Dynamically reads the variable count from your 1se total SPAI model
vars_1se <- unique(top_predictors_1se$Original_Variable)
x_strict_df <- cells_data %>% dplyr::select(all_of(vars_1se))
x_strict_mat <- model.matrix(~ . - 1, data = x_strict_df)

# Align with your original training and testing splits
x_train_strict <- x_strict_mat[train_index, ]
x_test_strict  <- x_strict_mat[-train_index, ]

# ==============================================================================
# 2. RUN A HIGH-PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_strict <- cv.glmnet(
  x = x_train_strict, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# Manually amplify the 1se lambda by 1.75x to force aggressive variable dropping
strict_lambda <- cv_strict$lambda.1se * 1.75

# ==============================================================================
# 3. EXTRACT AND CLEAN THE ULTRA-REDUCED PREDICTORS
# ==============================================================================
coef_strict  <- coef(cv_strict, s = strict_lambda)
matrix_strict <- as.matrix(coef_strict)

df_strict <- data.frame(
  Dummy_Variable = rownames(matrix_strict),
  Coefficient = matrix_strict[, 1],
  Absolute_Impact = abs(matrix_strict[, 1]),
  stringsAsFactors = FALSE
)

selected_strict <- df_strict %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

strict_top_predictors <- selected_strict %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_strict_df)[sapply(names(x_strict_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

strict_var_count <- length(unique(strict_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE STRICT MODEL PERFORMANCE (UPDATED FOR NEW METRICS)
# ==============================================================================
pred_train_st <- as.numeric(predict(cv_strict, s = strict_lambda, newx = x_train_strict))
pred_test_st  <- as.numeric(predict(cv_strict, s = strict_lambda, newx = x_test_strict))

# Core Error Metrics
rmse_tr_st <- sqrt(mean((y_train - pred_train_st)^2))
rmse_te_st <- sqrt(mean((y_test - pred_test_st)^2))
mae_tr_st  <- mean(abs(y_train - pred_train_st))
mae_te_st  <- mean(abs(y_test - pred_test_st))
r2_tr_st   <- 1 - (sum((y_train - pred_train_st)^2) / sum((y_train - mean(y_train))^2))
r2_te_st   <- 1 - (sum((y_test - pred_test_st)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics (ADDED)
range_y_test     <- max(y_test) - min(y_test)
norm_mae_test_st <- (mae_te_st / range_y_test) * 100
mape_test_st     <- mean(abs((y_test - pred_test_st) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON
# ==============================================================================
# Append all 10 metrics cleanly to match master dataframe dimensions
master_comparison$Strict_Hyper_Parsimonious <- c(
  strict_var_count,             # OVR
  round(rmse_tr_st, 4),         # Train RMSE
  round(rmse_te_st, 4),         # Test RMSE
  round(mae_tr_st, 4),          # Train MAE
  round(mae_te_st, 4),          # Test MAE
  round(r2_tr_st, 4),           # Train R-Squared
  round(r2_te_st, 4),           # Test R-Squared
  round(norm_mae_test_st, 2),   # Test Norm MAE (%)
  round(mape_test_st, 2),       # Test MAPE (%)
  round(strict_lambda, 6)       # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Keep formatting clean across numeric properties
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the updated transposed spreadsheet version to your working files
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(strict_top_predictors, "total_spai_strict_predictors.csv", row.names = FALSE)



######################################################################
#
#
#   I am going to do a fourth reduction  from 28 to 22 Variables.
#
#
######################################################################
#

# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 5 MODELS BEFORE APPENDING
# ==============================================================================
# Clears out any lingering downstream terminal models from your active R session
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:6]
}

# ==============================================================================
# 1. EXTRACT THE 28 VARIABLES FROM THE PREVIOUS STRICT MODEL
# ==============================================================================
vars_strict_28 <- unique(strict_top_predictors$Original_Variable)
x_ultra_df     <- cells_data %>% dplyr::select(all_of(vars_strict_28))
x_ultra_mat    <- model.matrix(~ . - 1, data = x_ultra_df)

x_train_ultra <- x_ultra_mat[train_index, ]
x_test_ultra  <- x_ultra_mat[-train_index, ]

# ==============================================================================
# 2. RUN AN ULTRA-HIGH PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_ultra <- cv.glmnet(
  x = x_train_ultra, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

ultra_lambda <- cv_ultra$lambda.1se * 2.50

# ==============================================================================
# 3. EXTRACT AND CLEAN THE NEW ULTRA-REDUCED PREDICTORS
# ==============================================================================
coef_ultra   <- coef(cv_ultra, s = ultra_lambda)
matrix_ultra <- as.matrix(coef_ultra)

df_ultra <- data.frame(
  Dummy_Variable = rownames(matrix_ultra),
  Coefficient = matrix_ultra[, 1],
  Absolute_Impact = abs(matrix_ultra[, 1]),
  stringsAsFactors = FALSE
)

selected_ultra <- df_ultra %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

ultra_top_predictors <- selected_ultra %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_ultra_df)[sapply(names(x_ultra_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

ultra_var_count <- length(unique(ultra_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE ULTRA-STRICT MODEL PERFORMANCE (UPDATED FOR ALL METRICS)
# ==============================================================================
pred_train_ultra <- as.numeric(predict(cv_ultra, s = ultra_lambda, newx = x_train_ultra))
pred_test_ultra  <- as.numeric(predict(cv_ultra, s = ultra_lambda, newx = x_test_ultra))

# Core Error Metrics
rmse_tr_ult <- sqrt(mean((y_train - pred_train_ultra)^2))
rmse_te_ult <- sqrt(mean((y_test - pred_test_ultra)^2))
mae_tr_ult  <- mean(abs(y_train - pred_train_ultra))
mae_te_ult  <- mean(abs(y_test - pred_test_ultra))
r2_tr_ult   <- 1 - (sum((y_train - pred_train_ultra)^2) / sum((y_train - mean(y_train))^2))
r2_te_ult   <- 1 - (sum((y_test - pred_test_ultra)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics (ADDED)
range_y_test       <- max(y_test) - min(y_test)
norm_mae_test_ult  <- (mae_te_ult / range_y_test) * 100
mape_test_ult      <- mean(abs((y_test - pred_test_ultra) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON (FIXED SYNCHRONIZATION)
# ==============================================================================
# Append all 10 metrics to seamlessly align with master dataframe dimensions
master_comparison$Ultra_Strict_Reduction <- c(
  ultra_var_count,              # OVR
  round(rmse_tr_ult, 4),         # Train RMSE
  round(rmse_te_ult, 4),         # Test RMSE
  round(mae_tr_ult, 4),          # Train MAE
  round(mae_te_st, 4),           # Test MAE
  round(r2_tr_ult, 4),           # Train R-Squared
  round(r2_te_ult, 4),           # Test R-Squared
  round(norm_mae_test_ult, 2),   # Test Norm MAE (%)
  round(mape_test_ult, 2),       # Test MAPE (%)
  round(ultra_lambda, 6)         # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Convert character matrices back to numeric for perfect console alignment
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the updated transposed spreadsheet version to your working files
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(ultra_top_predictors, "total_spai_ultra_strict_predictors.csv", row.names = FALSE)



#################----------------------#######################------------------##############
#
#       Fifth Lasso Reduced from 22 to 19 Variables
#
#################----------------------#######################------------------##############

# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 6 MODELS BEFORE APPENDING
# ==============================================================================
# CORRECTED: Safely retains your 6 foundational columns plus Ultra_Strict (7 columns total)
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:7]
}

# ==============================================================================
# 1. EXTRACT THE 22 VARIABLES FROM THE ULTRA-STRICT MODEL
# ==============================================================================
# Extract the unique names of variables that survived the ultra-strict pass
vars_ultra_22 <- unique(ultra_top_predictors$Original_Variable)
x_term_df     <- cells_data %>% dplyr::select(all_of(vars_ultra_22))
x_term_mat    <- model.matrix(~ . - 1, data = x_term_df)

# Align with your original training and testing splits for consistency
x_train_term <- x_term_mat[train_index, ]
x_test_term  <- x_term_mat[-train_index, ]

# ==============================================================================
# 2. RUN A TERMINAL MAXIMUM-PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_term <- cv.glmnet(
  x = x_train_term, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# Amplify the penalty multiplier to 3.5x to aggressively drop to the absolute core
term_lambda <- cv_term$lambda.1se * 3.50

# ==============================================================================
# 3. EXTRACT AND CLEAN THE TERMINAL PREDICTORS
# ==============================================================================
coef_term   <- coef(cv_term, s = term_lambda)
matrix_term <- as.matrix(coef_term)

df_term <- data.frame(
  Dummy_Variable = rownames(matrix_term),
  Coefficient = matrix_term[, 1],
  Absolute_Impact = abs(matrix_term[, 1]),
  stringsAsFactors = FALSE
)

selected_term <- df_term %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

term_top_predictors <- selected_term %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_term_df)[sapply(names(x_term_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

term_var_count <- length(unique(term_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE TERMINAL MODEL PERFORMANCE (UPDATED FOR ALL METRICS)
# ==============================================================================
pred_train_term <- as.numeric(predict(cv_term, s = term_lambda, newx = x_train_term))
pred_test_term  <- as.numeric(predict(cv_term, s = term_lambda, newx = x_test_term))

# Core Error Metrics
rmse_tr_term <- sqrt(mean((y_train - pred_train_term)^2))
rmse_te_term <- sqrt(mean((y_test - pred_test_term)^2))
mae_tr_term  <- mean(abs(y_train - pred_train_term))
mae_te_term  <- mean(abs(y_test - pred_test_term))
r2_tr_term   <- 1 - (sum((y_train - pred_train_term)^2) / sum((y_train - mean(y_train))^2))
r2_te_term   <- 1 - (sum((y_test - pred_test_term)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics
range_y_test      <- max(y_test) - min(y_test)
norm_mae_test_term <- (mae_te_term / range_y_test) * 100
mape_test_term     <- mean(abs((y_test - pred_test_term) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON
# ==============================================================================
# Append all 10 metrics cleanly to match master dataframe dimensions exactly
master_comparison$Terminal_Core_Reduction <- c(
  term_var_count,               # OVR
  round(rmse_tr_term, 4),        # Train RMSE
  round(rmse_te_term, 4),        # Test RMSE
  round(mae_tr_term, 4),         # Train MAE
  round(mae_te_term, 4),         # Test MAE
  round(r2_tr_term, 4),          # Train R-Squared
  round(r2_te_term, 4),          # Test R-Squared
  round(norm_mae_test_term, 2),  # Test Norm MAE (%)
  round(mape_test_term, 2),      # Test MAPE (%)
  round(term_lambda, 6)          # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Convert character matrices back to numeric for perfect console alignment
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the final clean wide format summary spreadsheet and the core predictors
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(term_top_predictors, "total_spai_terminal_core_predictors.csv", row.names = FALSE)
###################################

###################################
#
#   HTML-friendly LASSO Model Comparison Table
#
###################################

library(knitr)
library(kableExtra)

# Create cleaner names
html_model_comparison <- transposed_comparison %>%
  dplyr::select(
    Model,
    OVR,
    `Train RMSE`,
    `Test RMSE`,
    `Train MAE`,
    `Test MAE`,
    `Train R-Squared`,
    `Test R-Squared`,
    `Test Norm MAE (%)`,
    `Test MAPE (%)`,
    `Optimal Lambda`
  ) %>%
  dplyr::mutate(
    Model = dplyr::recode(
      Model,
      "Baseline_Full_Model" = "Baseline",
      "First_Lasso_Reduction" = "LASSO 1",
      "Second_Lasso_Reduction" = "LASSO 2",
      "Lambda_1se_Parsimonious" = "Lambda 1SE",
      "Strict_Hyper_Parsimonious" = "Strict",
      "Ultra_Strict_Reduction" = "Ultra",
      "Terminal_Core_Reduction" = "Terminal"
    )
  )

# Shorter column headings
colnames(html_model_comparison) <- c(
  "Model",
  "OVR",
  "Train RMSE",
  "Test RMSE",
  "Train MAE",
  "Test MAE",
  "Train R²",
  "Test R²",
  "Norm MAE (%)",
  "MAPE (%)",
  "Lambda"
)

# Render HTML table
html_model_comparison %>%
  kable(
    format = "html",
    digits = 3,
    caption = "LASSO Regression Model Comparison (Total SPAI)"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = TRUE,
    position = "center",
    font_size = 14
  )


```

Based on the test RMSE, test MAE, test R2, Test Normalized MAE and test
MAPE, the model with 33 predictor variables is the best model of all to
predict Total SPAI Score. Model 33 has the test RMSE = 5.1626 which is
the lowest error rate of all models considered so far. The model also
explains 74.98% of the variance in unseen test data. This model has also
seen a 59.25% reduction in complexity when compared to the baseline
model which is very significant. By removing those 48 variables the
overall performance of the model significantly increased. However, by
removing any additional variables, the prediction accuracy of the model
is reduced so the overall best performing model is the model with 33
variables.

Below are the variables that were selected in the optimal model with the
most influential at the top and displayed in decreasing order as the
influence decreases. :

```{r}
# Load ggplot2 for a high-quality visualization
library(ggplot2)

# ==============================================================================
# 1. ISOLATE AND RANK THE WINNING 33 PREDICTORS
# ==============================================================================
# MODIFIED: Pulls from top_predictors_1se to capture the 33-variable model
final_ranked_predictors <- top_predictors_1se %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  # Recalculate absolute impact to ensure proper ranking execution
  mutate(Absolute_Impact = abs(Coefficient)) %>%
  arrange(desc(Absolute_Impact)) %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient, Absolute_Impact)

# # Print the full ranked list to your console
# cat("\n==================================================================\n")
# cat("          FINAL RANKED PREDICTORS (LAMBDA 1SE MODEL)               \n")
# cat("==================================================================\n")
# print(final_ranked_predictors, n = Inf, row.names = FALSE)

# MODIFIED: Saved with a unique filename specific to this final 33-variable set
write.csv(final_ranked_predictors, "total_spai_final_33_predictors.csv", row.names = FALSE)

# ==============================================================================
# 2. GENERATE A COEFFICIENT IMPACT VISUALIZATION
# ==============================================================================
# Create a clear color flag for positive vs. negative coefficients
plot_df <- final_ranked_predictors %>%
  mutate(
    Direction = ifelse(Coefficient > 0, "Increases Addiction", "Decreases Addiction"),
    # Reorder the dummy variables so the chart ranks them from largest to smallest impact
    Dummy_Variable = reorder(Dummy_Variable, Absolute_Impact)
  )

# Build the publication-ready bar chart with centered titles
coef_plot <- ggplot(plot_df, aes(x = Dummy_Variable, y = Coefficient, fill = Direction)) +
  geom_bar(stat = "identity", width = 0.75, color = "black", linewidth = 0.2) +
  coord_flip() + 
  scale_fill_manual(values = c("Increases Addiction" = "#d95f02", "Decreases Addiction" = "#7570b3")) +
  labs(
    title = "Predictor Impact on Total SPAI Score",
    subtitle = "Final Parsimonious Lasso Model (33 Variables Remaining)",
    x = "Model Predictors (Dummy Variables)",
    y = "Lasso Coefficient Value (Impact Weight)",
    fill = "Effect Direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    # FIXED: Added hjust = 0.5 to perfectly center the title and subtitle
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5, margin = margin(b = 5)),
    plot.subtitle = element_text(color = "gray30", size = 10, hjust = 0.5, margin = margin(b = 15)),
    axis.text.y = element_text(size = 9, family = "mono"), 
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank() 
  )


# Display the plot in your RStudio viewer
print(coef_plot)
```

After performing the lasso regression and reducing the model to only 19
variables, additional analysis continued. With the optimal model
containing 33 questions that can be used to predict cellphone addiction,
Ordinary Least Squares was applied to just the top five influential
predictor variables from the 33 Variable Lasso Regression model as well
as the 52 lambda.min Lasso Regression model as the results from that
model were very close to the 33 variable model. An interesting
observation is that the top five variables from the 52 lambda.min model
were the exact same top five as the 22 variable model.

Here is a table that summarizes the results of all models:

```{r}
 #==============================================================================
# 1. ISOLATE TARGET AND SELECTED PREDICTORS (TOP 5 OF 22 CONFIGURATION)
# ==============================================================================
# Define the precise 5 features for the updated configuration
selected_vars_top5_22 <- c("SAPS_10", "SAPS_7", "MPATS4", "SAPS_9", "SAS_CA14")

# Extract features from your base dataframe
x_ols_df_22 <- cells_data %>% dplyr::select(all_of(selected_vars_top5_22))

# Split the isolated feature space into matching Train/Test matrices 
# using your script's active training split index
x_train_ols_22 <- x_ols_df_22[train_index, ]
x_test_ols_22  <- x_ols_df_22[-train_index, ]

# Combine with target arrays into clean training data frames for lm()
train_data_ols_22 <- data.frame(y = y_train, x_train_ols_22)
test_data_ols_22  <- data.frame(y = y_test, x_test_ols_22)

# ==============================================================================
# 2. FIT THE MULTIPLE LINEAR REGRESSION (OLS) MODEL
# ==============================================================================
ols_five_model_22 <- lm(y ~ ., data = train_data_ols_22)

# # ==============================================================================
# 3. EVALUATE MODEL PERFORMANCE METRICS
# ==============================================================================
# Generate numeric vector predictions for test and training pools
pred_train_ols5_22 <- as.numeric(predict(ols_five_model_22, newdata = train_data_ols_22))
pred_test_ols5_22  <- as.numeric(predict(ols_five_model_22, newdata = test_data_ols_22))

# Calculate Core Error metrics
rmse_tr_ols5_22 <- sqrt(mean((y_train - pred_train_ols5_22)^2))
rmse_te_ols5_22 <- sqrt(mean((y_test - pred_test_ols5_22)^2))
mae_tr_ols5_22  <- mean(abs(y_train - pred_train_ols5_22))
mae_te_ols5_22  <- mean(abs(y_test - pred_test_ols5_22))
r2_tr_ols5_22   <- 1 - (sum((y_train - pred_train_ols5_22)^2) / sum((y_train - mean(y_train))^2))
r2_te_ols5_22   <- 1 - (sum((y_test - pred_test_ols5_22)^2) / sum((y_test - mean(y_test))^2))

# Calculate Context Metrics
range_y_test              <- max(y_test) - min(y_test)
norm_mae_test_ols5_22     <- (mae_te_ols5_22 / range_y_test) * 100
mape_test_ols5_22         <- mean(abs((y_test - pred_test_ols5_22) / y_test)) * 100

# ==============================================================================
# 4. PRINT FORMATTED PERFORMANCE RESULTS (EXCLUDING OPTIMAL LAMBDA)
# ==============================================================================
ols5_22_results_horizontal <- data.frame(
  Model = "Top_5_of_22_Model", # MODIFIED
  OVR = length(selected_vars_top5_22),
  `Train RMSE` = round(rmse_tr_ols5_22, 4),
  `Test RMSE`  = round(rmse_te_ols5_22, 4),
  `Train MAE`   = round(mae_tr_ols5_22, 4),
  `Test MAE`    = round(mae_te_ols5_22, 4),
  `Train R-Squared` = round(r2_tr_ols5_22, 4),
  `Test R-Squared`  = round(r2_te_ols5_22, 4),
  `Test Norm MAE (%)` = round(norm_mae_test_ols5_22, 2), 
  `Test MAPE (%)`     = round(mape_test_ols5_22, 2),     
  check.names = FALSE
)

# ==============================================================================
# 1. ISOLATE TARGET AND SELECTED PREDICTORS (MPATS4 EXCLUDED, SAS_C7 ADDED)
# ==============================================================================
# Define the precise 5 features for the updated configuration
selected_vars_new5 <- c("SAPS_10", "SAPS_7", "SAPS_9", "SAS_CA14", "SAS_C7")

# Extract features from your base dataframe
x_ols5_new_df <- cells_data %>% dplyr::select(all_of(selected_vars_new5))

# Split the isolated feature space into matching Train/Test matrices 
# using your script's active training split index
x_train_ols5_new <- x_ols5_new_df[train_index, ]
x_test_ols5_new  <- x_ols5_new_df[-train_index, ]

# Combine with target arrays into clean training data frames for lm()
train_data_ols5_new <- data.frame(y = y_train, x_train_ols5_new)
test_data_ols5_new  <- data.frame(y = y_test, x_test_ols5_new)

# ==============================================================================
# 2. FIT THE MULTIPLE LINEAR REGRESSION (OLS) MODEL
# ==============================================================================
ols_five_new_model <- lm(y ~ ., data = train_data_ols5_new)


# ==============================================================================
# 3. EVALUATE MODEL PERFORMANCE METRICS
# ==============================================================================
# Generate numeric vector predictions for test and training pools
pred_train_ols5_new <- as.numeric(predict(ols_five_new_model, newdata = train_data_ols5_new))
pred_test_ols5_new  <- as.numeric(predict(ols_five_new_model, newdata = test_data_ols5_new))

# Calculate Core Error metrics
rmse_tr_ols5_new <- sqrt(mean((y_train - pred_train_ols5_new)^2))
rmse_te_ols5_new <- sqrt(mean((y_test - pred_test_ols5_new)^2))
mae_tr_ols5_new  <- mean(abs(y_train - pred_train_ols5_new))
mae_te_ols5_new  <- mean(abs(y_test - pred_test_ols5_new))
r2_tr_ols5_new   <- 1 - (sum((y_train - pred_train_ols5_new)^2) / sum((y_train - mean(y_train))^2))
r2_te_ols5_new   <- 1 - (sum((y_test - pred_test_ols5_new)^2) / sum((y_test - mean(y_test))^2))

# Calculate Context Metrics
range_y_test           <- max(y_test) - min(y_test)
norm_mae_test_ols5_new <- (mae_te_ols5_new / range_y_test) * 100
mape_test_ols5_new     <- mean(abs((y_test - pred_test_ols5_new) / y_test)) * 100

# ==============================================================================
# 4. PRINT FORMATTED PERFORMANCE RESULTS (EXCLUDING OPTIMAL LAMBDA FOR OLS)
# ==============================================================================
ols5_new_results_horizontal <- data.frame(
  Model = "Top_5_of_33_Model", # MODIFIED
  OVR = length(selected_vars_new5),
  `Train RMSE` = round(rmse_tr_ols5_new, 4),
  `Test RMSE`  = round(rmse_te_ols5_new, 4),
  `Train MAE`   = round(mae_tr_ols5_new, 4),
  `Test MAE`    = round(mae_te_ols5_new, 4),
  `Train R-Squared` = round(r2_tr_ols5_new, 4),
  `Test R-Squared`  = round(r2_te_ols5_new, 4),
  `Test Norm MAE (%)` = round(norm_mae_test_ols5_new, 2), 
  `Test MAPE (%)`     = round(mape_test_ols5_new, 2),     
  check.names = FALSE
)
# ==============================================================================
# FINAL MASTER SPECIFICATION CONSOLIDATION (WEBSITE HTML FORMAT)
# ==============================================================================

library(knitr)
library(kableExtra)

mega_dynamic_comparison <- data.frame(
  Model = c(
    "Baseline",
    "LASSO 1",
    "LASSO 2",
    "Lambda 1SE",
    "Strict",
    "Ultra",
    "Terminal",
    "Top 5 (33)",
    "Top 5 (52/22)"
  ),
  Predictors = c(81, 52, 52, 33, 28, 22, 19, 5, 5),
  `Test RMSE` = c(5.2338, 5.1702, 5.2142, 5.1626, 5.2553, 5.5280, 5.8296, 6.0199, 6.2256),
  `Test MAE` = c(3.9987, 3.9714, 3.9902, 3.9910, 4.1075, 4.1075, 4.6312, 4.7114, 4.8361),
  `Test R²` = c(0.7429, 0.7491, 0.7448, 0.7498, 0.7407, 0.7131, 0.6810, 0.6598, 0.6362),
  `Norm MAE (%)` = c(7.41, 7.35, 7.39, 7.39, 7.61, 8.06, 8.58, 8.72, 8.96),
  `MAPE (%)` = c(10.59, 10.58, 10.60, 10.71, 11.21, 12.14, 13.08, 12.65, 12.91),
  stringsAsFactors = FALSE
)


# ==============================================================================
# WEBSITE HTML TABLE
# ==============================================================================

mega_dynamic_comparison %>%
  kable(
    format = "html",
    digits = 3,
    caption = "OLS and LASSO Model Performance Comparison (Total SPAI)"
  ) %>%
  kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "responsive"
    ),
    full_width = TRUE,
    position = "center",
    font_size = 14
  )


```

Both models with only five variables performed very well with the top
five variables from the 33 variable model accounting for 65.98% of the
variances of unseen data which is remarkable high for a real model using
only five variables.\
This study substantiates that if you were only able to conduct a short,
five question survey, that the five questions that you should ask are:

1.  Even when I think I should stop using my mobile phone, I will
    continue to use it and cannot stop.
2.  It has become a habit of me to spend a lot of time on my mobile
    phone.
3.  I feel very panic when I can’t use my mobile phone.
4.  I don’t get enough sleep and feel tired from using mobile phone too
    much.
5.  Playing mobile phone has affected my academic performance.

These questions are the strongest indicator of cell phone addition.

#### LASSO regression for SPAI total

We fit plain LASSO (`glmnet`, `alpha = 1`, Gaussian family) and group
LASSO (`grpreg`, scales as groups) on the training set with 10-fold
cross-validation to select $\lambda$. Test-set performance is reported
with $R^2$ and RMSE (Equation 3). Following our Methods section, we
emphasize **plain LASSO at `lambda.1se`** as the recommended model: it
achieves strong predictive accuracy while retaining far fewer predictors
than group LASSO at comparable $R^2$ [@james2013; @huang2024].

```{r modeling-setup, warning=FALSE, message=FALSE}
library(glmnet)
library(grpreg)
library(caret)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  mat <- as.matrix(raw_coef)
  vars <- rownames(mat)[mat[, 1] != 0]
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(
    R2 = 1 - ss_res / ss_tot,
    RMSE = sqrt(mean((y_true - y_pred)^2))
  )
}

mod_dat <- dat
mod_dat$gender <- as.factor(mod_dat$gender)
mod_dat$grade <- as.factor(mod_dat$grade)
mod_dat$SPAI_12 <- as.factor(mod_dat$SPAI_12)

train_index <- createDataPartition(mod_dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(mod_dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
```

On the held-out test set ($n = `r nrow(mm_reg$x) - length(train_index)`$),
plain LASSO at `lambda.1se` achieved $R^2 = `r gauss_results$R2[1]`$ and
RMSE $= `r gauss_results$RMSE[1]`$, selecting `r gauss_results$N_selected[1]`
predictors. Group LASSO at `lambda.min` reached a similar $R^2 = `r gauss_results$R2[2]`$
but retained `r gauss_results$N_selected[2]` predictors—substantially more
complexity for little additional explanatory gain. In practical terms, the
non-SPAI scales explain roughly three-quarters of the variation in SPAI total
severity on respondents the model had not seen during training.

```{r tbl-gaussian, tbl.cap="Gaussian LASSO models predicting SPAI total from non-SPAI items (test set)."}
kable(gauss_results, align = c("l", "r", "r", "r"),
      col.names = c("Model", "R²", "RMSE", "Predictors selected"))
```

Item-level plain LASSO selection was concentrated in **SAPS** (addiction
proneness) and **SAS** (cognitive and behavioral addiction symptoms).
Nomophobia (NMP) contributed few or no items once these stronger scales
were penalized into the model, consistent with the weaker zero-order
correlation observed in exploration. Demographics played a minor role
relative to the addiction instrument items.

```{r fig-selected-scales, fig.cap="Number of item-level predictors selected by plain LASSO (lambda.1se), grouped by instrument."}
ggplot(scale_counts, aes(x = reorder(Scale, N_items), y = N_items, fill = Scale)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Selected items",
       title = "Plain LASSO Selected Predictors by Scale") +
  theme_minimal(base_size = 12)



```

These results align with the exploratory correlation structure: SAS and
SAPS track SPAI total most closely, and LASSO uses that shared variance
efficiently while zeroing out redundant nomophobia items. The high test
$R^2$ should be interpreted as **convergent validity** among overlapping
addiction measures rather than evidence of causal effects [@gao2024;
@lin2014].

## Conclusion

This study applied iterative LASSO variable reduction to a Chinese
Computer Adaptive Testing item bank on problematic mobile phone use
[@gao2024] in order to predict continuous SPAI total severity from
non-SPAI instrument items and demographics, with SPAI items excluded
from the predictor set to avoid circularity. Beginning with 81
candidates, Phase 1 selection at `lambda.min` retained 52 predictors;
Phase 2 escalation to `lambda.1se` yielded a 33-predictor model that
achieved the most favorable balance of test error and complexity. That
preferred specification attained a test RMSE of 5.16 and a test $R^2$ of
0.75 on held-out respondents, corresponding to an approximate 59%
reduction in predictor count relative to the full baseline. Further
shrinkage to 28, 22, and 19 predictors degraded predictive accuracy—with
test $R^2$ falling below 0.70 at 19 predictors—and therefore activated
the stopping rule.

Selected items concentrated in SAPS (addiction proneness) and SAS
(cognitive and behavioral addiction symptoms), with a smaller
contribution from MPATS; NMP (nomophobia) items were largely driven to
zero, consistent with nomophobia’s weaker zero-order association with
SPAI total. A post-LASSO ordinary least squares refit on the five
largest-magnitude items retained a test $R^2$ of 0.66—well below the
33-item model’s 0.75, yet still high for a five-item instrument. Those
five items capture compulsive use, habitual time spent, panic upon
separation, sleep loss, and impaired academic performance, spanning loss
of control, salience, withdrawal, and life interference.

The elevated $R^2$ is most appropriately interpreted as convergent
validity among overlapping addiction instruments rather than as evidence
of causal effects [@gao2024; @lin2014; @james2013]. Inter-scale
correlations frequently exceeding $r = 0.60$ produce the
multicollinearity that motivated LASSO over unregularized ordinary least
squares [@mcneish2015]. Because the sample was collected in China among
predominantly young adults (mean age $= 18.8$ years) and contained more
female than male respondents, the findings should not be generalized to
other populations without external validation [@gao2024]. Practically,
the 33-item LASSO model—and, where brevity is essential, the five-item
OLS subset—can estimate SPAI-like severity without administering the
full multi-instrument battery, subject to prospective validation.

Future work should assess robustness across alternative train–test
splits and random seeds, validate the short-form item sets in
independent samples, revisit the imbalanced categorical outcome SPAI
item 12 with methods suited to class imbalance, and extend the analysis
beyond the Chinese young-adult cohort in order to distinguish
context-specific from more universal predictors of smartphone addiction
severity.

## References