---
title: "Lasso Regression on Cellphone Use Addiction "
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}

## Introduction

In today's ever-growing digital era, what began as a tool of convenience,
smartphones and devices have become a vital piece of our daily life. While
these items enable instant communication across the globe, growing usage has
evolved into an increasing psychological and social crisis: smartphone
addiction. Characterized by the compulsive need to check notifications and
apps, withdrawal symptoms, and severe interference with daily life,
smartphone dependency shows an increasing link to rising rates of anxiety,
sleep deprivation, and diminished academic performance among young adults
[@lin2014]. Identifying the precise behavioral, social, and psychological
drivers of this addiction is critical for developing effective interventions.
Available data tracking screen time, app usage, and psychological metrics
frequently contain dozens of interconnected variables. Traditional linear
regression models struggle under this weight, frequently from overfitting and
multicollinearity, which obscures the true narrative hidden within the data
[@james2013].

In this project, we will explore the Least Absolute Shrinkage and Selection
Operator (LASSO) method, commonly known as LASSO regression, to isolate the
most critical predictors of smartphone addiction. Introduced by Robert
Tibshirani [-@tibshirani1996], LASSO is a powerful statistical method designed
for both regularization and variable selection. Traditional ordinary least
squares (OLS) regression treats every available variable as important, which
often introduces noise and reduces predictive accuracy when dealing with large
datasets. LASSO solves this problem by adding a mathematical constraint—known
as an L1 penalty—to the sum of the absolute values of the regression
coefficients [@tibshirani1996].

This L1 penalty performs a unique function: it actively shrinks the
coefficients of less important variables toward zero [@james2013]. Notably,
unlike other methods such as Ridge regression, LASSO can force coefficients to
become exactly zero [@tibshirani1996]. LASSO automatically performs variable
selection, completely removing irrelevant or redundant features from the final
model, resulting in a highly interpretable, sparse model consisting only of the
most statistically significant predictors.

Our dataset explores problematic smartphone usage from survey data from a
Computer Adaptive Testing (CAT) item bank. It is comprised of 1,619
respondents with complete data on six instrument families plus the demographics.

Here we will apply LASSO regression to a comprehensive smartphone addiction
dataset to break down which specific behaviors and demographics truly drive the
dependencies. Smartphone datasets are particularly noisy, with variables like
social media screen time, daily notification counts, and self-reported anxiety
scores often overlapping and distorting traditional models. By implementing
LASSO, we can mathematically weed out any statistical noise, isolating the
leading factors [@james2013]. Ideally, this approach will aid us in answering a
crucial question: with the countless digital metrics tracked today, which
factors truly lead to smartphone addictions? Our findings aim to provide a
streamlined, data-driven framework that peers and clinicians alike can use to
understand and combat digital dependency.

| Abbreviation | Scale |
|--------------|-------|
| NMP | Nomophobia |
| SAPS | Smartphone Addiction Proneness |
| SPAI | Smartphone Addiction Inventory |
| MPAS | Mobile Phone Addiction Scale |
| MPATS | Mobile Phone Addiction Tendency Scale |
| SAS | Smartphone Addiction Scale (SAS_C / SAS_CA items) |

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
