---
title: "LASSO Regression on Cellphone Use Addiction"
subtitle: "IDC6940 Capstone — Problematic Mobile Phone Use"
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
course: Capstone Projects in Data Science
bibliography: references.bib
self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(glmnet)
library(grpreg)
library(caret)
library(knitr)
library(patchwork)



RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

theme_set(
  theme_minimal(base_size = 16) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  mat <- as.matrix(raw_coef)
  vars <- rownames(mat)[mat[, 1] != 0]
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(
    R2 = 1 - ss_res / ss_tot,
    RMSE = sqrt(mean((y_true - y_pred)^2))
  )
}

dat <- read.csv("cellphone.csv")
dat <- dat[complete.cases(dat), ]

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$SPAI_total  <- rowSums(dat[, paste0("SPAI_",  1:20)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

dat$gender <- as.factor(dat$gender)
dat$grade <- as.factor(dat$grade)
dat$SPAI_12 <- as.factor(dat$SPAI_12)

gender_labels <- c("1" = "Male", "2" = "Female")
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)],
                         levels = unname(gender_labels))

train_index <- createDataPartition(dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
```
## Project Overview {.smaller}

Using LASSO Regression to Predict Cell Phone Addiction

Purpose

- Investigate factors associated with cell phone addiction.
- Identify the most important predictors of addiction.
- Use LASSO regression to reduce the number of variables while maintaining predictive accuracy.

Research Question

Which behavioral and demographic factors are the strongest predictors of cell phone addiction?

## Introduction and Literature Review {.smaller}

Background

 - Smartphone use has increased dramatically worldwide.
 - Excessive use has been linked to:
 
     - Anxiety
     - Depression
     - Poor sleep quality
     - Lower academic performance
     - Reduced productivity

## Introduction and Literature Review {.smaller}

Previous studies consistently show that:

 - Younger individuals tend to have higher addiction scores.
 - Mental health factors play an important role.
 - Social media and entertainment apps are major contributors.
 - Many studies rely on traditional regression methods.

Few studies use LASSO regression to determine which predictors truly matter most when many variables are considered simultaneously. [@busch2021]

## What is LASSO? {.smaller}

LASSO Regression (Least Absolute Shrinkage and Selection Operator) is a machine learning technique that:

 - Builds a prediction model
 - Shrinks less important coefficients toward zero
 - Removes unnecessary variables
 - Produces a simpler, more interpretable model

## Why LASSO? {.smaller}

 - Prevents overfitting
 - Handles datasets with many predictors
 - Improves model interpretability
 - Selects only variables that contribute meaningful information

Simple Visualization:

Many Variables -> LASSO Penalty -> Only Important Variables Remain

## The Problem {.smaller}

There are many possible factors that may influence cell phone addiction:

 - Age
 - Gender
 - Daily phone use
 - Social media habits
 - Sleep
 - Mental health
 - Academic performance
 - Lifestyle factors


## The Problem {.smaller}

Smartphone use has shifted from convenience to dependency.

- Compulsive checking, withdrawal, daily-life interference
- Links to anxiety, sleep loss, and academic strain [@lin2014]
- Dozens of overlapping survey variables → multicollinearity [@james2013]

Why This Project Matters

 - Demonstrates an application of machine learning in behavioral research.
 - Identifies the strongest predictors of cell phone addiction.
 - Creates a model that is easier to interpret than traditional regression.
 - Results may help guide future interventions and prevention efforts.


## Our Data {.smaller}

Survey data collected in **China** using a Computer Adaptive Testing (CAT)
item bank [@gao2024] — **$n = 1{,}619$** complete cases

| Abbrev. | Scale |
|---------|-------|
| NMP | Nomophobia |
| SAPS | Smartphone addiction proneness |
| SPAI | Smartphone Addiction Inventory (outcome) |
| MPAS | Mobile Phone Addiction Scale |
| MPATS | Mobile Phone Addiction Tendency Scale |
| SAS | Smartphone Addiction Scale (SAS-C / SAS-CA) |

Plus age, gender, and grade (mean age $= 18.8$; 991 F / 628 M).

## Outcome & Approach {.smaller}

**Primary outcome:** SPAI total — sum of 20 SPAI items (20–80; mean $= 41.8$)

**Predictors:** NMP, SAPS, MPAS, MPATS, SAS items + demographics — SPAI items excluded (they define the score)

**Method:** LASSO [@tibshirani1996] — $L_1$ penalty shrinks weak coefficients to zero [@james2013; @hesterberg2008]


$$
\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}
$$


## A Two-Phase Iterative Procedure {.smaller}

::: columns
::: {.column width="50%"}
### Phase 1 — Select & Validate {.smaller}

- Fit LASSO at `lambda.min`; drop zero-coefficient predictors
- Refit the survivors at the **same** `lambda.min` confirming the first fit already removed the noise and lets coefficients re-adjust.
:::

::: {.column width="50%"}
### Phase 2 — Escalating Search {.smaller}

- Switch to `lambda.1se`, which drops predictors more aggressively and refit
- Repeat: raise the penalty to a **larger multiple** of `lambda.1se`, refit, and check metrics with each next multiple chosen only *after* seeing the prior refit's results
:::
:::


---

## Model Selection Rule {.smaller}

Every candidate fit across **both phases** is compared on test RMSE, MAE, R², and MAPE.

::: {}
- The candidate with the strongest metrics overall leads
- When two or more candidates score nearly the same, the model with fewer predictors is preferred, since a simpler model that costs little to no measurable accuracy is the better choice
- From the best model we pick within each phase, we want to look at the top 5 predictors
:::

---

## Model Evaluation Metrics {.smaller}

::: {}
- **R²** — Proportion of test set variance in SPAI explained by model
- **RMSE** — Root Mean Squared Error: average magnitude of prediction errors
- **MAE** — Mean Absolute Error: average magnitude of errors between prediction and actual values
- **MAPE** — Mean Absolute Percentage Error: average magnitude of errors expressed as percentage
- **Normalized MAE** - Normalized Mean Absolute Error: MAE divided by a normalized factor
- **OVR** - Overall Variable Remaining: The amount of predictor variables left after shrinkage

:::

---

## Tuning Criteria {.smaller}

How each refit was penalized


#### Penalty Selection {.smaller}

::: {}
**`lambda.min`**

The λ with lowest cross-validated error. Maximizes predictive accuracy. Less aggressive about zeroing out coefficients.

**`lambda.1se`**

The largest λ within one standard error of the minimum CV error. Smaller amount of accuracy. Preferred model for shrinkage.

**Increasing multiples of `lambda.1se`**

Applied on later refits, each multiple chosen adaptively after reviewing the prior refit's metrics.
:::

---

## SPAI Severity in the Sample {.smaller}

```{r fig-spai-dist}
#| fig-width: 9
#| fig-height: 4.5

ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  annotate("text",
           x = mean(dat$SPAI_total) + 1.5, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Mean = ", round(mean(dat$SPAI_total), 1)),
           color = "#C44E52", size = 4) +
  labs(
    title = "Distribution of SPAI Total Score",
    x = "SPAI Total (20–80)", y = "Count"
  )
```


## Data Preprocessing {.smaller}

- $n = 1{,}619$ complete cases; no listwise missingness
- SPAI total = sum of 20 SPAI items; other SPAI items removed from predictors
- Model matrix from NMP, SAPS, MPAS, MPATS, SAS items + demographics
- 80/20 train–test split



## Model Fit: Actual vs. Predicted Total SPAI Score {.smaller}

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
#

library(glmnet)
library(dplyr)

###################################################
#
#      Reduces from full model if 81 Variables down to 52 using Lasso Lambda.min
#
##################################################


# ===============================
# 1. LOAD DATA
# ===============================
cells_data <- read.csv("cellphone.csv", header = TRUE)

# ===============================
# 2. CLEAN GRADE and Gender
# ===============================
grade_map <- c(
  "1" = "F", "2" = "Sph", "3" = "J", "4" = "Sen", 
  "5" = "M1", "6" = "M2", "7" = "M3", "8" = "Doc", 
  "12" = "8th", "13" = "9th", "14" = "10th", "15" = "11th", "16" = "12th"
)
cells_data$grade <- factor(grade_map[as.character(cells_data$grade)])

# ADD THIS LINE TO FIX GENDER:
cells_data$gender <- factor(cells_data$gender)

# ===============================
# 3. DEFINE OUTCOME (TOTAL SPAI ADDICTION)
# ===============================
# Generate names for all 20 SPAI items
spai_all <- paste0("SPAI_", 1:20)

# Convert all 20 columns to numeric
cells_data[spai_all] <- lapply(cells_data[spai_all], function(x) as.numeric(as.character(x)))

# MODIFIED: Calculate the sum of ALL 20 SPAI variables
cells_data$sum_SPAI <- rowSums(cells_data[, spai_all], na.rm = TRUE)

# MODIFIED: Set sum_SPAI as the new dependent variable
y <- cells_data$sum_SPAI

# ===============================
# 4. PREDICTORS (NO SPAI LEAKAGE)
# ===============================
# MODIFIED: Drop the new target variable (sum_SPAI) and all individual SPAI columns 
x <- cells_data %>% dplyr::select(-sum_SPAI, -all_of(spai_all))

# Safety check: Ensure no lingering SPAI or target columns exist in your predictors
stopifnot(!any(grepl("SPAI", names(x))))

# ===============================
# 5. FIXED: MODEL MATRIX BEFORE SPLIT
# ===============================
x_full_mat <- model.matrix(~ . - 1, data = x)

# ===============================
# 6. TRAIN / TEST SPLIT
# ===============================
set.seed(123)
n <- nrow(x_full_mat)
train_index <- sample(1:n, size = 0.8 * n)

x_train_mat <- x_full_mat[train_index, ]
x_test_mat  <- x_full_mat[-train_index, ]
y_train     <- y[train_index]
y_test      <- y[-train_index]

# ===============================
# 7. LASSO (TRAIN ONLY)
# ===============================
set.seed(123)
cv_lasso <- cv.glmnet(
  x = x_train_mat, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# ===============================
# 8. TRAIN VS TEST EVALUATION (UPDATED WITH NMAE & MAPE)
# ===============================
pred_train <- as.numeric(predict(cv_lasso, s = "lambda.min", newx = x_train_mat))
pred_test  <- as.numeric(predict(cv_lasso, s = "lambda.min", newx = x_test_mat))

rmse_train <- sqrt(mean((y_train - pred_train)^2))
mae_train  <- mean(abs(y_train - pred_train))
r2_train   <- 1 - (sum((y_train - pred_train)^2) / sum((y_train - mean(y_train))^2))

rmse_test  <- sqrt(mean((y_test - pred_test)^2))
mae_test   <- mean(abs(y_test - pred_test))
r2_test    <- 1 - (sum((y_test - pred_test)^2) / sum((y_test - mean(y_test))^2))

# FIXED: Removed the duplicate uncorrected line beneath this one
cv_rmse    <- sqrt(cv_lasso$cvm[cv_lasso$lambda == cv_lasso$lambda.min])

# ADDED: Calculation for Scale-Agnostic Percentage Metrics 
range_y_test   <- max(y_test) - min(y_test)
norm_mae_test  <- (mae_test / range_y_test) * 100
mape_test      <- mean(abs((y_test - pred_test) / y_test)) * 100


# ==============================================================================
# ADDITIONAL CODE: PLOT FIT AND EXTRACT TOP PREDICTORS (TOTAL SPAI MODEL)
# ==============================================================================

# ------------------------------------------------------------------------------
# 1. VISUALIZE MODEL FIT (ACTUAL VS. PREDICTED)
# ------------------------------------------------------------------------------
plot_data <- data.frame(
  Actual = y_test,
  Predicted = pred_test
)
# Determine coordinate boundaries for a perfectly square 1:1 plot
min_val <- min(c(plot_data$Actual, plot_data$Predicted), na.rm = TRUE)
max_val <- max(c(plot_data$Actual, plot_data$Predicted), na.rm = TRUE)

plot(
  x = plot_data$Actual, 
  y = plot_data$Predicted,
  xlab = "Actual Total SPAI Score",
  ylab = "Predicted Total SPAI Score",
  pch = 16,
  col = rgb(0.2, 0.4, 0.6, 0.5),
  xlim = c(min_val, max_val),
  ylim = c(min_val, max_val)
)

abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

legend(
  "topleft",
  legend = "Perfect Prediction Line",
  col = "red",
  lty = 2,
  lwd = 2
)

```

<center>

## Lasso Model Comparison Table {.smaller}

</center>


```{r}
# ------------------------------------------------------------------------------
# 2. EXTRACT AND CLEAN TOP PREDICTORS
# ------------------------------------------------------------------------------
# Extract Lasso coefficients at the optimal lambda (lambda.min)
coef_lasso <- coef(cv_lasso, s = "lambda.min")

# Convert to a clean matrix format
coef_matrix <- as.matrix(coef_lasso)
coef_df <- data.frame(
  Dummy_Variable = rownames(coef_matrix),
  Coefficient = coef_matrix[, 1],
  Absolute_Impact = abs(coef_matrix[, 1]),
  stringsAsFactors = FALSE
)

# Remove the Intercept and filter to only selected features (non-zero)
selected_features <- coef_df %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

# Combine everything into a clean presentation table with exact string matching
first_top_predictors <- selected_features %>%
  rowwise() %>%
  mutate(Original_Variable = {
    # Find which original column names match the start of the dummy name
    matches <- names(x)[sapply(names(x), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    # Pick the longest match to avoid sub-string confusion
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)


# MODIFIED: Saved with a unique filename specific to this new total SPAI outcome
write.csv(first_top_predictors, "total_spai_first_predictors.csv", row.names = FALSE)




# ===============================
# 9. OUTPUT RESULTS TABLE (UPDATED HORIZONTAL LAYOUT)
# ===============================
# 1. Build the base vertical results data frame
results_vertical <- data.frame(
  Metric = c("RMSE", "MAE", "R-Squared"),
  Train  = c(round(rmse_train, 4), round(mae_train, 4), round(r2_train, 4)),
  Test   = c(round(rmse_test, 4), round(mae_test, 4), round(r2_test, 4))
)

# 2. Transpose the metrics to make them columns and include scale-agnostic metrics
results_horizontal <- data.frame(
  Model = "First_Lasso_Reduction",
  OVR = length(unique(first_top_predictors$Original_Variable)),
  `Train RMSE` = results_vertical[1, "Train"],
  `Test RMSE`  = results_vertical[1, "Test"],
  `Train MAE`   = results_vertical[2, "Train"],
  `Test MAE`    = results_vertical[2, "Test"],
  `Train R-Squared` = results_vertical[3, "Train"],
  `Test R-Squared`  = results_vertical[3, "Test"],
  `Test Norm MAE (%)` = round(norm_mae_test, 2),
  `Test MAPE (%)`     = round(mape_test, 2),
  `Optimal Lambda`    = round(cv_lasso$lambda.min, 6), # ADDED: Lambda column
  check.names = FALSE
)


# ==============================================================================
# DISPLAY MODEL SIZE AND TOP 5 PREDICTOR VARIABLES
# ==============================================================================
# Calculate the count of unique original variables with non-zero coefficients
total_surviving_vars <- length(unique(first_top_predictors$Original_Variable))


####################################################################
###############################################################################
#
#       Second Lasso  lambda.min again on 52 to modify coefficients and still has 52 after this run
#
###############################################################################
####################################################################


# ==============================================================================
# 1. EXTRACT SELECTED VARIABLE NAMES (REMOVING DUMMY FACTOR SUFFIXES)
# ==============================================================================
# Get unique names of the original variables that survived the first round
remaining_vars <- unique(first_top_predictors$Original_Variable)

# ==============================================================================
# 2. SUBSET THE ORIGINAL X PREDICTORS
# ==============================================================================
# Create a new predictor dataframe containing only the surviving variables
x_reduced <- cells_data %>% dplyr::select(all_of(remaining_vars))

# ==============================================================================
# 3. GENERATE NEW MODEL MATRIX & SPLIT DATA
# ==============================================================================
x_reduced_mat <- model.matrix(~ . - 1, data = x_reduced)

# Re-use the exact same training indexes from your first split for consistency
x_train_mat2 <- x_reduced_mat[train_index, ]
x_test_mat2  <- x_reduced_mat[-train_index, ]
# Note: y_train and y_test remain exactly the same as your sum_SPAI vectors

# ==============================================================================
# 4. RUN SECOND LASSO MODEL
# ==============================================================================
set.seed(123) # Reset seed for reproducibility in cross-validation folds
cv_lasso2 <- cv.glmnet(
  x = x_train_mat2, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# ==============================================================================
# 6. EXTRACT, CLEAN, AND SAVE SECOND REDUCTION VARIABLES
#    (MOVED UP: Must create 'second_top_predictors' before evaluating performance)
# ==============================================================================
coef_lasso2  <- coef(cv_lasso2, s = "lambda.min")
coef_matrix2 <- as.matrix(coef_lasso2)

coef_df2 <- data.frame(
  Dummy_Variable = rownames(coef_matrix2),
  Coefficient = coef_matrix2[, 1],
  Absolute_Impact = abs(coef_matrix2[, 1]),
  stringsAsFactors = FALSE
)

selected_features2 <- coef_df2 %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

second_top_predictors <- selected_features2 %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_reduced)[sapply(names(x_reduced), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)


# Saved with a unique filename specific to this second total SPAI reduction
write.csv(second_top_predictors, "total_spai_second_predictors.csv", row.names = FALSE)

# ==============================================================================
# 5. EVALUATE SECOND MODEL PERFORMANCE (UPDATED WITH NMAE, MAPE, & LAMBDA)
# ==============================================================================
pred_train2 <- as.numeric(predict(cv_lasso2, s = "lambda.min", newx = x_train_mat2))
pred_test2  <- as.numeric(predict(cv_lasso2, s = "lambda.min", newx = x_test_mat2))

# Compute baseline metrics for vertical extraction
mae_train2      <- mean(abs(y_train - pred_train2)) # ADDED FOR MASTER TABLE CONSISTENCY
mae_test2       <- mean(abs(y_test - pred_test2))
range_y_test    <- max(y_test) - min(y_test)
norm_mae_test2  <- (mae_test2 / range_y_test) * 100
mape_test2      <- mean(abs((y_test - pred_test2) / y_test)) * 100
lambda_used2    <- cv_lasso2$lambda.min

# 1. Build the base vertical results data frame
results2_vertical <- data.frame(
  Metric = c("RMSE", "MAE", "R-Squared"),
  Train  = c(round(sqrt(mean((y_train - pred_train2)^2)), 4), round(mae_train2, 4), round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4)),
  Test   = c(round(sqrt(mean((y_test - pred_test2)^2)), 4), round(mae_test2, 4), round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4))
)

# 2. Transpose the metrics to make them columns with the correct OVR mapping
results2_horizontal <- data.frame(
  Model = "Second_Lasso_Reduction",
  OVR = length(unique(second_top_predictors$Original_Variable)),
  `Train RMSE` = results2_vertical[1, "Train"],
  `Test RMSE`  = results2_vertical[1, "Test"],
  `Train MAE`   = results2_vertical[2, "Train"],
  `Test MAE`    = results2_vertical[2, "Test"],
  `Train R-Squared` = results2_vertical[3, "Train"],
  `Test R-Squared`  = results2_vertical[3, "Test"],
  `Test Norm MAE (%)` = round(norm_mae_test2, 2), 
  `Test MAPE (%)`     = round(mape_test2, 2),     
  `Optimal Lambda`    = round(lambda_used2, 6),   
  check.names = FALSE
)


#------------------------------------------------------------------
#   Comparison Tables Code
#------------------------------------------------------------------

# ==============================================================================
# 1. CALCULATE VARIABLE COUNTS FOR EACH STAGE
# ==============================================================================
initial_var_count <- ncol(x)
first_var_count   <- length(unique(first_top_predictors$Original_Variable))
second_var_count  <- length(unique(second_top_predictors$Original_Variable))

# ==============================================================================
# 2. RUN BASELINE MULTIPLE LINEAR REGRESSION (FOR COMPARISON PERFORMANCE)
# ==============================================================================
baseline_model <- lm(y_train ~ ., data = as.data.frame(x_train_mat))

pred_train_base <- predict(baseline_model, newdata = as.data.frame(x_train_mat))
pred_test_base  <- predict(baseline_model, newdata = as.data.frame(x_test_mat))

# Baseline metrics (and new context variables)
rmse_tr_b  <- sqrt(mean((y_train - pred_train_base)^2, na.rm=TRUE))
rmse_te_b  <- sqrt(mean((y_test - pred_test_base)^2, na.rm=TRUE))
mae_tr_b   <- mean(abs(y_train - pred_train_base), na.rm=TRUE) # ADDED
mae_te_b   <- mean(abs(y_test - pred_test_base), na.rm=TRUE)
r2_tr_b    <- 1 - (sum((y_train - pred_train_base)^2, na.rm=TRUE) / sum((y_train - mean(y_train))^2))
r2_te_b    <- 1 - (sum((y_test - pred_test_base)^2, na.rm=TRUE) / sum((y_test - mean(y_test))^2))
norm_mae_b <- (mae_te_b / range_y_test) * 100
mape_b     <- mean(abs((y_test - pred_test_base) / y_test), na.rm=TRUE) * 100

# ==============================================================================
# 3. CONSTRUCT THE MASTER COMPARISON TABLE (UPDATED FOR ALL METRICS)
# ==============================================================================
master_comparison <- data.frame(
  Metric = c(
    "OVR", 
    "Train RMSE", 
    "Test RMSE", 
    "Train MAE",         # ADDED
    "Test MAE",          # ADDED
    "Train R-Squared", 
    "Test R-Squared",
    "Test Norm MAE (%)", 
    "Test MAPE (%)",     
    "Optimal Lambda"     
  ),
  Baseline_Full_Model = c(
    initial_var_count, 
    round(rmse_tr_b, 4), 
    round(rmse_te_b, 4), 
    round(mae_tr_b, 4),  # ADDED
    round(mae_te_b, 4),  # ADDED
    round(r2_tr_b, 4), 
    round(r2_te_b, 4),
    round(norm_mae_b, 2), 
    round(mape_b, 2),     
    NA                    
  ),
  First_Lasso_Reduction = c(
    first_var_count, 
    round(rmse_train, 4), 
    round(rmse_test, 4), 
    round(mae_train, 4),  # ADDED (Assumed variables exist from step 1)
    round(mae_test, 4),   # ADDED
    round(r2_train, 4), 
    round(r2_test, 4),
    round(norm_mae_test, 2),
    round(mape_test, 2),
    round(cv_lasso$lambda.min, 6)
  ),
  Second_Lasso_Reduction = c(
    second_var_count, 
    round(sqrt(mean((y_train - pred_train2)^2)), 4), 
    round(sqrt(mean((y_test - pred_test2)^2)), 4), 
    round(mae_train2, 4), # ADDED
    round(mae_test2, 4),  # ADDED
    round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4), 
    round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4),
    round(norm_mae_test2, 2),
    round(mape_test2, 2),
    round(lambda_used2, 6)
  ),
  stringsAsFactors = FALSE
)

# ==============================================================================
# 4. TRANSPOSE, PRINT, AND SAVE THE SUMMARY TABLE
# ==============================================================================
# Clean transposition keeping matrix string issues isolated
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Strip default sequence index numbering from print layout
row.names(transposed_comparison) <- NULL


write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)




#______________________________________________________________________________________
#######################################################################################
#
#
#  New Lasso to actually do a second reduction  Modified lambda to lambda.1se Reduces from 52 to 33
#
#
#######################################################################################
#______________________________________________________________________________________




# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 3 MODELS BEFORE APPENDING
# ==============================================================================
# CORRECTED: Ensures the matrix object clears any downstream models cleanly
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:4]
}

# ==============================================================================
# 1. EXTRACT COEFFICIENTS AND COUNT VARIABLES FOR LAMBDA.1SE
# ==============================================================================
# Extract Lasso coefficients using the stricter 1-standard-error rule
coef_lasso_1se  <- coef(cv_lasso, s = "lambda.1se")
coef_matrix_1se <- as.matrix(coef_lasso_1se)

coef_df_1se <- data.frame(
  Dummy_Variable = rownames(coef_matrix_1se),
  Coefficient = coef_matrix_1se[, 1],
  Absolute_Impact = abs(coef_matrix_1se[, 1]),
  stringsAsFactors = FALSE
)

# Filter out the intercept and zeroed-out variables
selected_features_1se <- coef_df_1se %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0)

# Clean variable names to count unique original variables
top_predictors_1se <- selected_features_1se %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x)[sapply(names(x), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup()

se1_var_count <- length(unique(top_predictors_1se$Original_Variable))

# ==============================================================================
# 2. EVALUATE LAMBDA.1SE MODEL PERFORMANCE (FIXED: COMPUTING ALL REQUIRED METRICS)
# ==============================================================================
# Generate predictions using the 1se penalty strength
pred_train_1se <- as.numeric(predict(cv_lasso, s = "lambda.1se", newx = x_train_mat))
pred_test_1se  <- as.numeric(predict(cv_lasso, s = "lambda.1se", newx = x_test_mat))

# Compute metrics
rmse_tr_1se   <- sqrt(mean((y_train - pred_train_1se)^2))
rmse_te_1se   <- sqrt(mean((y_test - pred_test_1se)^2))
mae_tr_1se    <- mean(abs(y_train - pred_train_1se))
mae_te_1se    <- mean(abs(y_test - pred_test_1se))
r2_tr_1se     <- 1 - (sum((y_train - pred_train_1se)^2) / sum((y_train - mean(y_train))^2))
r2_te_1se     <- 1 - (sum((y_test - pred_test_1se)^2) / sum((y_test - mean(y_test))^2))

# Compute new horizontal metrics for the 1se model
range_y_test      <- max(y_test) - min(y_test)
norm_mae_test_1se <- (mae_te_1se / range_y_test) * 100
mape_test_1se     <- mean(abs((y_test - pred_test_1se) / y_test)) * 100

# ==============================================================================
# 3. CONSTRUCT THE MASTER COMPARISON TABLE (FIXED VARIABLES SYNCHRONIZATION)
# ==============================================================================
master_comparison <- data.frame(
  Metric = c(
    "OVR", 
    "Train RMSE", 
    "Test RMSE", 
    "Train MAE",         
    "Test MAE",          
    "Train R-Squared", 
    "Test R-Squared",
    "Test Norm MAE (%)", 
    "Test MAPE (%)",     
    "Optimal Lambda"     
  ),
  Baseline_Full_Model = c(
    initial_var_count, 
    round(rmse_tr_b, 4), 
    round(rmse_te_b, 4), 
    round(mae_tr_b, 4),  
    round(mae_te_b, 4),  
    round(r2_tr_b, 4), 
    round(r2_te_b, 4),
    round(norm_mae_b, 2), 
    round(mape_b, 2),     
    NA                    
  ),
  First_Lasso_Reduction = c(
    first_var_count, 
    round(rmse_train, 4), 
    round(rmse_test, 4), 
    round(mae_train, 4),  
    round(mae_test, 4),   
    round(r2_train, 4), 
    round(r2_test, 4),
    round(norm_mae_test, 2),
    round(mape_test, 2),
    round(cv_lasso$lambda.min, 6)
  ),
  Second_Lasso_Reduction = c(
    second_var_count, 
    round(sqrt(mean((y_train - pred_train2)^2)), 4), 
    round(sqrt(mean((y_test - pred_test2)^2)), 4), 
    round(mae_train2, 4), 
    round(mae_test2, 4),  
    round(1 - (sum((y_train - pred_train2)^2) / sum((y_train - mean(y_train))^2)), 4), 
    round(1 - (sum((y_test - pred_test2)^2) / sum((y_test - mean(y_test))^2)), 4),
    round(norm_mae_test2, 2),
    round(mape_test2, 2),
    round(lambda_used2, 6)
  ),
  Lambda_1se_Parsimonious = c(
    se1_var_count,                                                  # OVR
    round(rmse_tr_1se, 4),                                          # Train RMSE
    round(rmse_te_1se, 4),                                          # Test RMSE
    round(mae_tr_1se, 4),                                           # Train MAE
    round(mae_te_1se, 4),                                           # Test MAE
    round(r2_tr_1se, 4),                                            # Train R-Squared
    round(r2_te_1se, 4),                                            # Test R-Squared
    round(norm_mae_test_1se, 2),                                    # Test Norm MAE (%)
    round(mape_test_1se, 2),                                        # Test MAPE (%)
    round(cv_lasso$lambda.1se, 6)                                   # Optimal Lambda (1se)
  ),
  stringsAsFactors = FALSE
)

# ==============================================================================
# 4. TRANSPOSE, PRINT, AND SAVE THE SUMMARY TABLE
# ==============================================================================
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Ensure numeric columns print correctly and don't convert back to raw characters
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL



write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)


#-----------------------------------------------------------------------------------------------------

#####################################################################################################
######################################################################################################
#
#      Third Lasso Reduction  Modified Lambda = 1.75Lambda.1se  Reduces from 33 to 28 
#
#####################################################################################################
#####################################################################################################



# ==============================================================================
# 0. ENVIRONMENT RESET: TRUNCATE TABLE TO THE INITIAL 4 MODELS BEFORE APPENDING
# ==============================================================================
# Clears out any lingering downstream models from your active R workspace session
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:5]
}

# ==============================================================================
# 1. EXTRACT THE SURVIVING VARIABLES FROM THE 1SE MODEL
# ==============================================================================
# Dynamically reads the variable count from your 1se total SPAI model
vars_1se <- unique(top_predictors_1se$Original_Variable)
x_strict_df <- cells_data %>% dplyr::select(all_of(vars_1se))
x_strict_mat <- model.matrix(~ . - 1, data = x_strict_df)

# Align with your original training and testing splits
x_train_strict <- x_strict_mat[train_index, ]
x_test_strict  <- x_strict_mat[-train_index, ]

# ==============================================================================
# 2. RUN A HIGH-PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_strict <- cv.glmnet(
  x = x_train_strict, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# Manually amplify the 1se lambda by 1.75x to force aggressive variable dropping
strict_lambda <- cv_strict$lambda.1se * 1.75

# ==============================================================================
# 3. EXTRACT AND CLEAN THE ULTRA-REDUCED PREDICTORS
# ==============================================================================
coef_strict  <- coef(cv_strict, s = strict_lambda)
matrix_strict <- as.matrix(coef_strict)

df_strict <- data.frame(
  Dummy_Variable = rownames(matrix_strict),
  Coefficient = matrix_strict[, 1],
  Absolute_Impact = abs(matrix_strict[, 1]),
  stringsAsFactors = FALSE
)

selected_strict <- df_strict %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

strict_top_predictors <- selected_strict %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_strict_df)[sapply(names(x_strict_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

strict_var_count <- length(unique(strict_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE STRICT MODEL PERFORMANCE (UPDATED FOR NEW METRICS)
# ==============================================================================
pred_train_st <- as.numeric(predict(cv_strict, s = strict_lambda, newx = x_train_strict))
pred_test_st  <- as.numeric(predict(cv_strict, s = strict_lambda, newx = x_test_strict))

# Core Error Metrics
rmse_tr_st <- sqrt(mean((y_train - pred_train_st)^2))
rmse_te_st <- sqrt(mean((y_test - pred_test_st)^2))
mae_tr_st  <- mean(abs(y_train - pred_train_st))
mae_te_st  <- mean(abs(y_test - pred_test_st))
r2_tr_st   <- 1 - (sum((y_train - pred_train_st)^2) / sum((y_train - mean(y_train))^2))
r2_te_st   <- 1 - (sum((y_test - pred_test_st)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics (ADDED)
range_y_test     <- max(y_test) - min(y_test)
norm_mae_test_st <- (mae_te_st / range_y_test) * 100
mape_test_st     <- mean(abs((y_test - pred_test_st) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON
# ==============================================================================
# Append all 10 metrics cleanly to match master dataframe dimensions
master_comparison$Strict_Hyper_Parsimonious <- c(
  strict_var_count,             # OVR
  round(rmse_tr_st, 4),         # Train RMSE
  round(rmse_te_st, 4),         # Test RMSE
  round(mae_tr_st, 4),          # Train MAE
  round(mae_te_st, 4),          # Test MAE
  round(r2_tr_st, 4),           # Train R-Squared
  round(r2_te_st, 4),           # Test R-Squared
  round(norm_mae_test_st, 2),   # Test Norm MAE (%)
  round(mape_test_st, 2),       # Test MAPE (%)
  round(strict_lambda, 6)       # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Keep formatting clean across numeric properties
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the updated transposed spreadsheet version to your working files
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(strict_top_predictors, "total_spai_strict_predictors.csv", row.names = FALSE)



######################################################################
#
#
#   I am going to do a fourth reduction  from 28 to 22 Variables.
#
#
######################################################################
#

# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 5 MODELS BEFORE APPENDING
# ==============================================================================
# Clears out any lingering downstream terminal models from your active R session
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:6]
}

# ==============================================================================
# 1. EXTRACT THE 28 VARIABLES FROM THE PREVIOUS STRICT MODEL
# ==============================================================================
vars_strict_28 <- unique(strict_top_predictors$Original_Variable)
x_ultra_df     <- cells_data %>% dplyr::select(all_of(vars_strict_28))
x_ultra_mat    <- model.matrix(~ . - 1, data = x_ultra_df)

x_train_ultra <- x_ultra_mat[train_index, ]
x_test_ultra  <- x_ultra_mat[-train_index, ]

# ==============================================================================
# 2. RUN AN ULTRA-HIGH PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_ultra <- cv.glmnet(
  x = x_train_ultra, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

ultra_lambda <- cv_ultra$lambda.1se * 2.50

# ==============================================================================
# 3. EXTRACT AND CLEAN THE NEW ULTRA-REDUCED PREDICTORS
# ==============================================================================
coef_ultra   <- coef(cv_ultra, s = ultra_lambda)
matrix_ultra <- as.matrix(coef_ultra)

df_ultra <- data.frame(
  Dummy_Variable = rownames(matrix_ultra),
  Coefficient = matrix_ultra[, 1],
  Absolute_Impact = abs(matrix_ultra[, 1]),
  stringsAsFactors = FALSE
)

selected_ultra <- df_ultra %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

ultra_top_predictors <- selected_ultra %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_ultra_df)[sapply(names(x_ultra_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

ultra_var_count <- length(unique(ultra_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE ULTRA-STRICT MODEL PERFORMANCE (UPDATED FOR ALL METRICS)
# ==============================================================================
pred_train_ultra <- as.numeric(predict(cv_ultra, s = ultra_lambda, newx = x_train_ultra))
pred_test_ultra  <- as.numeric(predict(cv_ultra, s = ultra_lambda, newx = x_test_ultra))

# Core Error Metrics
rmse_tr_ult <- sqrt(mean((y_train - pred_train_ultra)^2))
rmse_te_ult <- sqrt(mean((y_test - pred_test_ultra)^2))
mae_tr_ult  <- mean(abs(y_train - pred_train_ultra))
mae_te_ult  <- mean(abs(y_test - pred_test_ultra))
r2_tr_ult   <- 1 - (sum((y_train - pred_train_ultra)^2) / sum((y_train - mean(y_train))^2))
r2_te_ult   <- 1 - (sum((y_test - pred_test_ultra)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics (ADDED)
range_y_test       <- max(y_test) - min(y_test)
norm_mae_test_ult  <- (mae_te_ult / range_y_test) * 100
mape_test_ult      <- mean(abs((y_test - pred_test_ultra) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON (FIXED SYNCHRONIZATION)
# ==============================================================================
# Append all 10 metrics to seamlessly align with master dataframe dimensions
master_comparison$Ultra_Strict_Reduction <- c(
  ultra_var_count,              # OVR
  round(rmse_tr_ult, 4),         # Train RMSE
  round(rmse_te_ult, 4),         # Test RMSE
  round(mae_tr_ult, 4),          # Train MAE
  round(mae_te_st, 4),           # Test MAE
  round(r2_tr_ult, 4),           # Train R-Squared
  round(r2_te_ult, 4),           # Test R-Squared
  round(norm_mae_test_ult, 2),   # Test Norm MAE (%)
  round(mape_test_ult, 2),       # Test MAPE (%)
  round(ultra_lambda, 6)         # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Convert character matrices back to numeric for perfect console alignment
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the updated transposed spreadsheet version to your working files
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(ultra_top_predictors, "total_spai_ultra_strict_predictors.csv", row.names = FALSE)



#################----------------------#######################------------------##############
#
#       Fifth Lasso Reduced from 22 to 19 Variables
#
#################----------------------#######################------------------##############

# ==============================================================================
# 0. MEMORY RESET: TRUNCATE TABLE TO THE INITIAL 6 MODELS BEFORE APPENDING
# ==============================================================================
# CORRECTED: Safely retains your 6 foundational columns plus Ultra_Strict (7 columns total)
if(exists("master_comparison")) {
  master_comparison <- master_comparison[, 1:7]
}

# ==============================================================================
# 1. EXTRACT THE 22 VARIABLES FROM THE ULTRA-STRICT MODEL
# ==============================================================================
# Extract the unique names of variables that survived the ultra-strict pass
vars_ultra_22 <- unique(ultra_top_predictors$Original_Variable)
x_term_df     <- cells_data %>% dplyr::select(all_of(vars_ultra_22))
x_term_mat    <- model.matrix(~ . - 1, data = x_term_df)

# Align with your original training and testing splits for consistency
x_train_term <- x_term_mat[train_index, ]
x_test_term  <- x_term_mat[-train_index, ]

# ==============================================================================
# 2. RUN A TERMINAL MAXIMUM-PENALTY LASSO CORRECTION
# ==============================================================================
set.seed(123)
cv_term <- cv.glmnet(
  x = x_train_term, y = y_train, 
  alpha = 1, family = "gaussian", nfolds = 10
)

# Amplify the penalty multiplier to 3.5x to aggressively drop to the absolute core
term_lambda <- cv_term$lambda.1se * 3.50

# ==============================================================================
# 3. EXTRACT AND CLEAN THE TERMINAL PREDICTORS
# ==============================================================================
coef_term   <- coef(cv_term, s = term_lambda)
matrix_term <- as.matrix(coef_term)

df_term <- data.frame(
  Dummy_Variable = rownames(matrix_term),
  Coefficient = matrix_term[, 1],
  Absolute_Impact = abs(matrix_term[, 1]),
  stringsAsFactors = FALSE
)

selected_term <- df_term %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(Absolute_Impact))

term_top_predictors <- selected_term %>%
  rowwise() %>%
  mutate(Original_Variable = {
    matches <- names(x_term_df)[sapply(names(x_term_df), function(orig) grepl(paste0("^", orig), Dummy_Variable))]
    if(length(matches) > 0) matches[which.max(nchar(matches))] else Dummy_Variable
  }) %>%
  ungroup() %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient)

term_var_count <- length(unique(term_top_predictors$Original_Variable))

# ==============================================================================
# 4. EVALUATE TERMINAL MODEL PERFORMANCE (UPDATED FOR ALL METRICS)
# ==============================================================================
pred_train_term <- as.numeric(predict(cv_term, s = term_lambda, newx = x_train_term))
pred_test_term  <- as.numeric(predict(cv_term, s = term_lambda, newx = x_test_term))

# Core Error Metrics
rmse_tr_term <- sqrt(mean((y_train - pred_train_term)^2))
rmse_te_term <- sqrt(mean((y_test - pred_test_term)^2))
mae_tr_term  <- mean(abs(y_train - pred_train_term))
mae_te_term  <- mean(abs(y_test - pred_test_term))
r2_tr_term   <- 1 - (sum((y_train - pred_train_term)^2) / sum((y_train - mean(y_train))^2))
r2_te_term   <- 1 - (sum((y_test - pred_test_term)^2) / sum((y_test - mean(y_test))^2))

# Horizontal Context Metrics
range_y_test      <- max(y_test) - min(y_test)
norm_mae_test_term <- (mae_te_term / range_y_test) * 100
mape_test_term     <- mean(abs((y_test - pred_test_term) / y_test)) * 100

# ==============================================================================
# 5. APPEND, TRANSPOSE, AND PRINT HORIZONTAL COMPARISON
# ==============================================================================
# Append all 10 metrics cleanly to match master dataframe dimensions exactly
master_comparison$Terminal_Core_Reduction <- c(
  term_var_count,               # OVR
  round(rmse_tr_term, 4),        # Train RMSE
  round(rmse_te_term, 4),        # Test RMSE
  round(mae_tr_term, 4),         # Train MAE
  round(mae_te_term, 4),         # Test MAE
  round(r2_tr_term, 4),          # Train R-Squared
  round(r2_te_term, 4),          # Test R-Squared
  round(norm_mae_test_term, 2),  # Test Norm MAE (%)
  round(mape_test_term, 2),      # Test MAPE (%)
  round(term_lambda, 6)          # Optimal Lambda
)

# Run transposition logic immediately within the execution sequence
transposed_comparison <- as.data.frame(t(master_comparison[, -1]), check.names = FALSE)
colnames(transposed_comparison) <- master_comparison$Metric

transposed_comparison <- data.frame(
  Model = rownames(transposed_comparison),
  transposed_comparison,
  check.names = FALSE
)

# Convert character matrices back to numeric for perfect console alignment
transposed_comparison[, 2:ncol(transposed_comparison)] <- lapply(
  transposed_comparison[, 2:ncol(transposed_comparison)], 
  function(x) as.numeric(as.character(x))
)

row.names(transposed_comparison) <- NULL


# Save the final clean wide format summary spreadsheet and the core predictors
write.csv(transposed_comparison, "total_spai_master_comparison_transposed.csv", row.names = FALSE)
write.csv(term_top_predictors, "total_spai_terminal_core_predictors.csv", row.names = FALSE)
###################################
#
#   Printing only the test Metrics
#
##################################

# ==============================================================================
# KEEP ONLY TEST OUTPUT METRICS + LAMBDA
# ==============================================================================

test_comparison <- transposed_comparison %>%
  dplyr::select(
    Model,
    OVR,
    `Test RMSE`,
    `Test MAE`,
    `Test R-Squared`,
    `Test Norm MAE (%)`,
    `Test MAPE (%)`,
    `Optimal Lambda`
  )
###################################
#
#   Printing only the test Metrics
#   with shortened model names
#   (removed Norm MAE and MAPE)
#
##################################

# ==============================================================================
# KEEP ONLY TEST OUTPUT METRICS + LAMBDA
# ==============================================================================

test_comparison <- transposed_comparison %>%
  dplyr::select(
    Model,
    OVR,
    `Test RMSE`,
    `Test MAE`,
    `Test R-Squared`,
    `Optimal Lambda`
  ) %>%
  dplyr::mutate(
    Model = dplyr::recode(
      Model,
      "Baseline_Full_Model" = "Baseline",
      "First_Lasso_Reduction" = "LASSO 1",
      "Second_Lasso_Reduction" = "LASSO 2",
      "Lambda_1se_Parsimonious" = "Lambda 1SE",
      "Strict_Hyper_Parsimonious" = "Strict",
      "Ultra_Strict_Reduction" = "Ultra",
      "Terminal_Core_Reduction" = "Terminal"
    )
  )

# ==============================================================================
# PRINT FINAL SLIDE TABLE
# ==============================================================================

###################################
#
#   Print centered slide table
#
##################################
# install.packages("knitr")
# install.packages("kableExtra")
library(knitr)
library(kableExtra)

test_comparison %>%
  kable(
    format = "html",
    digits = 3
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = FALSE,
    position = "center",
    font_size = 24
  )
```

## Final Lasso Regression Model {.smaller}

```{r}
# Load ggplot2 for a high-quality visualization
library(ggplot2)

# ==============================================================================
# 1. ISOLATE AND RANK THE WINNING 33 PREDICTORS
# ==============================================================================
# MODIFIED: Pulls from top_predictors_1se to capture the 33-variable model
final_ranked_predictors <- top_predictors_1se %>%
  filter(Dummy_Variable != "(Intercept)", Coefficient != 0) %>%
  # Recalculate absolute impact to ensure proper ranking execution
  mutate(Absolute_Impact = abs(Coefficient)) %>%
  arrange(desc(Absolute_Impact)) %>%
  dplyr::select(Original_Variable, Dummy_Variable, Coefficient, Absolute_Impact)

# # Print the full ranked list to your console
# cat("\n==================================================================\n")
# cat("          FINAL RANKED PREDICTORS (LAMBDA 1SE MODEL)               \n")
# cat("==================================================================\n")
# print(final_ranked_predictors, n = Inf, row.names = FALSE)

# MODIFIED: Saved with a unique filename specific to this final 33-variable set
write.csv(final_ranked_predictors, "total_spai_final_33_predictors.csv", row.names = FALSE)

# ==============================================================================
# 2. GENERATE A COEFFICIENT IMPACT VISUALIZATION
# ==============================================================================
# Create a clear color flag for positive vs. negative coefficients
plot_df <- final_ranked_predictors %>%
  mutate(
    Direction = ifelse(Coefficient > 0, "Increases Addiction", "Decreases Addiction"),
    # Reorder the dummy variables so the chart ranks them from largest to smallest impact
    Dummy_Variable = reorder(Dummy_Variable, Absolute_Impact)
  )

# Build the publication-ready bar chart with centered titles
coef_plot <- ggplot(plot_df, aes(x = Dummy_Variable, y = Coefficient, fill = Direction)) +
  geom_bar(stat = "identity", width = 0.75, color = "black", linewidth = 0.2) +
  coord_flip() + 
  scale_fill_manual(values = c("Increases Addiction" = "#d95f02", "Decreases Addiction" = "#7570b3")) +
  labs(
    title = "Predictor Impact on Total SPAI Score",
    subtitle = "Final Parsimonious Lasso Model (33 Variables Remaining)",
    x = "Model Predictors (Dummy Variables)",
    y = "Lasso Coefficient Value (Impact Weight)",
    fill = "Effect Direction"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    # FIXED: Added hjust = 0.5 to perfectly center the title and subtitle
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5, margin = margin(b = 5)),
    plot.subtitle = element_text(color = "gray30", size = 10, hjust = 0.5, margin = margin(b = 15)),
    axis.text.y = element_text(size = 9, family = "mono"), 
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank() 
  )


# Display the plot in your RStudio viewer
print(coef_plot)
```

##  Optimal Number of Predictor Variables {.smaller}

```{r}
# MODIFIED: Saved with a clear, specific filename for your final chosen 33-variable model
png("total_spai_33var_cross_validation_diagnostic.png", width = 7, height = 5, units = "in", res = 300)

# Set clean margins for the plot
par(mar = c(4.5, 4.5, 3, 1))

# Generate the standard glmnet cross-validation plot using your current cv_lasso object
plot(
  cv_lasso, 
  xlab = "Lasso Penalty Strength - Log(Lambda)", 
  ylab = "10-Fold Cross-Validation Error (MSE)"
)

# MODIFIED: Title updated to explicitly highlight the final 33-variable threshold selection
title("Total SPAI Model: Lambda Selection (Final 33-Variable Model Chosen)", line = 2, font.main = 1, cex.main = 1.1)

# Close the file device and save the image
invisible(dev.off())

# Display the plot directly in your RStudio viewer as well
plot(cv_lasso)

```

## Final Model Selection for Real Life Application {.smaller}
```{r}
# ==============================================================================
# 1. ISOLATE TARGET AND SELECTED PREDICTORS (TOP 5 OF 22 CONFIGURATION)
# ==============================================================================
# Define the precise 5 features for the updated configuration
selected_vars_top5_22 <- c("SAPS_10", "SAPS_7", "MPATS4", "SAPS_9", "SAS_CA14")

# Extract features from your base dataframe
x_ols_df_22 <- cells_data %>% dplyr::select(all_of(selected_vars_top5_22))

# Split the isolated feature space into matching Train/Test matrices 
# using your script's active training split index
x_train_ols_22 <- x_ols_df_22[train_index, ]
x_test_ols_22  <- x_ols_df_22[-train_index, ]

# Combine with target arrays into clean training data frames for lm()
train_data_ols_22 <- data.frame(y = y_train, x_train_ols_22)
test_data_ols_22  <- data.frame(y = y_test, x_test_ols_22)

# ==============================================================================
# 2. FIT THE MULTIPLE LINEAR REGRESSION (OLS) MODEL
# ==============================================================================
ols_five_model_22 <- lm(y ~ ., data = train_data_ols_22)

# # ==============================================================================
# 3. EVALUATE MODEL PERFORMANCE METRICS
# ==============================================================================
# Generate numeric vector predictions for test and training pools
pred_train_ols5_22 <- as.numeric(predict(ols_five_model_22, newdata = train_data_ols_22))
pred_test_ols5_22  <- as.numeric(predict(ols_five_model_22, newdata = test_data_ols_22))

# Calculate Core Error metrics
rmse_tr_ols5_22 <- sqrt(mean((y_train - pred_train_ols5_22)^2))
rmse_te_ols5_22 <- sqrt(mean((y_test - pred_test_ols5_22)^2))
mae_tr_ols5_22  <- mean(abs(y_train - pred_train_ols5_22))
mae_te_ols5_22  <- mean(abs(y_test - pred_test_ols5_22))
r2_tr_ols5_22   <- 1 - (sum((y_train - pred_train_ols5_22)^2) / sum((y_train - mean(y_train))^2))
r2_te_ols5_22   <- 1 - (sum((y_test - pred_test_ols5_22)^2) / sum((y_test - mean(y_test))^2))

# Calculate Context Metrics
range_y_test              <- max(y_test) - min(y_test)
norm_mae_test_ols5_22     <- (mae_te_ols5_22 / range_y_test) * 100
mape_test_ols5_22         <- mean(abs((y_test - pred_test_ols5_22) / y_test)) * 100

# ==============================================================================
# 4. PRINT FORMATTED PERFORMANCE RESULTS (EXCLUDING OPTIMAL LAMBDA)
# ==============================================================================
ols5_22_results_horizontal <- data.frame(
  Model = "Top_5_of_22_Model", # MODIFIED
  OVR = length(selected_vars_top5_22),
  `Train RMSE` = round(rmse_tr_ols5_22, 4),
  `Test RMSE`  = round(rmse_te_ols5_22, 4),
  `Train MAE`   = round(mae_tr_ols5_22, 4),
  `Test MAE`    = round(mae_te_ols5_22, 4),
  `Train R-Squared` = round(r2_tr_ols5_22, 4),
  `Test R-Squared`  = round(r2_te_ols5_22, 4),
  `Test Norm MAE (%)` = round(norm_mae_test_ols5_22, 2), 
  `Test MAPE (%)`     = round(mape_test_ols5_22, 2),     
  check.names = FALSE
)

# ==============================================================================
# 1. ISOLATE TARGET AND SELECTED PREDICTORS (MPATS4 EXCLUDED, SAS_C7 ADDED)
# ==============================================================================
# Define the precise 5 features for the updated configuration
selected_vars_new5 <- c("SAPS_10", "SAPS_7", "SAPS_9", "SAS_CA14", "SAS_C7")

# Extract features from your base dataframe
x_ols5_new_df <- cells_data %>% dplyr::select(all_of(selected_vars_new5))

# Split the isolated feature space into matching Train/Test matrices 
# using your script's active training split index
x_train_ols5_new <- x_ols5_new_df[train_index, ]
x_test_ols5_new  <- x_ols5_new_df[-train_index, ]

# Combine with target arrays into clean training data frames for lm()
train_data_ols5_new <- data.frame(y = y_train, x_train_ols5_new)
test_data_ols5_new  <- data.frame(y = y_test, x_test_ols5_new)

# ==============================================================================
# 2. FIT THE MULTIPLE LINEAR REGRESSION (OLS) MODEL
# ==============================================================================
ols_five_new_model <- lm(y ~ ., data = train_data_ols5_new)


# ==============================================================================
# 3. EVALUATE MODEL PERFORMANCE METRICS
# ==============================================================================
# Generate numeric vector predictions for test and training pools
pred_train_ols5_new <- as.numeric(predict(ols_five_new_model, newdata = train_data_ols5_new))
pred_test_ols5_new  <- as.numeric(predict(ols_five_new_model, newdata = test_data_ols5_new))

# Calculate Core Error metrics
rmse_tr_ols5_new <- sqrt(mean((y_train - pred_train_ols5_new)^2))
rmse_te_ols5_new <- sqrt(mean((y_test - pred_test_ols5_new)^2))
mae_tr_ols5_new  <- mean(abs(y_train - pred_train_ols5_new))
mae_te_ols5_new  <- mean(abs(y_test - pred_test_ols5_new))
r2_tr_ols5_new   <- 1 - (sum((y_train - pred_train_ols5_new)^2) / sum((y_train - mean(y_train))^2))
r2_te_ols5_new   <- 1 - (sum((y_test - pred_test_ols5_new)^2) / sum((y_test - mean(y_test))^2))

# Calculate Context Metrics
range_y_test           <- max(y_test) - min(y_test)
norm_mae_test_ols5_new <- (mae_te_ols5_new / range_y_test) * 100
mape_test_ols5_new     <- mean(abs((y_test - pred_test_ols5_new) / y_test)) * 100

# ==============================================================================
# 4. PRINT FORMATTED PERFORMANCE RESULTS (EXCLUDING OPTIMAL LAMBDA FOR OLS)
# ==============================================================================
ols5_new_results_horizontal <- data.frame(
  Model = "Top_5_of_33_Model", # MODIFIED
  OVR = length(selected_vars_new5),
  `Train RMSE` = round(rmse_tr_ols5_new, 4),
  `Test RMSE`  = round(rmse_te_ols5_new, 4),
  `Train MAE`   = round(mae_tr_ols5_new, 4),
  `Test MAE`    = round(mae_te_ols5_new, 4),
  `Train R-Squared` = round(r2_tr_ols5_new, 4),
  `Test R-Squared`  = round(r2_te_ols5_new, 4),
  `Test Norm MAE (%)` = round(norm_mae_test_ols5_new, 2), 
  `Test MAPE (%)`     = round(mape_test_ols5_new, 2),     
  check.names = FALSE
)

#
# ==============================================================================
# FINAL MASTER SPECIFICATION CONSOLIDATION (SLIDE-FRIENDLY FORMAT)
# ==============================================================================

mega_dynamic_comparison <- data.frame(
  Model = c(
    "Baseline",
    "LASSO 1",
    "LASSO 2",
    "Lambda 1SE",
    "Strict",
    "Ultra",
    "Terminal",
    "Top 5 (33)",
    "Top 5 (52/22)"
  ),
  OVR = c(81, 52, 52, 33, 28, 22, 19, 5, 5),
  `Test RMSE` = c(5.2338, 5.1702, 5.2142, 5.1626, 5.2553, 5.5280, 5.8296, 6.0199, 6.2256),
  `Test MAE` = c(3.9987, 3.9714, 3.9902, 3.9910, 4.1075, 4.1075, 4.6312, 4.7114, 4.8361),
  `Test R-Squared` = c(0.7429, 0.7491, 0.7448, 0.7498, 0.7407, 0.7131, 0.6810, 0.6598, 0.6362),
  `Test Norm MAE (%)` = c(7.41, 7.35, 7.39, 7.39, 7.61, 8.06, 8.58, 8.72, 8.96),
  `Test MAPE (%)` = c(10.59, 10.58, 10.60, 10.71, 11.21, 12.14, 13.08, 12.65, 12.91),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

# ==============================================================================
# SHORTEN COLUMN NAMES FOR PRESENTATION
# ==============================================================================

colnames(mega_dynamic_comparison) <- c(
  "Model",
  "OVR",
  "RMSE",
  "MAE",
  "R²",
  "Norm MAE (%)",
  "MAPE (%)"
)

# ==============================================================================
# PRINT SLIDE-READY TABLE
# ==============================================================================

library(knitr)
library(kableExtra)

mega_dynamic_comparison %>%
  kable(
    format = "html",
    digits = 3
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width = FALSE,
    position = "center",
    font_size = 24
  )
```

## Questions to ask {.smaller}

1. Even when I think I should stop using my mobile phone, I will continue to use it and cannot stop.
2. It has become a habit of me to spend a lot of time on my mobile phone.
3. I feel very panic when I can’t use my mobile phone.
4. I don’t get enough sleep and feel tired from using mobile phone too much.
5. Playing mobile phone has affected my academic performance.  


## Conclusion {.smaller}

- Iterative LASSO on Chinese CAT bank [@gao2024]: predict SPAI total from non-SPAI items + demographics
- 81 → 52 (`lambda.min`) → **33** (`lambda.1se`): best balance of test error and complexity
- Test RMSE $= 5.16$, test $R^2 = 0.75$ (~59% fewer predictors than baseline)
- Further shrinkage (28 / 22 / 19) degraded accuracy ($R^2 < 0.70$ at 19) → stopping rule
- **SAPS** and **SAS** dominate; smaller MPATS role; **NMP** largely zeroed out
- Post-LASSO OLS on five items: test $R^2 = 0.66$ (below 0.75, strong for five items)
- Those items: compulsive use, habitual time spent, separation panic, sleep loss, academic interference
- High $R^2$ = **convergent validity**, not causation [@gao2024; @lin2014; @james2013]

## China & Generalizability {.smaller}

Sample collected in **China** among predominantly young adults (mean age $= 18.8$), more female than male [@gao2024]

- Multicollinearity ($r > 0.60$) motivates LASSO over OLS [@mcneish2015]
- Do not generalize without external validation
- 33-item LASSO—and, where brevity matters, the five-item OLS subset—can estimate SPAI-like severity pending prospective validation

## Future Work {.smaller}

- Robustness across train–test splits and random seeds
- Validate short-form item sets in independent samples
- Pursue group LASSO more systematically (scale groupings and penalties vs. iterative plain LASSO)
- Extend beyond the Chinese young-adult cohort to separate context-specific from universal predictors

## References {.smaller}
