---
title: "LASSO Regression on Cellphone Use Addiction"
subtitle: "IDC6940 Capstone — Problematic Mobile Phone Use"
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
course: Capstone Projects in Data Science
bibliography: references.bib
self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(glmnet)
library(grpreg)
library(caret)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_12_num", "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  if (is.list(raw_coef)) {
    vars <- unique(unlist(lapply(raw_coef, function(m) {
      mat <- as.matrix(m)
      rownames(mat)[mat[, 1] != 0]
    })))
  } else {
    mat <- as.matrix(raw_coef)
    vars <- rownames(mat)[mat[, 1] != 0]
  }
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(R2 = 1 - ss_res / ss_tot, RMSE = sqrt(mean((y_true - y_pred)^2)))
}

dat <- read.csv("data/DATA_Problematic Mobile Phone Use CAT.csv")
dat <- dat[complete.cases(dat), ]
dat$gender <- as.factor(dat$gender)
dat$grade <- as.factor(dat$grade)
dat$SPAI_total <- rowSums(dat[, paste0("SPAI_", 1:20)])
dat$SPAI_12 <- as.factor(dat$SPAI_12)

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

train_index <- createDataPartition(dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
scale_counts <- scale_counts %>%
  filter(Scale != "other") %>%
  arrange(desc(N_items))

corr_df <- dat %>%
  summarise(
    NMP   = cor(NMP_total,   SPAI_total),
    SAPS  = cor(SAPS_total,  SPAI_total),
    MPAS  = cor(MPAS_total,  SPAI_total),
    MPATS = cor(MPATS_total, SPAI_total),
    SAS   = cor(SAS_total,   SPAI_total)
  ) %>%
  pivot_longer(everything(), names_to = "Scale", values_to = "r") %>%
  mutate(Scale = reorder(Scale, r))
```

## The Problem

Smartphone use has shifted from convenience to dependency.

- Compulsive checking, withdrawal, daily-life interference
- Links to anxiety, sleep loss, and academic strain [@lin2014]
- Dozens of overlapping survey variables → multicollinearity [@james2013]

**Question:** Which scales best predict overall addiction severity?

## Our Data

Computer Adaptive Testing (CAT) survey [@gao2024] — **1,619** complete cases

| Abbrev. | Scale |
|---------|-------|
| NMP | Nomophobia |
| SAPS | Addiction proneness |
| SPAI | Addiction inventory (outcome) |
| MPAS | Mobile phone addiction |
| MPATS | Addiction tendency |
| SAS | Smartphone addiction scale |

Plus age, gender, and grade (mean age ≈ 18.8; 991 F / 628 M).

## Outcome & Approach

**Outcome:** SPAI total (sum of 20 SPAI items, 20–80; mean ≈ 41.8)

**Predictors:** all other scales + demographics — SPAI items excluded (they define the score)

**Method:** LASSO [@tibshirani1996] — $L_1$ penalty shrinks weak coefficients to zero [@james2013; @hesterberg2008]

## LASSO in One Equation

$$
\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}
$$

- $\lambda$ tuned by 10-fold CV (`lambda.min` vs `lambda.1se`)
- **Plain LASSO** — item-level selection (`glmnet`)
- **Group LASSO** — whole-scale selection (`grpreg`) [@yuan2006; @huang2024]

## Two Modeling Dials

| | Category (multinomial) | Number (Gaussian) |
|---|---|---|
| **Item-level** (Plain) | Predict SPAI_12 class | Predict SPAI total |
| **Scale-level** (Group) | Predict SPAI_12 class | Predict SPAI total |

12 models total: 8 classification + 4 Gaussian regression

## Exploratory Findings

- **NMP** highest mean endorsement (~69% of max) but **weakest** SPAI correlation
- **SAS** strongest predictor ($r \approx 0.81$), then SAPS / MPATS (~0.71)
- All six scales highly intercorrelated → convergent validity + multicollinearity
- **SPAI item 12** imbalanced (~6% in rarest category)

## SPAI Severity in the Sample

```{r spa-total-hist}
#| fig-width: 8
#| fig-height: 4.5

ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  labs(
    x = "SPAI total score",
    y = "Respondents",
    title = "Distribution of smartphone addiction severity (SPAI total)"
  ) +
  theme_minimal(base_size = 18)
```

## Correlations with SPAI Total

```{r corr-bar}
#| fig-width: 8
#| fig-height: 4.5

ggplot(corr_df, aes(x = Scale, y = r, fill = r)) +
  geom_col(width = 0.6, alpha = 0.9) +
  geom_text(aes(label = round(r, 2)), hjust = -0.15, size = 5, fontface = "bold") +
  scale_fill_gradient(low = "#9ecae1", high = "#08519c", guide = "none") +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(x = NULL, y = "Pearson r", title = "Non-SPAI scales vs SPAI total") +
  theme_minimal(base_size = 18)
```

## Live Results — Gaussian LASSO

Single 80/20 split · test set *n* = `r nrow(mm_reg$x) - length(train_index)`

```{r gauss-table}
kable(gauss_results, align = c("l", "r", "r", "r"))
```

Other scales explain about **three-quarters** of SPAI-total variation on held-out respondents.

## Selected Scales (Plain LASSO, lambda.1se)

```{r scale-bar}
#| fig-width: 8
#| fig-height: 4.5

ggplot(scale_counts, aes(x = reorder(Scale, N_items), y = N_items, fill = Scale)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Selected items", title = "Which instruments matter most?") +
  theme_minimal(base_size = 18)
```

**SAPS** and **SAS** dominate · **NMP** (nomophobia) contributes little

## Recommendation & Next Steps

**Recommended model:** Plain LASSO at `lambda.1se` — strong R² with far fewer predictors than group LASSO

- Scales overlap → convergent validity, not causation
- External validation on a new sample
- Robustness across seeds/splits
- Write up why SAPS/SAS track SPAI total

## References
