---
title: "LASSO Regression on Cellphone Use Addiction"
subtitle: "IDC6940 Capstone — Problematic Mobile Phone Use"
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
course: Capstone Projects in Data Science
bibliography: references.bib
self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(glmnet)
library(grpreg)
library(caret)
library(knitr)
library(patchwork)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

theme_set(
  theme_minimal(base_size = 16) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  mat <- as.matrix(raw_coef)
  vars <- rownames(mat)[mat[, 1] != 0]
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(
    R2 = 1 - ss_res / ss_tot,
    RMSE = sqrt(mean((y_true - y_pred)^2))
  )
}

dat <- read.csv("data/DATA_Problematic Mobile Phone Use CAT.csv")
dat <- dat[complete.cases(dat), ]

dat$NMP_total   <- rowSums(dat[, paste0("NMP_",   1:16)])
dat$SAPS_total  <- rowSums(dat[, paste0("SAPS_",  1:10)])
dat$SPAI_total  <- rowSums(dat[, paste0("SPAI_",  1:20)])
dat$MPAS_total  <- rowSums(dat[, paste0("MPAS_",  1:11)])
dat$MPATS_total <- rowSums(dat[, paste0("MPATS",  1:16)])
dat$SAS_total   <- rowSums(dat[, c(paste0("SAS_C", 1:11),
                                    paste0("SAS_CA", 1:14))])

dat$gender <- as.factor(dat$gender)
dat$grade <- as.factor(dat$grade)
dat$SPAI_12 <- as.factor(dat$SPAI_12)

gender_labels <- c("1" = "Male", "2" = "Female")
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)],
                         levels = unname(gender_labels))

train_index <- createDataPartition(dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
```

## The Problem

Smartphone use has shifted from convenience to dependency.

- Compulsive checking, withdrawal, daily-life interference
- Links to anxiety, sleep loss, and academic strain [@lin2014]
- Dozens of overlapping survey variables → multicollinearity [@james2013]

**Question:** Which scales best predict overall addiction severity?

## Our Data

Survey data collected in **China** using a Computer Adaptive Testing (CAT)
item bank [@gao2024] — **$n = 1{,}619$** complete cases

| Abbrev. | Scale |
|---------|-------|
| NMP | Nomophobia |
| SAPS | Smartphone addiction proneness |
| SPAI | Smartphone Addiction Inventory (outcome) |
| MPAS | Mobile Phone Addiction Scale |
| MPATS | Mobile Phone Addiction Tendency Scale |
| SAS | Smartphone Addiction Scale (SAS-C / SAS-CA) |

Plus age, gender, and grade (mean age $= 18.8$; 991 F / 628 M).

## Outcome & Approach

**Primary outcome:** SPAI total — sum of 20 SPAI items (20–80; mean $= 41.8$)

**Predictors:** NMP, SAPS, MPAS, MPATS, SAS items + demographics — SPAI items excluded (they define the score)

**Secondary outcome:** SPAI item 12 (perceived impact on study/work; 4-level Likert)

**Method:** LASSO [@tibshirani1996] — $L_1$ penalty shrinks weak coefficients to zero [@james2013; @hesterberg2008]

## LASSO Objective

$$
\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij}\right)^2
+ \lambda \sum_{j=1}^{p}|\beta_j|
\right\}
$$

- $\lambda$ tuned by 10-fold CV on an 80/20 split (`lambda.min` vs `lambda.1se`)
- **Plain LASSO** — item-level selection (`glmnet`) [@james2013]
- **Group LASSO** — whole-scale selection (`grpreg`) [@yuan2006; @huang2024]
- **Main result:** plain LASSO at `lambda.1se` (parsimony vs fit)

## SPAI Severity in the Sample

```{r fig-spai-dist}
#| fig-width: 9
#| fig-height: 4.5

ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(binwidth = 3, fill = "#4C72B0", color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(dat$SPAI_total), linetype = "dashed",
             color = "#C44E52", linewidth = 0.9) +
  annotate("text",
           x = mean(dat$SPAI_total) + 1.5, y = Inf, vjust = 1.5, hjust = 0,
           label = paste0("Mean = ", round(mean(dat$SPAI_total), 1)),
           color = "#C44E52", size = 4) +
  labs(
    title = "Distribution of SPAI Total Score",
    x = "SPAI Total (20–80)", y = "Count"
  )
```

## Strongest Correlation: SAS vs SPAI

SAS ($r = 0.81$) > SAPS / MPATS ($r = 0.71$) > MPAS ($r = 0.65$) > NMP ($r = 0.55$)

```{r fig-scatter-sas}
#| fig-width: 9
#| fig-height: 4.5

ggplot(dat, aes(x = SAS_total, y = SPAI_total)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#4C72B0") +
  geom_smooth(method = "lm", se = TRUE, color = "#C44E52",
              fill = "#f4a7a9", linewidth = 1.1) +
  annotate("text",
           x = max(dat$SAS_total) * 0.05,
           y = max(dat$SPAI_total) * 0.96, hjust = 0,
           label = paste0("r = ", round(cor(dat$SAS_total, dat$SPAI_total), 2)),
           size = 5, fontface = "bold", color = "#C44E52") +
  labs(
    title = "SAS Total vs SPAI Total",
    x = "SAS Total (Smartphone Addiction Scale)",
    y = "SPAI Total (Addiction Severity)"
  )
```

## Convergent Validity & Multicollinearity

All six scale totals move together (inter-scale $r$ often exceeds 0.60) — motivates LASSO over OLS.

```{r fig-corr-heatmap}
#| fig-width: 8
#| fig-height: 5

corr_cols <- c("NMP_total", "SAPS_total", "SPAI_total", "MPAS_total",
               "MPATS_total", "SAS_total")
corr_mat <- cor(dat[, corr_cols], use = "complete.obs")

corr_long <- as.data.frame(corr_mat) %>%
  mutate(Scale1 = rownames(.)) %>%
  pivot_longer(-Scale1, names_to = "Scale2", values_to = "r") %>%
  mutate(
    Scale1 = gsub("_total", "", Scale1),
    Scale2 = gsub("_total", "", Scale2)
  )

ggplot(corr_long, aes(x = Scale2, y = Scale1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.5, color = "black") +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r", title = "Scale Total Correlations")
```

## Patterns Before Modeling

- **SPAI item 12** highly imbalanced — rarest category $= 5.7\%$ of sample
- **NMP** highest mean endorsement ($= 68.8\%$ of max) but **weakest** SPAI correlation
- Nomophobia may be redundant once SAPS and SAS are in the model

```{r fig-demographics}
#| fig-width: 9
#| fig-height: 3.5

p_age <- ggplot(dat, aes(x = age)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(title = "Age", x = "Years", y = "Count")

p_gender <- dat %>% count(gender_lab) %>%
  ggplot(aes(x = gender_lab, y = n, fill = gender_lab)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Gender", x = NULL, y = "Count")

p_age + p_gender + plot_layout(widths = c(2, 1))
```

## Data Preprocessing

- $n = 1{,}619$ complete cases; no listwise missingness
- SPAI total = sum of 20 SPAI items; other SPAI items removed from predictors
- Model matrix from NMP, SAPS, MPAS, MPATS, SAS items + demographics
- 80/20 train–test split, stratified on SPAI item 12 (`caret`)

## Gaussian LASSO Results

Test set $n = `r nrow(mm_reg$x) - length(train_index)`$

```{r tbl-gaussian}
kable(gauss_results, align = c("l", "r", "r", "r"),
      col.names = c("Model", "R²", "RMSE", "Predictors selected"))
```

Plain LASSO at `lambda.1se`: $R^2 = `r gauss_results$R2[1]`$, RMSE $= `r gauss_results$RMSE[1]`$, **`r gauss_results$N_selected[1]`** predictors

Group LASSO at `lambda.min`: $R^2 = `r gauss_results$R2[2]`$, **`r gauss_results$N_selected[2]`** predictors — more complexity, little extra gain

## Plain LASSO Selected Predictors

**SAPS** and **SAS** dominate · **NMP** contributes little · demographics minor

```{r fig-selected-scales}
#| fig-width: 8
#| fig-height: 4.5

ggplot(scale_counts, aes(x = reorder(Scale, N_items), y = N_items, fill = Scale)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Selected items",
       title = "Plain LASSO Selected Predictors by Scale")
```

## Conclusion

- Sparse plain LASSO predicts SPAI total with strong held-out accuracy ($R^2 \approx 0.75$)
- **SAPS** and **SAS** are the dominant predictors; **NMP** adds little once they are included
- High $R^2$ reflects **convergent validity** among overlapping scales, not causation [@gao2024; @lin2014]

## China & Generalizability

Sample collected in **China** from predominantly young adults (mean age $= 18.8$) [@gao2024]

- Results should not be generalized uncritically to other countries or age groups
- Cultural norms around mobile use and education differ across settings
- External validation in U.S. or other populations is needed

## Future Work

- Post-LASSO refitting and robustness across splits / random seeds
- Ordered-outcome models for SPAI item 12
- Prospective validation of a short SAPS + SAS screening subset

## References
