---
title: "LASSO Regression on Cellphone Use Addiction"
subtitle: "IDC6940 Capstone — Problematic Mobile Phone Use"
author: "Richard Odgers, Melissa Bolen, Jillian Mouser, Jonathan De La Torre (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: false
course: Capstone Projects in Data Science
bibliography: references.bib
self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(glmnet)
library(grpreg)
library(caret)
library(dplyr)
library(knitr)
library(ggplot2)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_12_num", "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  if (is.list(raw_coef)) {
    vars <- unique(unlist(lapply(raw_coef, function(m) {
      mat <- as.matrix(m)
      rownames(mat)[mat[, 1] != 0]
    })))
  } else {
    mat <- as.matrix(raw_coef)
    vars <- rownames(mat)[mat[, 1] != 0]
  }
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

lambda_1se_grp <- function(cv_fit) {
  i <- which.min(cv_fit$cve)
  max(cv_fit$lambda[cv_fit$cve <= cv_fit$cve[i] + cv_fit$cvse[i]])
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(R2 = 1 - ss_res / ss_tot, RMSE = sqrt(mean((y_true - y_pred)^2)))
}

dat <- read.csv("data/DATA_Problematic Mobile Phone Use CAT.csv")
dat <- dat[complete.cases(dat), ]
dat$gender <- as.factor(dat$gender)
dat$grade <- as.factor(dat$grade)
dat$SPAI_total <- rowSums(dat[, paste0("SPAI_", 1:20)])
dat$SPAI_12 <- as.factor(dat$SPAI_12)

train_index <- createDataPartition(dat$SPAI_12, p = 0.80, list = FALSE)
mm_reg <- make_reg_mm(dat)
xtr <- mm_reg$x[train_index, , drop = FALSE]
xte <- mm_reg$x[-train_index, , drop = FALSE]
ytr <- mm_reg$y[train_index]
yte <- mm_reg$y[-train_index]

cv_plain <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
pred_plain <- as.numeric(predict(cv_plain, xte, s = "lambda.1se"))
met_plain <- reg_metrics(yte, pred_plain)
sel_plain <- plain_lasso_vars(cv_plain, "lambda.1se", colnames(mm_reg$x))

cv_grp <- cv.grpreg(xtr, ytr, mm_reg$group, family = "gaussian", nfolds = 10)
lam_grp <- cv_grp$lambda.min
pred_grp <- as.numeric(predict(cv_grp, xte, lambda = lam_grp))
met_grp <- reg_metrics(yte, pred_grp)
sel_grp <- group_lasso_vars(cv_grp, lam_grp, colnames(mm_reg$x))

gauss_results <- data.frame(
  Model = c("Plain LASSO | lambda.1se", "Group LASSO | lambda.min"),
  R2 = round(c(met_plain$R2, met_grp$R2), 3),
  RMSE = round(c(met_plain$RMSE, met_grp$RMSE), 2),
  N_selected = c(length(sel_plain), length(sel_grp))
)

scale_counts <- sel_plain %>%
  sapply(assign_group) %>%
  table() %>%
  as.data.frame(stringsAsFactors = FALSE)
names(scale_counts) <- c("Scale", "N_items")
scale_counts <- scale_counts %>%
  filter(Scale != "other") %>%
  arrange(desc(N_items))
```

## The Problem

Smartphone use has shifted from convenience to dependency.

- Compulsive checking, withdrawal, daily-life interference
- Links to anxiety, sleep loss, and academic strain [@lin2014]
- Dozens of overlapping survey variables → multicollinearity [@james2013]

**Question:** Which scales best predict overall addiction severity?

## Our Data

Computer Adaptive Testing (CAT) survey — **1,619** complete cases

| Abbrev. | Scale |
|---------|-------|
| NMP | Nomophobia |
| SAPS | Addiction proneness |
| SPAI | Addiction inventory (outcome) |
| MPAS | Mobile phone addiction |
| MPATS | Addiction tendency |
| SAS | Smartphone addiction scale |

Plus age, gender, and grade.

## Outcome & Approach

**Outcome:** SPAI total (sum of 20 SPAI items, ~20–80)

**Predictors:** all other scales + demographics — SPAI items excluded (they define the score)

**Method:** LASSO [@tibshirani1996] — shrinks weak coefficients to zero and keeps a sparse, interpretable model [@james2013]

## Two Modeling Dials

| | Category (multinomial) | Number (Gaussian) |
|---|---|---|
| **Item-level** (Plain) | Predict SPAI_12 class | Predict SPAI total |
| **Scale-level** (Group) | Predict SPAI_12 class | Predict SPAI total |

- **Plain LASSO** — selects individual items
- **Group LASSO** — keeps or drops whole scales
- **lambda.1se** — simplest model nearly as good as best fit

## SPAI Severity in the Sample

```{r spa-total-hist}
#| fig-width: 8
#| fig-height: 4.5

ggplot(dat, aes(x = SPAI_total)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(
    x = "SPAI total score",
    y = "Respondents",
    title = "Distribution of smartphone addiction severity (SPAI total)"
  ) +
  theme_minimal(base_size = 18)
```

## Live Results — Gaussian LASSO

Single 80/20 split · test set *n* = `r nrow(mm_reg$x) - length(train_index)`

```{r gauss-table}
kable(gauss_results, align = c("l", "r", "r", "r"))
```

Other scales explain about **three-quarters** of SPAI-total variation on held-out respondents.

## Selected Scales (Plain LASSO, lambda.1se)

```{r scale-bar}
#| fig-width: 8
#| fig-height: 4.5

ggplot(scale_counts, aes(x = reorder(Scale, N_items), y = N_items, fill = Scale)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(x = NULL, y = "Selected items", title = "Which instruments matter most?") +
  theme_minimal(base_size = 18)
```

**SAPS** and **SAS** dominate · **NMP** (nomophobia) contributes little

## Recommendation & Next Steps

**Recommended model:** Plain LASSO at `lambda.1se` — strong R² with far fewer predictors than group LASSO

- Scales overlap → convergent validity, not causation
- External validation on a new sample
- Robustness across seeds/splits
- Write up why SAPS/SAS track SPAI total

## References
