---
title: "Descriptive Statistics & Visualizations"
subtitle: "Problematic Mobile Phone Use CAT Data"
author: "Jonathan De La Torre"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# data/ next to this .qmd, or ../data/ if the .qmd lives in a subfolder (e.g. scripts/).
data_path <- function(fname) {
  doc_dir <- dirname(knitr::current_input())
  candidates <- c(
    file.path(doc_dir, "data", fname),
    file.path(doc_dir, "..", "data", fname)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) {
    stop("Cannot find data/", fname, " — expected data/ beside this .qmd or in ../data/")
  }
  normalizePath(hit)
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(scales)
library(purrr)
library(gridExtra)

scale_defs <- tibble(
  Scale = c("NMP", "SAPS", "SPAI", "MPAS", "MPATS", "SAS_C", "SAS_CA"),
  Label = c(
    "Nomophobia (NMP-C)",
    "Smartphone Addiction Proneness (SAPS)",
    "Smartphone Addiction Inventory (SPAI)",
    "Mobile Phone Addiction Scale (MPAS)",
    "Mobile Phone Addiction Tendency Scale (MPATS)",
    "Smartphone Addiction Scale — College (SAS-C)",
    "Smartphone Addiction Scale — Adults (SAS-CA)"
  ),
  Items = c(16, 10, 20, 11, 16, 11, 14),
  Likert = c("1–7", "1–4", "1–4", "1–5", "1–5", "1–5", "1–5"),
  Item_prefix = c("NMP_", "SAPS_", "SPAI_", "MPAS_", "MPATS", "SAS_C", "SAS_CA")
)

gender_labels <- c("1" = "Male", "2" = "Female")
grade_labels <- c(
  "1" = "Freshman", "2" = "Sophomore", "3" = "Junior", "4" = "Senior",
  "5" = "Master Y1", "6" = "Master Y2", "7" = "Master Y3", "8" = "Doctoral",
  "11" = "Grade 7", "12" = "Grade 8", "13" = "Grade 9", "14" = "Grade 10",
  "15" = "Grade 11", "16" = "Grade 12"
)

scale_item_cols <- function(dat, prefix) {
  if (prefix == "MPATS") {
    grep("^MPATS", names(dat), value = TRUE)
  } else {
    grep(paste0("^", prefix), names(dat), value = TRUE)
  }
}

add_scale_totals <- function(dat) {
  out <- dat
  for (i in seq_len(nrow(scale_defs))) {
    cols <- scale_item_cols(dat, scale_defs$Item_prefix[i])
    out[[paste0(scale_defs$Scale[i], "_total")]] <- rowSums(out[, cols], na.rm = FALSE)
  }
  spai_cols <- grep("^SPAI_[0-9]+$", names(dat), value = TRUE)
  out$SPAI_total <- rowSums(out[, spai_cols], na.rm = FALSE)
  out$SAS_total <- out$SAS_C_total + out$SAS_CA_total
  out
}

desc_table <- function(x) {
  tibble(
    N = sum(!is.na(x)),
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Q1 = quantile(x, 0.25, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Q3 = quantile(x, 0.75, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
}
```

# Overview

This document summarizes the **Computer Adaptive Testing (CAT)** survey on
problematic mobile phone use. Instrument definitions follow the project item
bank (`ITEM BANK_Problematic Mobile Phone CAT.docx`). All tables and figures
below are computed live from `DATA_Problematic Mobile Phone Use CAT.csv`.

# Instrument reference

From the item bank documentation:

```{r instrument-table}
#| label: tbl-instruments

kable(scale_defs, align = c("l", "l", "r", "c", "l"))
```

**Demographics:** gender (1 = Male, 2 = Female), grade (see codebook), age.

**Note:** `SPAI_12` is item 12 of the SPAI inventory (4-point response) and is
also used as a categorical severity indicator in classification models.

# Data loading & completeness

```{r load-data}
#| label: load-data

raw <- read.csv(data_path("DATA_Problematic Mobile Phone Use CAT.csv"))
dat <- raw[complete.cases(raw), ]
dat <- add_scale_totals(dat)
dat$gender_lab <- factor(gender_labels[as.character(dat$gender)], levels = unname(gender_labels))
grade_chr <- as.character(dat$grade)
dat$grade_lab <- ifelse(
  grade_chr %in% names(grade_labels),
  grade_labels[grade_chr],
  paste("Code", grade_chr)
)
dat$grade_lab <- factor(dat$grade_lab)

cat("Raw rows:", nrow(raw), "| Complete cases:", nrow(dat), "\n")
cat("Variables:", ncol(raw), "\n")
```

```{r missingness}
#| label: tbl-missingness
#| tbl-cap: "Missing values by variable (raw data). Only variables with any missing shown."

miss <- colSums(is.na(raw))
miss_tbl <- tibble(Variable = names(miss), Missing = as.integer(miss)) %>%
  filter(Missing > 0) %>%
  arrange(desc(Missing)) %>%
  mutate(Pct = percent(Missing / nrow(raw), accuracy = 0.1))

if (nrow(miss_tbl) == 0) {
  miss_tbl <- tibble(Variable = "(none)", Missing = 0, Pct = "0%")
}

kable(miss_tbl, align = c("l", "r", "r"))
```

```{r completeness-plot}
#| label: fig-completeness
#| fig-cap: "Share of respondents with complete vs. incomplete records."

comp_df <- tibble(
  Status = c("Complete", "Incomplete"),
  N = c(nrow(dat), nrow(raw) - nrow(dat))
)

ggplot(comp_df, aes(x = Status, y = N, fill = Status)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = N), vjust = -0.4, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Respondents", title = "Data completeness") +
  theme_minimal(base_size = 12)
```

# Demographics

```{r demo-age}
#| label: tbl-demo-age
#| tbl-cap: "Age summary (complete cases)."

kable(desc_table(dat$age) %>% mutate(across(where(is.numeric), ~ round(.x, 2))),
      col.names = c("N", "Mean", "SD", "Min", "Q1", "Median", "Q3", "Max"))
```

```{r demo-gender-grade}
#| label: tbl-demo-gender-grade
#| tbl-cap: "Gender and grade counts (complete cases)."

demo_tbl <- bind_rows(
  dat %>% count(gender_lab, name = "N") %>% rename(Category = gender_lab) %>% mutate(Variable = "Gender"),
  dat %>% count(grade_lab, name = "N") %>% rename(Category = grade_lab) %>% mutate(Variable = "Grade")
) %>%
  select(Variable, Category, N) %>%
  arrange(Variable, desc(N))

kable(demo_tbl, align = c("l", "l", "r"))
```

```{r fig-demographics}
#| label: fig-demographics
#| fig-cap: "Age distribution and demographic breakdown (complete cases)."
#| fig-width: 10
#| fig-height: 5

p_age <- ggplot(dat, aes(x = age)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(title = "Age", x = "Years", y = "Count") +
  theme_minimal(base_size = 11)

gender_df <- dat %>% count(gender_lab, .drop = FALSE)
p_gender <- ggplot(gender_df, aes(x = gender_lab, y = n, fill = gender_lab)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Gender", x = NULL, y = "Count") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

grade_df <- dat %>% count(grade_lab, .drop = FALSE) %>%
  mutate(grade_lab = reorder(grade_lab, n))
p_grade <- ggplot(grade_df, aes(x = grade_lab, y = n, fill = grade_lab)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Grade / school level", x = NULL, y = "Count") +
  theme_minimal(base_size = 11)

gridExtra::grid.arrange(p_age, p_gender, p_grade, ncol = 3)
```

# Scale totals — descriptive statistics

Scale totals are the **sum of item scores** within each instrument (higher =
more problematic use, per item bank direction).

```{r scale-desc}
#| label: tbl-scale-desc
#| tbl-cap: "Descriptive statistics for scale total scores (complete cases)."

total_cols <- c(paste0(scale_defs$Scale, "_total"), "SPAI_total", "SAS_total")
total_labels <- c(scale_defs$Label, "SPAI total (20 items)", "SAS combined (SAS-C + SAS-CA)")

scale_desc <- map2_dfr(total_cols, total_labels, function(col, lab) {
  desc_table(dat[[col]]) %>% mutate(Scale = lab)
}) %>%
  select(Scale, everything()) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kable(scale_desc, align = c("l", rep("r", 8)))
```

```{r fig-scale-histograms}
#| label: fig-scale-histograms
#| fig-cap: "Distributions of scale total scores."
#| fig-width: 10
#| fig-height: 12

plot_cols <- c(paste0(scale_defs$Scale, "_total"), "SPAI_total")
plot_labels <- c(scale_defs$Label, "SPAI total (20 items)")

hist_df <- dat %>%
  select(all_of(plot_cols)) %>%
  pivot_longer(everything(), names_to = "scale", values_to = "score") %>%
  mutate(scale = factor(scale, levels = plot_cols, labels = plot_labels))

ggplot(hist_df, aes(x = score)) +
  geom_histogram(bins = 25, fill = "#72B7B2", color = "white") +
  facet_wrap(~ scale, scales = "free", ncol = 2) +
  labs(x = "Scale total", y = "Respondents", title = "Scale total score distributions") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(size = 8))
```

```{r fig-scale-boxplots}
#| label: fig-scale-boxplots
#| fig-cap: "Boxplots of scale totals (complete cases)."

box_df <- dat %>%
  select(all_of(plot_cols)) %>%
  pivot_longer(everything(), names_to = "scale", values_to = "score") %>%
  mutate(scale = factor(scale, levels = plot_cols, labels = plot_labels))

ggplot(box_df, aes(x = scale, y = score, fill = scale)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.35) +
  coord_flip() +
  labs(x = NULL, y = "Scale total", title = "Scale totals — spread and outliers") +
  theme_minimal(base_size = 11)
```

# SPAI item 12 (categorical)

Item 12 measures perceived negative impact on study/work (4-point Likert in the
item bank).

```{r spai12-table}
#| label: tbl-spai12
#| tbl-cap: "SPAI_12 response distribution (complete cases)."

spai12_tbl <- dat %>%
  count(SPAI_12, name = "N") %>%
  mutate(Pct = percent(N / sum(N), accuracy = 0.1))

kable(spai12_tbl, col.names = c("SPAI_12 level", "N", "Percent"), align = c("l", "r", "r"))
```

```{r fig-spai12}
#| label: fig-spai12
#| fig-cap: "SPAI_12 category frequencies."

ggplot(spai12_tbl, aes(x = factor(SPAI_12), y = N, fill = factor(SPAI_12))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = N), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "SPAI item 12 response", y = "Respondents", title = "SPAI_12 distribution") +
  theme_minimal(base_size = 12)
```

# Correlations among scale totals

```{r correlation}
#| label: tbl-correlation
#| tbl-cap: "Pearson correlations among scale totals."

corr_cols <- c(paste0(scale_defs$Scale, "_total"), "SPAI_total")
corr_mat <- cor(dat[, corr_cols], use = "complete.obs")
corr_df <- as.data.frame(round(corr_mat, 2))
corr_df <- cbind(Scale = rownames(corr_df), corr_df)
rownames(corr_df) <- NULL

kable(corr_df, align = c("l", rep("r", ncol(corr_df) - 1)))
```

```{r fig-correlation}
#| label: fig-correlation
#| fig-cap: "Correlation heatmap of scale totals."

corr_long <- as.data.frame(corr_mat) %>%
  mutate(Scale1 = rownames(.)) %>%
  pivot_longer(-Scale1, names_to = "Scale2", values_to = "r") %>%
  mutate(
    Scale1 = gsub("_total", "", Scale1),
    Scale2 = gsub("_total", "", Scale2)
  )

ggplot(corr_long, aes(x = Scale2, y = Scale1, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.8, color = "black") +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r", title = "Scale total correlations") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Item-level means by scale

Average response per item (complete cases), faceted by instrument.

```{r item-means}
#| label: fig-item-means
#| fig-cap: "Mean item score within each scale (1 = lowest endorsement per item bank)."
#| fig-width: 10
#| fig-height: 14

item_means_list <- lapply(seq_len(nrow(scale_defs)), function(i) {
  cols <- scale_item_cols(dat, scale_defs$Item_prefix[i])
  tibble(
    item = cols,
    mean_score = colMeans(dat[, cols], na.rm = TRUE),
    scale = scale_defs$Label[i]
  )
})
item_means_df <- bind_rows(item_means_list) %>%
  group_by(scale) %>%
  mutate(item = reorder(item, mean_score)) %>%
  ungroup()

ggplot(item_means_df, aes(x = mean_score, y = item, fill = mean_score)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ scale, scales = "free_y", ncol = 1) +
  labs(x = "Mean item score", y = NULL, title = "Average endorsement by item") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8))
```

# SPAI total by demographics

```{r spai-by-demo}
#| label: tbl-spai-by-demo
#| tbl-cap: "SPAI total by gender and grade (mean ± SD)."

spai_gender <- dat %>%
  group_by(gender_lab) %>%
  summarise(N = n(), Mean = mean(SPAI_total), SD = sd(SPAI_total), .groups = "drop") %>%
  mutate(across(c(Mean, SD), ~ round(.x, 2)), Group = "Gender", Level = as.character(gender_lab))

spai_grade <- dat %>%
  group_by(grade_lab) %>%
  summarise(N = n(), Mean = mean(SPAI_total), SD = sd(SPAI_total), .groups = "drop") %>%
  mutate(across(c(Mean, SD), ~ round(.x, 2)), Group = "Grade", Level = as.character(grade_lab))

spai_demo <- bind_rows(
  spai_gender %>% select(Group, Level, N, Mean, SD),
  spai_grade %>% select(Group, Level, N, Mean, SD)
)

kable(spai_demo, align = c("l", "l", "r", "r", "r"))
```

```{r fig-spai-by-demo}
#| label: fig-spai-by-demo
#| fig-cap: "SPAI total by gender and grade."
#| fig-width: 10
#| fig-height: 5

p1 <- ggplot(dat, aes(x = gender_lab, y = SPAI_total, fill = gender_lab)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  labs(title = "By gender", x = NULL, y = "SPAI total") +
  theme_minimal(base_size = 11)

grade_order <- dat %>%
  group_by(grade_lab) %>%
  summarise(m = mean(SPAI_total), .groups = "drop") %>%
  arrange(m) %>%
  pull(grade_lab)

p2 <- ggplot(dat, aes(x = factor(grade_lab, levels = grade_order), y = SPAI_total, fill = grade_lab)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  coord_flip() +
  labs(title = "By grade", x = NULL, y = "SPAI total") +
  theme_minimal(base_size = 11)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

# Age vs. scale totals

```{r fig-age-scatter}
#| label: fig-age-scatter
#| fig-cap: "Age versus SPAI total and SAS combined total."

scatter_df <- dat %>%
  select(age, SPAI_total, SAS_total) %>%
  pivot_longer(-age, names_to = "outcome", values_to = "score") %>%
  mutate(outcome = recode(outcome,
    SPAI_total = "SPAI total",
    SAS_total = "SAS combined total"
  ))

ggplot(scatter_df, aes(x = age, y = score)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#4C78A8") +
  geom_smooth(method = "lm", se = TRUE, color = "#E45756", linewidth = 0.8) +
  facet_wrap(~ outcome, scales = "free_y") +
  labs(x = "Age", y = "Scale total", title = "Age and addiction severity scores") +
  theme_minimal(base_size = 11)
```

# Summary

- **`r nrow(raw)`** respondents in the raw file; **`r nrow(dat)`** complete cases
  used for all summaries above.
- Seven instrument families plus demographics, documented in the CAT item bank.
- Scale totals are strongly correlated with one another; SAPS, SPAI, MPAS,
  MPATS, and SAS tend to move together (see correlation heatmap).
- Demographic breakdowns and item-level means help locate where respondents
  endorse problematic phone-use symptoms most strongly.
