---
title: "LASSO Cellphone Addiction Analysis Proposal"
subtitle: "Smartphone Addiction Project — Initial Results & Next Steps"
author: "Jonathan De La Torre"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    geometry:
      - margin=1in
execute:
  echo: true
  warning: false
  message: false
---

<!-- If PDF rendering fails with a LaTeX error, install a lightweight engine:
     install.packages("tinytex"); tinytex::install_tinytex()
     Then re-render this document. -->

```{r setup}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# data/ next to this .qmd, or ../data/ if the .qmd lives in a subfolder (e.g. scripts/).
data_csv <- function(fname) {
  doc_dir <- dirname(knitr::current_input())
  candidates <- c(
    file.path(doc_dir, "data", fname),
    file.path(doc_dir, "..", "data", fname)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) {
    stop("Cannot find data/", fname, " — expected data/ beside this .qmd or in ../data/")
  }
  normalizePath(hit)
}

library(glmnet)
library(grpreg)
library(caret)
library(nnet)
library(dplyr)
library(knitr)

RANDOM_SEED <- 123
set.seed(RANDOM_SEED)
# Cross-validation fold assignment may vary slightly even with a fixed seed.
```

::: {.callout-note}
## Reproducibility note

All tables below are produced by executing the embedded R code at render time. Results may vary slightly between renders because 10-fold cross-validation inside `glmnet` and `grpreg` is not fully deterministic, even with `set.seed(123)`.
:::

# Overview — What I'm Proposing {#sec-overview}

Hey team — here's where I landed on our smartphone-addiction analysis, and what I think we can build on together toward the paper.

**The goal:** find out which survey scales best predict overall smartphone-addiction severity, using our dataset of **1,619** people and six validated instruments (NMP, SAPS, SPAI, MPAS, MPATS, SAS) plus demographics.

I've run an initial LASSO analysis (all the code and numbers below are live — they re-run when you render this doc). This write-up is my proposal for how we approach the project going forward: same data, same core question, room for all of us to extend it.

# The Data {#sec-data}

Each person answered items from six scale families plus demographics. Here's a quick cheat sheet:

| Scale | Abbreviation | What it measures |
|-------|--------------|------------------|
| Nomophobia | NMP | Fear of being without one's phone |
| Smartphone Addiction Proneness | SAPS | Tendency toward addictive phone use |
| Smartphone Addiction Inventory | SPAI | Symptom inventory for phone addiction |
| Mobile Phone Addiction Scale | MPAS | Addiction-related behaviors and feelings |
| Mobile Phone Addiction Tendency Scale | MPATS | Tendency toward problematic use |
| Smartphone Addiction Scale | SAS | Cognitive and behavioral addiction symptoms |

We also have age, gender, and grade. After dropping rows with any missing answers, we're left with **1,619** complete cases.

# What We Predict (and Why) {#sec-outcome}

Our main outcome is **SPAI total** — the sum of all 20 SPAI items. It's a continuous severity score (roughly 20 to 80).

We **leave the SPAI items out of the predictors** because they *are* the score. Using them to predict the total would be circular — like predicting someone's final exam score from the individual exam questions you already added up.

Instead, we ask: **do the other scales (nomophobia, proneness, MPAS, MPATS, SAS) and demographics track how severe someone's SPAI total is?** That's the question we can take into the paper.

# How LASSO Works (Quick Version) {#sec-lasso-primer}

LASSO is a regression method that **automatically drops predictors that don't help** and keeps the ones that do. It does this by shrinking coefficients toward zero; unhelpful ones get shrunk all the way to zero.

The penalty strength is controlled by **lambda**:

- **`lambda.min`** — best predictive fit on average (lowest cross-validation error).
- **`lambda.1se`** — the simplest model that's nearly as good (within one standard error of the best). Think: "almost the same R², but fewer variables."

LASSO automatically picks which items matter — we don't hand-select them beforehand.

# The Methods, Explained {#sec-methods}

The naming gets confusing, so here's the plain version. There are really **two independent choices**, not three competing "methods."

## Dial 1: What gets selected — plain vs group

**Plain LASSO** (`glmnet`) decides **item by item**. It might keep one SAPS question and drop another from the same scale.

**Group LASSO** (`grpreg`) is told that each scale is a **group**. It keeps or drops **whole scales at once** — either all of SAPS stays, or all of it goes.

- Plain answers: *which individual items matter?*
- Group answers: *which whole scales matter?*

In my results, group LASSO kept **many more** variables without much gain in R². For the paper we lean toward **plain LASSO at `lambda.1se`** — a deliberate parsimony choice that keeps far fewer variables for nearly the same fit.

## Dial 2: What type of outcome — multinomial vs Gaussian

This is set by `family =` in the model. It's a **separate dial** from plain vs group.

- **`family = "multinomial"`** — predict an **unordered category**. We used this when predicting a single SPAI item (`SPAI_12`) as one of four response options.
- **`family = "gaussian"`** — predict a **continuous number**. We used this for **SPAI total**.

Important: **"Gaussian LASSO" is not a third method.** It's just plain-or-group LASSO applied to a numeric outcome. You still choose plain *or* group; Gaussian only tells the model you're predicting a number instead of a category.

## Two dials at a glance

| | **Predict a category** (multinomial) | **Predict a number** (Gaussian) |
|---|---|---|
| **Drop items individually** (Plain LASSO) | Plain LASSO → SPAI_12 class | Plain LASSO → SPAI total |
| **Drop whole scales** (Group LASSO) | Group LASSO → SPAI_12 class | Group LASSO → SPAI total |

**Anchor line:** plain vs group = drop items or drop whole scales. Multinomial vs Gaussian = predict a category or a number. These two choices are **independent** — you mix and match them.

## Everything I ran so far

All models share **one 80/20 train/test split** (same people held out every time), so the numbers are comparable.

**Classification side** (multinomial family — predicting `SPAI_12` as a 4-level category):

- Plain and Group LASSO, each in a few predictor-set variants (all predictors, SPAI items removed, SPAI + SAS removed).
- Eight models total. I report balanced accuracy and Kappa on the test set.

`SPAI_12` is imbalanced — the rarest category is only about 6% of responses:

```{r class-imbalance}
#| label: tbl-class-imbalance
#| tbl-cap: "Distribution of SPAI_12 response categories (complete cases only)."

dat_raw <- read.csv(data_csv("DATA_Problematic Mobile Phone Use CAT.csv"))
dat_raw <- dat_raw[complete.cases(dat_raw), ]

class_tbl <- as.data.frame(table(SPAI_12 = dat_raw$SPAI_12))
class_tbl$Percent <- round(100 * class_tbl$Freq / sum(class_tbl$Freq), 1)
kable(class_tbl, align = c("l", "r", "r"))
```

**Regression side** (Gaussian family — predicting **SPAI total**):

- Plain and Group LASSO, each at `lambda.min` and `lambda.1se`.
- Four models. This is our **main result** for the paper.

That's **12 models** in one comparison table. The code below loads the data, defines helper functions, fits everything, and builds the tables.

```{r data-prep}
#| label: data-prep

dat <- read.csv(data_csv("DATA_Problematic Mobile Phone Use CAT.csv"))
dat <- dat[complete.cases(dat), ]

dat$gender <- as.factor(dat$gender)
dat$grade <- as.factor(dat$grade)
dat$SPAI_total <- rowSums(dat[, paste0("SPAI_", 1:20)])
dat$SPAI_12 <- as.factor(dat$SPAI_12)
dat$SPAI_12_num <- as.numeric(dat$SPAI_12)

spai_drop <- setdiff(grep("^SPAI_", names(dat), value = TRUE), "SPAI_12")
sas_drop <- grep("^SAS_", names(dat), value = TRUE)

train_index <- createDataPartition(dat$SPAI_12, p = 0.80, list = FALSE)
y_class_test <- dat$SPAI_12[-train_index]

cat("Training n =", length(train_index), "| Test n =", length(y_class_test), "\n")
```

```{r helpers}
#| label: helpers

class_metrics <- function(pred, y) {
  pred <- factor(pred, levels = levels(y))
  cm <- confusionMatrix(pred, y)
  bc <- as.data.frame(cm$byClass)
  list(
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Balanced_Accuracy = mean(bc$Sensitivity, na.rm = TRUE),
    Kappa = as.numeric(cm$overall["Kappa"])
  )
}

reg_metrics <- function(y_true, y_pred) {
  ss_res <- sum((y_true - y_pred)^2)
  ss_tot <- sum((y_true - mean(y_true))^2)
  list(
    R2 = 1 - ss_res / ss_tot,
    RMSE = sqrt(mean((y_true - y_pred)^2))
  )
}

assign_group <- function(nm) {
  if (grepl("^gender", nm)) return("gender")
  if (grepl("^grade", nm)) return("grade")
  if (nm == "age") return("age")
  if (grepl("^NMP_", nm)) return("NMP")
  if (grepl("^SAPS_", nm)) return("SAPS")
  if (grepl("^SPAI_", nm)) return("SPAI")
  if (grepl("^MPAS_", nm)) return("MPAS")
  if (grepl("^MPATS", nm)) return("MPATS")
  if (grepl("^SAS_CA", nm)) return("SAS_CA")
  if (grepl("^SAS_C", nm)) return("SAS_C")
  "other"
}

make_class_mm <- function(dat, drop_cols = character(0)) {
  d <- dat[, !names(dat) %in% c("SPAI_12_num", "SPAI_total", drop_cols), drop = FALSE]
  x <- model.matrix(SPAI_12 ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_12,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

make_reg_mm <- function(dat) {
  spai_items <- paste0("SPAI_", 1:20)
  d <- dat[, !names(dat) %in% c(spai_items, "SPAI_12_num", "SPAI_total"), drop = FALSE]
  d$SPAI_total <- dat$SPAI_total
  x <- model.matrix(SPAI_total ~ ., data = d)[, -1]
  list(
    x = x,
    y = d$SPAI_total,
    group = as.numeric(factor(sapply(colnames(x), assign_group)))
  )
}

plain_lasso_vars <- function(cv_fit, lambda_label, cols) {
  raw_coef <- coef(cv_fit, s = lambda_label)
  if (is.list(raw_coef)) {
    vars <- unique(unlist(lapply(raw_coef, function(m) {
      mat <- as.matrix(m)
      rownames(mat)[mat[, 1] != 0]
    })))
  } else {
    mat <- as.matrix(raw_coef)
    vars <- rownames(mat)[mat[, 1] != 0]
  }
  intersect(setdiff(vars, "(Intercept)"), cols)
}

group_lasso_vars <- function(cv_fit, lambda_val, cols) {
  coefs <- coef(cv_fit, lambda = lambda_val)
  intersect(names(coefs)[coefs != 0 & names(coefs) != "(Intercept)"], cols)
}

lambda_1se_grp <- function(cv_fit) {
  i <- which.min(cv_fit$cve)
  max(cv_fit$lambda[cv_fit$cve <= cv_fit$cve[i] + cv_fit$cvse[i]])
}

add_result <- function(results, row) {
  new_row <- data.frame(
    Model = row$Model,
    Part = row$Part,
    Variant = row$Variant,
    Outcome = row$Outcome,
    Lambda = row$Lambda,
    Balanced_Accuracy = row$Balanced_Accuracy,
    Kappa = row$Kappa,
    R2 = row$R2,
    RMSE = row$RMSE,
    N_selected = row$N_selected,
    stringsAsFactors = FALSE
  )
  rbind(results, new_row)
}

run_plain_class <- function(x, y, train_index, tag) {
  xtr <- x[train_index, , drop = FALSE]
  xte <- x[-train_index, , drop = FALSE]
  ytr <- y[train_index]
  yte <- y[-train_index]

  cv <- cv.glmnet(xtr, ytr, alpha = 1, family = "multinomial", nfolds = 10)
  sel <- plain_lasso_vars(cv, "lambda.min", colnames(x))
  pred_a <- predict(cv, xte, s = "lambda.min", type = "class")

  if (length(sel) > 0) {
    tr <- cbind(SPAI_12 = ytr, as.data.frame(xtr[, sel, drop = FALSE]))
    te <- cbind(SPAI_12 = yte, as.data.frame(xte[, sel, drop = FALSE]))
    pred_b <- predict(multinom(SPAI_12 ~ ., tr, trace = FALSE), te, type = "class")
  } else {
    pred_b <- rep(levels(yte)[1], length(yte))
  }

  list(
    a = list(name = paste(tag, "A: LASSO"), pred = pred_a, n = length(sel), sel = sel),
    b = list(name = paste(tag, "B: Post-LASSO"), pred = pred_b, n = length(sel), sel = sel)
  )
}

run_group_class <- function(x, y, group, train_index, tag) {
  xtr <- x[train_index, , drop = FALSE]
  xte <- x[-train_index, , drop = FALSE]
  ytr <- y[train_index]
  yte <- y[-train_index]
  ytr_num <- as.numeric(ytr)

  cv <- cv.grpreg(xtr, ytr_num, group, family = "gaussian", nfolds = 10)
  lam <- cv$lambda.min
  sel <- group_lasso_vars(cv, lam, colnames(x))
  pred_num <- as.numeric(predict(cv, xte, lambda = lam))
  pred_a <- factor(
    pmax(1, pmin(nlevels(yte), round(pred_num))),
    levels = levels(yte)
  )

  if (length(sel) > 0) {
    tr <- cbind(SPAI_12 = ytr, as.data.frame(xtr[, sel, drop = FALSE]))
    te <- cbind(SPAI_12 = yte, as.data.frame(xte[, sel, drop = FALSE]))
    pred_b <- predict(multinom(SPAI_12 ~ ., tr, trace = FALSE), te, type = "class")
  } else {
    pred_b <- rep(levels(yte)[1], length(yte))
  }

  list(
    a = list(name = paste(tag, "A: Group LASSO"), pred = pred_a, n = length(sel), sel = sel),
    b = list(name = paste(tag, "B: Post-Group-LASSO"), pred = pred_b, n = length(sel), sel = sel)
  )
}

run_gaussian <- function(x, y, train_index, tag, method = c("plain", "group"),
                         group = NULL, lambda_label = "lambda.min") {
  method <- match.arg(method)
  xtr <- x[train_index, , drop = FALSE]
  xte <- x[-train_index, , drop = FALSE]
  ytr <- y[train_index]
  yte <- y[-train_index]

  if (method == "plain") {
    cv <- cv.glmnet(xtr, ytr, alpha = 1, family = "gaussian", nfolds = 10)
    if (lambda_label == "lambda.min") {
      pred <- as.numeric(predict(cv, xte, s = "lambda.min"))
      sel <- plain_lasso_vars(cv, "lambda.min", colnames(x))
      lam <- cv$lambda.min
    } else {
      pred <- as.numeric(predict(cv, xte, s = "lambda.1se"))
      sel <- plain_lasso_vars(cv, "lambda.1se", colnames(x))
      lam <- cv$lambda.1se
    }
  } else {
    cv <- cv.grpreg(xtr, ytr, group, family = "gaussian", nfolds = 10)
    if (lambda_label == "lambda.min") {
      lam <- cv$lambda.min
    } else {
      lam <- lambda_1se_grp(cv)
    }
    pred <- as.numeric(predict(cv, xte, lambda = lam))
    sel <- group_lasso_vars(cv, lam, colnames(x))
  }

  list(name = tag, pred = pred, yte = yte, n = length(sel), sel = sel, lambda = lam)
}
```

```{r model-matrices}
#| label: model-matrices

mm_full <- make_class_mm(dat)
mm_c1 <- make_class_mm(dat, spai_drop)
mm_c2 <- make_class_mm(dat, c(spai_drop, sas_drop))
mm_reg <- make_reg_mm(dat)

stopifnot(
  nrow(mm_full$x) == nrow(mm_c1$x),
  nrow(mm_full$x) == nrow(mm_c2$x),
  nrow(mm_full$x) == nrow(mm_reg$x)
)
```

```{r fit-classification}
#| label: fit-classification

results <- data.frame(
  Model = character(),
  Part = character(),
  Variant = character(),
  Outcome = character(),
  Lambda = character(),
  Balanced_Accuracy = numeric(),
  Kappa = numeric(),
  R2 = numeric(),
  RMSE = numeric(),
  N_selected = integer(),
  stringsAsFactors = FALSE
)

plain_full <- run_plain_class(mm_full$x, mm_full$y, train_index, "Plain")
plain_c1 <- run_plain_class(mm_c1$x, mm_c1$y, train_index, "Plain C1")
plain_c2 <- run_plain_class(mm_c2$x, mm_c2$y, train_index, "Plain C2")

plain_rows <- list(
  list(Model = "Plain A: LASSO (all predictors)", Variant = "A", m = plain_full$a),
  list(Model = "Plain B: Post-LASSO (all predictors)", Variant = "B", m = plain_full$b),
  list(Model = "Plain C1: LASSO (no SPAI predictors)", Variant = "C1", m = plain_c1$a),
  list(Model = "Plain C2: LASSO (no SPAI/SAS predictors)", Variant = "C2", m = plain_c2$a)
)

for (row in plain_rows) {
  met <- class_metrics(row$m$pred, y_class_test)
  results <- add_result(results, list(
    Model = row$Model,
    Part = "Plain LASSO",
    Variant = row$Variant,
    Outcome = "SPAI_12",
    Lambda = "lambda.min",
    Balanced_Accuracy = met$Balanced_Accuracy,
    Kappa = met$Kappa,
    R2 = NA,
    RMSE = NA,
    N_selected = row$m$n
  ))
}

grp_full <- run_group_class(mm_full$x, mm_full$y, mm_full$group, train_index, "Group")
grp_c1 <- run_group_class(mm_c1$x, mm_c1$y, mm_c1$group, train_index, "Group C1")
grp_c2 <- run_group_class(mm_c2$x, mm_c2$y, mm_c2$group, train_index, "Group C2")

group_rows <- list(
  list(Model = "Group A: Group LASSO (all predictors)", Variant = "A", m = grp_full$a),
  list(Model = "Group B: Post-Group-LASSO (all predictors)", Variant = "B", m = grp_full$b),
  list(Model = "Group C1: Group LASSO (no SPAI predictors)", Variant = "C1", m = grp_c1$a),
  list(Model = "Group C2: Group LASSO (no SPAI/SAS predictors)", Variant = "C2", m = grp_c2$a)
)

for (row in group_rows) {
  met <- class_metrics(row$m$pred, y_class_test)
  results <- add_result(results, list(
    Model = row$Model,
    Part = "Group LASSO",
    Variant = row$Variant,
    Outcome = "SPAI_12",
    Lambda = "lambda.min",
    Balanced_Accuracy = met$Balanced_Accuracy,
    Kappa = met$Kappa,
    R2 = NA,
    RMSE = NA,
    N_selected = row$m$n
  ))
}
```

Plain LASSO classification on `SPAI_12` under three predictor sets (variant A: all predictors; C1: SPAI items excluded; C2: SPAI and SAS items excluded):

```{r tbl-class-predictor-sets}
#| label: tbl-class-predictor-sets
#| tbl-cap: "Plain LASSO classification: test-set performance under different predictor sets."

class_pred_sets <- results %>%
  filter(Part == "Plain LASSO", Variant %in% c("A", "C1", "C2")) %>%
  select(Model, Balanced_Accuracy, Kappa, N_selected) %>%
  mutate(
    Balanced_Accuracy = round(Balanced_Accuracy, 3),
    Kappa = round(Kappa, 3)
  )

kable(class_pred_sets, align = c("l", "r", "r", "r"))
```

```{r fit-gaussian}
#| label: fit-gaussian

gauss_models <- list(
  run_gaussian(mm_reg$x, mm_reg$y, train_index,
               "Gaussian Plain | lambda.min", "plain", lambda_label = "lambda.min"),
  run_gaussian(mm_reg$x, mm_reg$y, train_index,
               "Gaussian Plain | lambda.1se", "plain", lambda_label = "lambda.1se"),
  run_gaussian(mm_reg$x, mm_reg$y, train_index,
               "Gaussian Group | lambda.min", "group",
               group = mm_reg$group, lambda_label = "lambda.min"),
  run_gaussian(mm_reg$x, mm_reg$y, train_index,
               "Gaussian Group | lambda.1se", "group",
               group = mm_reg$group, lambda_label = "lambda.1se")
)

gauss_rows <- list(
  list(Model = "Gaussian Plain | lambda.min", Variant = "Plain", Lambda = "lambda.min"),
  list(Model = "Gaussian Plain | lambda.1se", Variant = "Plain", Lambda = "lambda.1se"),
  list(Model = "Gaussian Group | lambda.min", Variant = "Group", Lambda = "lambda.min"),
  list(Model = "Gaussian Group | lambda.1se", Variant = "Group", Lambda = "lambda.1se")
)

for (i in seq_along(gauss_models)) {
  m <- gauss_models[[i]]
  met <- reg_metrics(m$yte, m$pred)
  results <- add_result(results, list(
    Model = gauss_rows[[i]]$Model,
    Part = "Gaussian LASSO",
    Variant = gauss_rows[[i]]$Variant,
    Outcome = "SPAI_total",
    Lambda = gauss_rows[[i]]$Lambda,
    Balanced_Accuracy = NA,
    Kappa = NA,
    R2 = met$R2,
    RMSE = met$RMSE,
    N_selected = m$n
  ))
}

stopifnot(nrow(results) == 12)
```

# Results So Far {#sec-results}

## Master comparison — all 12 models

```{r master-table}
#| label: tbl-master

results_display <- results %>%
  mutate(
    Balanced_Accuracy = ifelse(is.na(Balanced_Accuracy), NA, round(Balanced_Accuracy, 3)),
    Kappa = ifelse(is.na(Kappa), NA, round(Kappa, 3)),
    R2 = ifelse(is.na(R2), NA, round(R2, 3)),
    RMSE = ifelse(is.na(RMSE), NA, round(RMSE, 2))
  )

kable(
  results_display,
  booktabs = TRUE,
  longtable = TRUE,
  caption = paste0(
    "All models on the shared test set (n = ", length(y_class_test),
    "). Classification rows use balanced accuracy and Kappa; Gaussian rows use R² and RMSE."
  )
)
```

## Gaussian regression — our main result

```{r gaussian-table}
#| label: tbl-gaussian
#| tbl-cap: "Gaussian LASSO models predicting SPAI total from non-SPAI scales (test set)."

gauss_tbl <- results %>%
  filter(Outcome == "SPAI_total") %>%
  select(Model, Lambda, R2, RMSE, N_selected) %>%
  mutate(R2 = round(R2, 3), RMSE = round(RMSE, 2))

kable(gauss_tbl, align = c("l", "l", "r", "r", "r"))
```

## The plain read

Test R² is around **`r round(max(results$R2, na.rm = TRUE), 2)`** — so the other scales explain about **three-quarters** of the variation in SPAI severity on people the model hasn't seen.

**SAPS** (addiction proneness) and **SAS** do most of the work. **Nomophobia (NMP)** barely shows up — once proneness and SAS are in the model, fear-of-disconnection items don't add much.

For the paper we recommend **plain LASSO at `lambda.1se`** (R² ≈ **`r round(results_display$R2[results_display$Model == "Gaussian Plain | lambda.1se"], 2)`** in this render) because it keeps far fewer variables for nearly the same R² — a parsimony choice, not a hunt for the single highest decimal. Group LASSO at `lambda.min` lands in a similar R² range but with many more predictors in the mix.

# Things to Keep in Mind {#sec-caveats}

- **The scales overlap.** They all measure facets of problematic phone use, so a high R² partly means "they move together" — convergent validity, not proof that one scale *causes* another.
- **Numbers wobble a bit.** Cross-validation picks the penalty, and fold assignment isn't perfectly repeatable. Treat the results as solid but not exact to the third decimal.

# Where the Group Can Take It Next {#sec-next}

Here are alternative directions we could split up and develop for the paper — totally open, not assignments:

- **Ordinal models** — try an ordered-outcome version for the categorical SPAI_12 analysis instead of plain multinomial.
- **Write up the SAPS/SAS story** — why proneness and the SAS scale track SPAI total so closely, and what that means substantively.
- **Robustness checks** — different random seeds or train/test splits to confirm the R² and variable-selection pattern holds up.
- **Why nomophobia drops out** — dig into whether NMP is redundant with SAPS/SAS or just weaker on its own.

That's the foundation. Happy to walk through any of the ideas above.
